---
layout: page
title: "Normal analyses"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
header-includes:
- \usepackage{blkarray}
- \usepackage{amsmath}
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](18-normal_analyses.R)

```{r}
library("tidyverse"); theme_set(theme_bw())
library("Sleuth3")
set.seed(20230319)
```

# Normal analyses

Normal analyses are (typically) relevant for independent, continuous 
observations. 
These observations may need to be transformed to satisfy assumptions of 
normality. 
Pedagogically, we often begin with a single group with a known variance.
A known variance is uncommon in practice and thus, here, we skip directly
to the unknown variance situation. 

## One group

The single group normal analysis begins with the assumptions
\[ 
Y_i \stackrel{ind}{\sim} N(mu,\sigma^2)
\]
for $i=1,\ldots,n$.

An alternative way to write this model is 
\[ 
Y_i = \mu + \epsilon_i, \qquad \epsilon_i \stackrel{ind}{\sim} N(0,\sigma^2)
\]

which has the following assumptions

- errors are independent,
- errors are normally distributed,
- errors have a constant variance,

and 

- common mean. 

### Summary statistics

```{r}
creativity <- case0101 %>%
  filter(Treatment == "Intrinsic")
```

```{r}
summary(creativity)

creativity %>%
  summarize(
    n = n(),
    mean = mean(Score),
    sd = sd(Score)
  )
```

```{r}
ggplot(creativity, aes(x = Score)) + 
  geom_histogram()
```

```{r}
ggplot(creativity, aes(x = Score)) + 
  geom_boxplot(outlier.shape = NA)
```


### t.test

In a t-test, we are typically interested in the hypothesis
\[ 
H_0: \mu = \mu_0
\]
for some value $\mu_0$. 



```{r}
t.test(creativity$Score)
```
Alternatively, using a formula

```{r}
t.test(Score ~ 1, data = creativity)
```


### Additional arguments

```{r}
t.test(creativity$Score, mu = 20, alternative = "greater", conf.level = 0.9)
```



## Two groups

When you have two groups, the assumption is 

\[ 
Y_{ig} \stackrel{ind}{\sim} N(\mu_g, \sigma_g^2)
\]
for $i = 1, \ldots, n_g$ and $g=1,2$. Alternatively

\[ 
Y_{ig} = \mu_g + \epsilon_{ig}, \qquad \epsilon_{ig} \stackrel{ind}{\sim} N(0,\sigma_g^2)
\]
which assumes

- errors are independent,
- errors are normally distributed, 
- errors have constant variance within a group, 

and 

- each group has its own constant mean.

### Summary statistics

```{r}
case0101 %>%
  group_by(Treatment) %>%
  summarize(
    n = n(),
    mean = mean(Score),
    sd = sd(Score)
  )
```

### Tests and confidence intervals

Typically we are interested in the hypothesis
\[ 
H_0: \mu_1 = \mu_2 \qquad \mbox{or, equivalently,} \qquad H_0: \mu_1-\mu_2 = 0 
\]
and a confidence interval for the difference $\mu_1-\mu_2$. 


#### Unequal variances

To run the test, 
we first need to get the appropriate data

```{r}
intrinsic <- case0101$Score[case0101$Treatment == "Intrinsic"]
extrinsic <- case0101$Score[case0101$Treatment == "Extrinsic"]
```

then we can run the test

```{r}
t.test(extrinsic, intrinsic)
```
Alternatively, we can use the formula version and simplify our code

```{r}
t.test(Score ~ Treatment, data = case0101)
```



#### Equal variances

Often, we will make the assumption that the two groups have equal variances. 

```{r}
t.test(Score ~ Treatment, data = case0101, var.equal = TRUE)
```
#### Additional arguments

```{r}
t.test(Score ~ Treatment, data = case0101, var.equal = TRUE,
       mu = 5, alternative = "greater", conf.level = 0.9)
```

### Paired data

A paired analysis can be used when the data are have a natural grouping 
structure into a pair of observations.
Within each pair of observations, 
one observation has one level of the explanatory variable while the other
observation has the other level of the explanatory variable. 

In the following example, the data are naturally paired because each set of
observations is on a set of identical twins. 
One twin has schizophrenia while the other does not. 
These data compares the volume of the left hippocampus to understand its 
relationship to schizophrenia.

#### Summary statistics

```{r}
summary(case0202)
```

```{r}
schizophrenia <- case0202 %>%
  mutate(pair = 1:n()) %>%
  pivot_longer(-pair, names_to = "diagnosis", values_to = "volume")
```

```{r}
ggplot(schizophrenia, aes(x = diagnosis, y = volume, group = pair)) + 
  geom_line()
```

It is important to connect these points between each twin because we are 
looking for a pattern amongst the comparisons. 

#### Paired t-test

When the data are paired, we can perform a paired t-test. 

```{r}
t.test(case0202$Unaffected, case0202$Affected, paired = TRUE)
```
This analysis is equivalent to taking the difference in each pair and 
performing a one-sample t-test. 

```{r}
t.test(case0202$Unaffected - case0202$Affected)
```

### More than two groups

If we have a single categorical variable with more than two groups, then $G > 2$. 

#### Summary statistics

```{r}
case0501 %>%
  group_by(Diet) %>%
  summarize(n = n(),
            mean = mean(Lifetime),
            sd = sd(Lifetime))
```

```{r}
ggplot(case0501, aes(x = Diet, y = Lifetime)) +
  geom_jitter(width = 0.1)
```

## Regression

As a general approach, regression allows the response variable to depend on
categorical and continuous variables in complex patterns. 
For the current discussion, we will limit our discussion to one categorical
variable. 

The general structure of a regression model is 
\[ 
Y_i \stackrel{ind}{\sim} N(\beta_0 + \beta_1X_{i,1} + \cdots + \beta_p X_{i,p}, \sigma^2)
\]
for $i=1,\ldots,n$
or, equivalently,
\[ 
Y_i = \beta_0 + \beta_1X_{i,1} + \cdots + \beta_p X_{i,p}, \qquad \epsilon_i \stackrel{ind}{\sim} N(0, \sigma^2).
\]
The assumptions a regression model are 

- errors are independent,
- errors are normally distributed,
- errors have constant variance, 

and 

- the expected response, $E[Y_i]$, depends on the explanatory variables according
to a linear function (of the parameters). 

### Two groups

We can repeat the equal variance analysis using regression by encoding a 
*dummy* or (*indicator*) variable to represent the group. 
So let 
\[ 
X_{i,1} = \left\{ 
\begin{array}{ll}
1 & \mbox{if Treatment}_i \mbox{ is Intrinsic} \\
0 & \mbox{otherwise}
\end{array}
\right.
\]

```{r}
m <- lm(Score ~ Treatment, data = case0101)
```

We can summarize the analysis in a few different ways. 

#### Standard regression output

```{r}
summary(m)
```

#### Confidence intervals

```{r}
confint(m)
```

#### ANOVA

```{r}
anova(m)
```


### More than two groups

If we have a single categorical variable with more than two groups, then $G > 2$. 



#### Summary statistics


### Two categorical variables
