---
title: "Lab08 - Categorical explanatory variables"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


To follow along, use the [lab08 code](lab08.R) and make sure the following 
packages are installed:

- Sleuth3

You can use the following code to perform the installation:

```{r install_packages, eval=FALSE}
install.packages(c("Sleuth3"))
```

Now load the packages

```{r load_packages}
library("dplyr")
library("ggplot2")
library("Sleuth3")
```

## Overview

Categorical variables in R are represented either by a 

- character
- factor

A factor is a special character variable that 

- has an underlying numerical representation,
- a lookup table to find the values, and
- an inherent ordering.

When using characters are factors in a regression model in R, 
the character will automatically be converted to a factor. 


## Characters vs factors

Most `data.frame`s in R will have character variables automatically converted
into factors. 

```{r}
str(case0501)
```

Note that `Diet` is a factor. 

```{r}
str(case0501$Diet)
```

Although printing out this variable makes it look like a character:

```{r}
case0501$Diet
```
There is actually an internal representation via integers:

```{r}
as.numeric(case0501$Diet)
```

This representation can be accessed using `levels()`:

```{r}
nlevels(case0501$Diet)
levels(case0501$Diet)
```

The ordering here is alphabetical.

When you run `summary()` or `table()`, 
R uses this ordering for displaying results:

```{r}
summary(case0501$Diet)
table(case0501$Diet)
```

You can convert characters to factors. 

```{r}
this <- as.factor(c("a","a","b","c"))
this
```

You can also convert factors back to characters.

```{r}
as.character(this)
```

When reading data into R, `read.csv()` will automatically convert 
characters into factors while `read_csv()` from the `dplyr` package doesn't. 

#### Activity

Practice reading and writing files using `read.csv()` and `dlpyr::read_csv()`.
Specifically, do the following

1. Write `case0501` to a csv using `write.csv()`. Use the help file to ensure that you do not write the rownames.
1. Read the created csv file in using both `read.csv()` and `dlpyr::read_csv()`. What differences are there?
1. Use the help files for `read.csv()` and `dlpyr::read_csv()` to make the `data.frame`s equivalent. 

<div id="readcsv_activity_solution" style="display:none"> 
```{r, purl=FALSE}

```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('readcsv_activity_solution') .style.display=='none') {document.getElementById('readcsv_activity_solution') .style.display=''}else{document.getElementById('readcsv_activity_solution') .style.display='none'}">Show/Hide Solution</button>
