---
layout: page
title: STAT 486/586
tagline: "Functions"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](11-functions.R)

R is a [functional programming language](https://en.wikipedia.org/wiki/Functional_programming).

```{r}
class(log)
```



# Function basics

Functions take in some input and return some output. 
The input are a collection of arguments to the function and the output is 
the return value. 

## Arguments

```{r}
log(10)
log(x = 10)
log(10, base = exp(1))
log(10, base = 10)
log(x = 10, base = 10)
```

Take a look at the arguments. 

```{r}
args(log)
```



### Default arguments

In the `log` function, 
the default value for the `base` argument is `exp(1)`. 

```{r}
all.equal(
  log(10),
  log(10, base = exp(1))
)
```

### Positional matching

```{r}
log(10, exp(1))
log(exp(1), 10)
```

### Name matching

```{r}
log(x = 10, base = exp(1))
log(base = exp(1), x = 10)
```



### Partial matching

```{r}
log(10, b = exp(1))
log(10, ba = exp(1))
log(10, bas = exp(1))
log(10, base = exp(1))
```


### R objects as input

```{r}
y <- 100
log(y)
```




## Return value

```{r}
class(log(10))
class(as.data.frame(10))
```

```{r}
class(all.equal(1,1))
class(all.equal(1,2))
```

```{r}
m <- lm(len ~ dose, data = ToothGrowth)
class(m)
class(summary(m))
```

# Building functions

## Function definition

```{r}
# Create a function
add <- function(x, y) {
  x + y
}
```

```{r}
add(1,2)
add(x = 1, y = 2)
```

```{r}
add(1:2, 3:4)
```

```{r}
add(1:2, 3)
```

```{r}
add(1:2, 3:5)
```



## Default arguments

```{r}
add <- function(x = 1, y = 2) {
  x + y
}
```

```{r}
add()
add(3)
add(y = 5)
```



## Explicit return

R functions will return the last 

```{r explicit_return}
add <- function(x, y) {
  return(x + y)
}
```

```{r}
add(1,2)
```


Suppose you want to return a TRUE/FALSE depending on whether a specific 
character is in a string.
As soon as you find the character, you can immediately return TRUE
If you don't find the character, you can return FALSE. 

```{r}
is_char_in_string <- function(string, char) {
  for (i in 1:nchar(string)) {
    if (char == substr(string, i, i))
      return(TRUE)
  }
  return(FALSE)
}
```

```{r}
is_char_in_string("this is my string", "a")
is_char_in_string("this is my string", "s")
```



## Error handling

```{r, error=TRUE}
add(1, "a")
```

### message()

```{r}
add <- function(x, y) {
  message("Here is a message!")
  return(x + y)
}
```

```{r}
add(1,2)
```


### warning()

```{r}
add <- function(x, y) {
  warning("Here is a warning!")
  return(x + y)
}
```

```{r}
add(1,2)
```

### stop()

```{r}
add <- function(x, y) {
  stop("Here is an error!")
  return(x + y)
}
```

```{r, error=TRUE}
add(1,2)
```



### stopifnot()

```{r}
add <- function(x, y) {
  stopifnot(x < 0)
  return(x + y)
}
```

```{r, error=TRUE}
add(1,2)
```
```{r}
add <- function(x, y) {
  stopifnot(is.numeric(x))
  stopifnot(is.numeric(y))
  return(x + y)
}
```

```{r, error=TRUE}
add(1,2)
add(1,"a")
add("b")
```



# Function issues

These are some issues I want you to be aware of so you (I hope) avoid issues
in the future. 

## Argument vs object assignment

```{r}
my_fancy_function <- function(x, y) {
  return(x + y*100)
}
```

What is the result of the following?

```{r}
my_fancy_function(y <- 5, x <- 4)
```
What happened?
We assigned `y` the value 5 and `x` the value 4 outside the function. 
Then, we passed `y`(5) as the first argument of the function and
`x`(4) as the second argument fo the function. 

This was equivalent to 

```{r}
y <- 5
x <- 4
my_fancy_function(x = y, y = x)
```
So, when assigning function arguments, use `=`. 
Also, it is probably helpful to avoid naming objects the same name as the 
argument names. 



## Scoping

Here is a function

```{r scoping}
f <- function() {
  return(y)
}
```

What is the result of the following?

```{r}
f()
```

Basically, R searches through a series of *environments* to find the variable
called `y`. 


## Closure errors

Sometimes you get baffling error messages due to `closure` errors or
`special` errors. 

```{r, error=TRUE}
mean[1]
log[1]
```

This is related to functions having a `typeof` `closure` or `special`. 

```{r}
typeof(mean)
typeof(log)
```

You will see `closure` errors much more commonly than `special` errors. 


## Generic functions

### mean()

```{r}
print(mean)
```

Take a look at the help file

```{r, eval=FALSE, comment=""}
?mean
```

Notice the words "Generic function". 
This means what the function does will depend on the class of the first argument 
to the function. 

```{r}
mean(1:5)
mean(as.Date(c("2023-01-01","2022-01-01")))
```

I bring up generic functions primarily to point out that it can be hard to 
track down the appropriate helpful. 
Generally you will look up `<function>.<class>`. 

For example,
```{r}
# Determine the class
class(as.Date(c("2023-01-01","2022-01-01")))
```

```{r, eval=FALSE, comment=""}
# Look up the function
?mean.Date
```

This didn't provide the actual help information. 
Because it went somewhere, this was the intended behavior. 
Why it isn't documented, I have no idea. 

```{r}
class(1:5)
```

```{r, eval=FALSE, comment=""}
?mean.integer
```

There is typically a default method that will be used if a specific method
can't be found.

```{r, eval=FALSE, comment=""}
?mean.default
```

### summary()

```{r}
summary(ToothGrowth$len)
summary(ToothGrowth$supp)
summary(ToothGrowth)
summary(lm(len ~ supp, data = ToothGrowth))
```

```{r, eval=FALSE, comment=""}
?summary
?summary.numeric
?summary.factor
?summary.data.frame
?summary.lm
```



## ... argument

```{r, eval=FALSE, comment=""}
?sum
```

```{r}
sum(1,2,3)
```

```{r}
sum(5:6)
```

```{r}
sum(1,2,3,5:6)
```

Typos get ignored

```{r}
sum(c(1,2,NA), na.mr = TRUE) # vs 
sum(c(1,2,NA), na.rm = TRUE) # vs 
```





# Suggestions

- Define functions for tasks that you do 2 or more times
- Use informative names (verbs)
- Use consistent return values
