---
layout: page
title: "R scripts"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
header-includes:
- \usepackage{blkarray}
- \usepackage{amsmath}
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](23-rscripts.R)

# Tidyverse style guide

What we will cover here is basically the 
[tidyverse style guide](https://style.tidyverse.org/).
Here are some additional comments.

## R scripts

It is helpful to organize the content in your R script. 
Here are some recommendations.

### Include meta-data

Include meta-data about the R script including

- purpose
- author
- dates

```{r}
# This script will perform ...
#
# Author: Jarad Niemi
# Written: 09 Apr 2023
# Modified: 10 Apr 2023 by Alex Wold
```

### Include packages

Include all packages necessary for the script as `library()` calls at the 
top of the script (after meta-data).
Generally put packages in decreasing order of important, 
i.e. most used packages [in the script] go first and least used packages go last. 

```{r}
library("tidyverse"); theme_set(theme_bw())
library("Sleuth3")
library("ggResidpanel")
library("emmeans")
```

### Source other files

You may need to source other R scripts within your R script. 
This step, if needed, should be done right after loading packages.

```{r, eval=FALSE, comment=""}
source("01-read_data.R")
source("02-wrangle_data.R")
```

### Keep only necessary commands

Only put necessary commands in the R script. 
If there are other steps that you perform, e.g. `summary()` of the data,
leave these out of the R script. 
You can, and should, run these interactively, but they should not be included
in the R script. 

### Put functions in a separate file

A good way to write R, a functional programming language, is to write 
functions for performing actions. 
Each function should be in its own R script whose name is the same as the 
function. 
These files can contain helper functions, but only if their sole purpose is for
functions within this file. 

For functions, utilize 
[roxygen2 documentation](https://r-pkgs.org/man.html). 
