\documentclass[12pt]{article}
\usepackage[margin=2cm]{geometry}
\usepackage{hyperref,verbatim}
\hypersetup{colorlinks=true,urlcolor=black}
\begin{document}

{\LARGE
\begin{tabular}{@{}l}
Iowa State University \\
Department of Statistics \\
STAT 401A, Statistical Methods for Research Workers \\
Fall 2014 \\
\end{tabular}
} % End \LARGE

\bigskip

\begin{tabular}{@{}ll@{\hspace{.2in}}ll}
Instructor: &Jarad Niemi & Office: &Snedecor 2208 \\
Email: &\href{mailto:niemi@iastate.edu}{\texttt{niemi@iastate.edu}} & Phone: & 515.294.8679 \\
Course hours: & MWF 8:00-8:50 in Sweeney 1134 & Office hours: & M 9-11 \\
& M 2:10-4:00 @ Gilman 2272 && @ Snedecor 2208\\
TA: & Nehemias Ulloa & Office hours: & T 1-3 \\
& \href{mailto:nulloa1@iastate.edu}{\texttt{nulloa1@iastate.edu}} &&@ Snedecor 1205 \\
\\
\multicolumn{2}{@{}l}{Course webpage: on Blackboard Learn and \href{http://jarad.me/stat401A}{jarad.me/stat401A}} \\
\multicolumn{2}{@{}l}{Textbook:} \\
\multicolumn{4}{l}{\hspace{0.1in}
\href{http://www.amazon.com/The-Statistical-Sleuth-Methods-Analysis/dp/1133490670/ref=sr_1_2?ie=UTF8&qid=1343246343&sr=8-2&keywords=statistical+sleuth}{
{\normalsize Ramsey and Shafer (2012). The Statistical Sleuth. Duxbury Press. 3rd ed. }}} \\
\multicolumn{2}{@{}l}{Prerequisites:} \\
\multicolumn{4}{l}{\hspace{0.1in} STAT 101, 104, 105, or 226}
\end{tabular}
\bigskip
\subsubsection*{Course description}
Graduate students without an equivalent [prerequisite] course should contact the department. Methods of analyzing and interpreting experimental and survey [observational] data. Statistical concepts and models; estimation; hypothesis tests with continuous and discrete data; simple and multiple linear regression and correlation; introduction to analysis of variance and blocking. Nonmajor graduate credit.
\subsubsection*{Course objectives}
\begin{itemize}
\item Understand variation and its consequences for drawing conclusions from data.
\item How to use statistical methods, e.g. t-tests, rank sum tests, ANOVA, regression, etc., to analyze experimental and observational data.
\item Perform, check assumptions, and interpret multiple linear regression.
%\item Appropriately perform and interpret logistic and Poisson regression.
\item Write, interpret, and critically evaluate statements such as
\begin{itemize}
\item
\begin{quote}
No significant differences were observed between A/H5N1$_{\mbox{HA N182K}}$- and \\A/H5N1$_{\mbox{HA Q222L,G2245}}$-inoculated animals, as calculated by comparing the viral titer (Mann-Whitney test, P=0.589 and 0.818 for nose and throat titers, respectively).
\end{quote}
\item
\begin{quote}
... estimate gross carbon emissions across tropical regions between 2000 and 2005 as 0.81 petagram of carbon per year, with a 90\% prediction interval of 0.57 to 1.22 petagrams of carbon per year.
\end{quote}
\end{itemize}
\end{itemize}
\subsubsection*{Assessment}
\begin{center}
\begin{tabular}{ccc}
Component & Percentage (\%) \\
\hline
Homework (weekly) & 20 \\
Midterm I & 25 \\
Midterm II & 25 \\
Final exam & 30 \\
\hline
\end{tabular}
\end{center}
\subsubsection*{Tentative schedule}
You are expected to read the appropriate chapters prior to their discussion in class.
\begin{center}
\begin{tabular}{ccl}
Week & Chapter(s) & Topic \\
\hline
1 & 1 & Statistical Inference \\
2 & 2 & Two-sample t-tests \\
3 & 3 & T-test assumptions \\
4 & 4 & Nonparametric approaches \\
\hline
5 & & {\bf MIDTERM I} (Sep 24) \\
\hline
6 & 5 & Comparison for multiple groups \\
7 & 6 & Linear combinations and multiple comparisons \\
8 & 7 & Linear regression \\
9 & 8 & Linear regression assumptions \\
\hline
10 & & {\bf MIDTERM II} (Oct 29) \\
\hline
11 & 9 & Multiple regression \\
12 & 10 & Inference in multiple regression \\
13 & 11,12 & Dealing with many explanatory variables \\
14 & 13 & Two-way ANOVA \\
15 & 20,21,22 & Data with a count response (intro) \\
\hline
&& {\bf FINAL EXAM} (Dec 19 7:30-9:30) \\
\hline
\end{tabular}
\end{center}
\subsubsection*{Academic Dishonesty, Disability Accommodation, Dead Week, Harassment and Discrimination, Religious Accommodation}
This course abides by the Faculty Senate Recommendations provided at \url{http://www.celt.iastate.edu/teaching/syllabus_inclusion.html}.
\end{document}
