---
layout: page
title: STAT 486/586
tagline: "Basic R - ggplot2"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](05-ggplot2.R)

We will start with loading up the necessary packages using the `library()` 
command.

```{r}
library("tidyverse")
```

These slides utilize the 
[ggplot2](https://ggplot2-book.org/) 
graphics system to construct a number of graphical statistics, 
i.e. data visualizations, 
for data sets ranging from a single variable to a small number of 
continuous and categorical variables. 
The first step in constructing ggplot2 graphics is to organize your data in a 
`data.frame`. 

# data.frame

Here we introduce the basics of `data.frame`s as extensions of 
matrices. 
As a reminder, matrices can only contain 1 data type in the entire matrix.
In contrast, `data.frame`s can only contain 1 data type **per column** and 
thus make `data.frame`s an ideal storage unit for basic data sets. 

Many data sets exist within the `datasets` package that is automatically 
installed when you install R. 

```{r}
data()
```

## ToothGrowth

We'll take a look at the `ToothGrowth` data set. 
Since this data set is built into R, 
there is a helpfile with useful information about the data set. 

```{r, eval=FALSE}
?ToothGrowth
```

A variety of utility functions are available for assessing aspects of 
`data.frame` objects including

```{r}
dim(ToothGrowth)
head(ToothGrowth)
tail(ToothGrowth)
summary(ToothGrowth)
```

# ggplot2

As described in 
[Section 3.2.3 of R4DS](https://r4ds.had.co.nz/data-visualisation.html#a-graphing-template),
the basic structure ggplot2 code is 

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

or

```{r, eval=FALSE}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + 
  <GEOM_FUNCTION>()
```

In both cases, you need to determine ```<DATA>```, 
```<GEOM_FUNCTION>```, and ```<MAPPINGS>```. 
In the former case, 
the mapping is only applied to that ```<GEOM_FUNCTION>```.
In the latter case, 
the mapping is applied to all future ```<GEOM_FUNCTION>```s. 
I tend to default to the latter. 



## Univariate plots

Let's start by considering the response variable in this data set,
`len`, i.e. the length of the odontoblasts. 

### Histogram

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram()
```

As instructed, let's choose the number of bins. 

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram(bins = 20)
```
Or set the binwidth (which is often more convenient)

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram(binwidth = 1)
```


### Boxplot

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_boxplot()
```

### Density plot

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_density()
```

### Violin plot

```{r}
ggplot(ToothGrowth, aes(x = len, y = 1)) + # note the y = 1
  geom_violin()
```


## Multiple univariate plots

### Violin plots

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin()
```

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin(trim = FALSE)
```

### Combine geoms

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin(trim = FALSE) + 
  geom_point()
```

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin(trim = FALSE) + 
  geom_jitter()
```

If we want the plot vertical rather than horizontal, 
just change the ```<MAPPINGS>```.

```{r}
ggplot(ToothGrowth, aes(x = supp, y = len)) + 
  geom_violin(trim = FALSE) + 
  geom_jitter()
```

Alternatively, you can use the `coord_flip()` function:

```{r}
ggplot(ToothGrowth, aes(y = supp, x = len)) + 
  geom_violin(trim = FALSE) + 
  geom_jitter() + 
  coord_flip()
```

### Alternative DATA

Let's modify the ```<DATA>``` argument using the `chickwts` data set. 

```{r}
ggplot(chickwts, aes(x = feed, y = weight)) + 
  geom_violin(trim = TRUE) + 
  geom_jitter() 
```



## Scatterplots

Recall that the tooth growth experiment included a dose that we have, so far,
ignored. 

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len)) + 
  geom_point()
```

You have already seen the use of the `geom_jitter()` function, 
but here is another option. 

### Jitterplots

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len)) + 
  geom_jitter(width=0.1)
```

### Logarithmic axis

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len)) + 
  geom_jitter(width=0.01) +
  scale_x_log10()
```

### Colors and shapes

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len, color = supp, shape = supp)) + 
  geom_jitter(width=0.01) +
  scale_x_log10()
```

### Regression line

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len, color = supp, shape = supp)) + 
  geom_jitter(width=0.01) +
  scale_x_log10() +
  geom_smooth(method = "lm")
```

Are these lines parallel?

```{r}
m <- lm(len ~ dose*supp, data = ToothGrowth)
anova(m)
```

## Professional graphics

To construct professional looking graphics,
we need to modify and customizing much of the look of these default graphics.
We will use a combination of modifying the underlying data as well as 
modify the output graphic. 

### Labels

Improve underlying data. 

```{r}
d <- ToothGrowth %>%
  rename(
    Length = len,
    Delivery = supp,
    Dose = dose) %>%
  mutate(Delivery = fct_recode(Delivery, 
                               "Orange Juice" = "OJ",
                               "Ascorbic Acid" = "VC"))
```



```{r}
ggplot(d,                       # Note the new data.frame
       aes(
         x = Dose, 
         y = Length, 
         color = Delivery, 
         shape = Delivery)) + 
  geom_point(position = position_jitterdodge(
    jitter.width = 0.005,
    dodge.width = 0.01           # moves Delivery method left and right of Dose
  )) +
  scale_x_log10() +
  scale_color_manual(
    values = c("Orange Juice" = "orange",
               "Ascorbic Acid" = "blue")
  ) +
  geom_smooth(method = "lm") + 
  labs(
    x = "Dose (mg/day)", 
    y = "Length (\u00b5m)", # unicode \u005b is the Greek letter mu
    title = "Odontoblast length vs Vitamin C in Guinea Pigs",
    color = "Delivery Method",
    shape = "Delivery Method") +
  theme_bw() +
  theme(legend.position = c(0.8, 0.2),
        legend.background = element_rect(fill = "white",
                                        color = "black"))
```

### Save the graphic

```{r}
# Save to an object
g <- ggplot(d,                       # Note the new data.frame
       aes(
         x = Dose, 
         y = Length, 
         color = Delivery, 
         shape = Delivery)) + 
  geom_point(position = position_jitterdodge(
    jitter.width = 0.005,
    dodge.width = 0.01
  )) +
  scale_x_log10() +
  scale_color_manual(
    values = c("Orange Juice" = "orange",
               "Ascorbic Acid" = "blue")
  ) +
  geom_smooth(method = "lm") + 
  labs(
    x = "Dose (mg/day)", 
    y = "Length (\u00b5m)", # unicode \u005b is the Greek letter mu
    title = "Odontoblast length vs Vitamin C in Guinea Pigs",
    color = "Delivery Method",
    shape = "Delivery Method") +
  theme_bw() +
  theme(legend.position = c(0.8, 0.2),
        legend.background = element_rect(fill = "white",
                                        color = "black"))

# Write to a file
fn <- "toothgrowth.png"
ggsave(fn, plot = g, width = 6, height = 4)

# Clean up
unlink(fn) # remove the file
```

## Diamonds data set

```{r, eval=FALSE}
?diamonds
```

What kind of informative plot can you construct?

```{r}
ggplot(diamonds, 
       aes(
         x = carat,
         y = price,
         color = depth)) +
  scale_color_gradientn(colors = rainbow(5)) +
  geom_point(alpha = 0.2) +
  facet_grid(color ~ cut) + 
  scale_y_log10() + 
  scale_x_log10() + 
  theme_bw()
```


Multiple regression model (all interactions)

```{r}
m <- lm(log(price) ~ log(carat) * depth * cut * color, data = diamonds)
drop1(m, test = "F")
```

## Additional resources

- [ggplot2 book](https://ggplot2-book.org/)
- [ggplot2 gallery](https://r-graph-gallery.com/)