---
layout: page
title: STAT 486/586
tagline: "Basic R - ggplot2"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](05-ggplot2.R)

We will start with loading up the necessary packages using the `library()` 
command.

```{r}
library("tidyverse")
```

These slides utilize the 
[ggplot2](https://ggplot2-book.org/) 
graphics system to construct a number of graphical statistics, 
i.e. data visualizations, 
for data sets ranging from a single variable to a small number of 
continuous and categorical variables. 
The first step in constructing ggplot2 graphics is to organize your data in a 
`data.frame`. 

# data.frame

Here we introduce the basics of `data.frame`s as extensions of 
matrices. 
As a reminder, matrices can only contain 1 data type in the entire matrix.
In contrast, `data.frame`s can only contain 1 data type **per column** and 
thus make `data.frame`s an ideal storage unit for basic data sets. 

Many data sets exist within the `datasets` package that is automatically 
installed when you install R. 

```{r}
data()
```

## ToothGrowth

We'll take a look at the `ToothGrowth` data set. 
Since this data set is built into R, 
there is a helpfile with useful information about the data set. 

```{r, eval=FALSE}
?ToothGrowth
```

A variety of utility functions are available for assessing aspects of 
`data.frame` objects including

```{r}
dim(ToothGrowth)
head(ToothGrowth)
tail(ToothGrowth)
summary(ToothGrowth)
```

# ggplot2

As described in 
[Section 3.2.3 of R4DS](https://r4ds.had.co.nz/data-visualisation.html#a-graphing-template),
the basic structure ggplot2 code is 

```{r, eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

or

```{r, eval=FALSE}
ggplot(data = <DATA>, mapping = aes(<MAPPINGS>)) + 
  <GEOM_FUNCTION>()
```

In both cases, you need to determine ```<DATA>```, 
```<GEOM_FUNCTION>```, and ```<MAPPINGS>```. 
In the former case, 
the mapping is only applied to that ```<GEOM_FUNCTION>```.
In the latter case, 
the mapping is applied to all future ```<GEOM_FUNCTION>```s. 
I tend to default to the latter. 



## Univariate plots

Let's start by considering the response variable in this data set,
`len`, i.e. the length of the odontoblasts. 

### Histogram

ChatGPT: 
A histogram is a graphical representation of data distribution. It is a bar graph that represents the frequency of occurrence of continuous variables in the form of ranges or bins. The height of each bar represents the number of data points that fall within the range represented by that bin. Histograms are useful for analyzing the distribution of numerical data and identifying any patterns or trends in the data, such as skewness, outliers, and modality. They also help to visualize the shape of the distribution, which can be symmetrical, skewed, or multimodal. In addition, histograms are used in a wide range of applications, including image processing, signal processing, and statistical analysis.

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram()
```

As instructed, let's choose the number of bins. 

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram(bins = 20)
```
Or set the binwidth (which is often more convenient)

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram(binwidth = 1)
```

If we would like to compare a histogram to a density,
we may want to use proportion of counts rather than 
the number of counts, e.g. 

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_histogram(aes(y = ..density...), binwidth = 1) +
  stat_function(fun = dnorm,
                args = list(mean = mean(data$x),
                            sd = sd(data$x)),
                col = "#red",
                size = 5)
```


### Boxplot

ChatGPT: 
A boxplot is a type of graph commonly used in statistics to represent and summarize the distribution of a set of numerical data. It is constructed by plotting a box between the first and third quartile (the 25th and 75th percentile), and a vertical line that extends from the box to the minimum and maximum value of the data (excluding outliers). The median (50th percentile) of the data is also plotted as a horizontal line inside the box. This graphical representation provides a quick visual summary of the central tendency, variability, and skewness of the data. Outliers, which are values that lie significantly outside the range of the majority of the data, are also easily identified in a boxplot.

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_boxplot()
```

### Density plot

```{r}
ggplot(ToothGrowth, aes(x = len)) + 
  geom_density()
```

### Violin plot

```{r}
ggplot(ToothGrowth, aes(x = len, y = 1)) + # note the y = 1
  geom_violin()
```


## Multiple univariate plots

### Violin plots

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin()
```

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin(trim = FALSE)
```

### Combine geoms

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin(trim = FALSE) + 
  geom_point()
```

```{r}
ggplot(ToothGrowth, aes(x = len, y = supp)) + 
  geom_violin(trim = FALSE) + 
  geom_jitter()
```

If we want the plot vertical rather than horizontal, 
just change the ```<MAPPINGS>```.

```{r}
ggplot(ToothGrowth, aes(x = supp, y = len)) + 
  geom_violin(trim = FALSE) + 
  geom_jitter()
```

Alternatively, you can use the `coord_flip()` function:

```{r}
ggplot(ToothGrowth, aes(y = supp, x = len)) + 
  geom_violin(trim = FALSE) + 
  geom_jitter() + 
  coord_flip()
```

### Alternative DATA

Let's modify the ```<DATA>``` argument using the `chickwts` data set. 

```{r}
ggplot(chickwts, aes(x = feed, y = weight)) + 
  geom_violin(trim = TRUE) + 
  geom_jitter() 
```



## Scatterplots

Recall that the tooth growth experiment included a dose that we have, so far,
ignored. 

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len)) + 
  geom_point()
```

You have already seen the use of the `geom_jitter()` function, 
but here is another option. 

### Jitterplots

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len)) + 
  geom_jitter(width=0.1)
```

### Logarithmic axis

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len)) + 
  geom_jitter(width=0.01) +
  scale_x_log10()
```

### Colors and shapes

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len, color = supp, shape = supp)) + 
  geom_jitter(width=0.01) +
  scale_x_log10()
```

### Regression line

```{r}
ggplot(ToothGrowth, aes(x = dose, y = len, color = supp, shape = supp)) + 
  geom_jitter(width=0.01) +
  scale_x_log10() +
  geom_smooth(method = "lm")
```

Are these lines parallel?

```{r}
m <- lm(len ~ dose*supp, data = ToothGrowth)
anova(m)
```

## Professional graphics

To construct professional looking graphics,
we need to modify and customizing much of the look of these default graphics.
We will use a combination of modifying the underlying data as well as 
modify the output graphic. 

### Labels

Improve underlying data. 

```{r}
d <- ToothGrowth %>%
  rename(
    Length = len,
    Delivery = supp,
    Dose = dose) %>%
  mutate(Delivery = fct_recode(Delivery, 
                               "Orange Juice" = "OJ",
                               "Ascorbic Acid" = "VC"))
```



```{r}
ggplot(d,                       # Note the new data.frame
       aes(
         x = Dose, 
         y = Length, 
         color = Delivery, 
         shape = Delivery)) + 
  geom_point(position = position_jitterdodge(
    jitter.width = 0.005,
    dodge.width = 0.01           # moves Delivery method left and right of Dose
  )) +
  scale_x_log10() +
  scale_color_manual(
    values = c("Orange Juice" = "orange",
               "Ascorbic Acid" = "blue")
  ) +
  geom_smooth(method = "lm") + 
  labs(
    x = "Dose (mg/day)", 
    y = "Length (\u00b5m)", # unicode \u005b is the Greek letter mu
    title = "Odontoblast length vs Vitamin C in Guinea Pigs",
    color = "Delivery Method",
    shape = "Delivery Method") +
  theme_bw() +
  theme(legend.position = c(0.8, 0.2),
        legend.background = element_rect(fill = "white",
                                        color = "black"))
```

### Save the graphic

```{r}
# Save to an object
g <- ggplot(d,                       # Note the new data.frame
       aes(
         x = Dose, 
         y = Length, 
         color = Delivery, 
         shape = Delivery)) + 
  geom_point(position = position_jitterdodge(
    jitter.width = 0.005,
    dodge.width = 0.01
  )) +
  scale_x_log10() +
  scale_color_manual(
    values = c("Orange Juice" = "orange",
               "Ascorbic Acid" = "blue")
  ) +
  geom_smooth(method = "lm") + 
  labs(
    x = "Dose (mg/day)", 
    y = "Length (\u00b5m)", # unicode \u005b is the Greek letter mu
    title = "Odontoblast length vs Vitamin C in Guinea Pigs",
    color = "Delivery Method",
    shape = "Delivery Method") +
  theme_bw() +
  theme(legend.position = c(0.8, 0.2),
        legend.background = element_rect(fill = "white",
                                        color = "black"))

# Write to a file
fn <- "toothgrowth.png"
ggsave(fn, plot = g, width = 6, height = 4)

# Clean up
unlink(fn) # remove the file
```

## Diamonds data set

```{r, eval=FALSE}
?diamonds
```

What kind of informative plot can you construct?

```{r}
ggplot(diamonds, 
       aes(
         x = carat,
         y = price,
         color = depth)) +
  scale_color_gradientn(colors = rainbow(5)) +
  geom_point(alpha = 0.2) +
  facet_grid(color ~ cut) + 
  scale_y_log10() + 
  scale_x_log10() + 
  theme_bw()
```


Multiple regression model (all interactions)

```{r}
m <- lm(log(price) ~ log(carat) * depth * cut * color, data = diamonds)
drop1(m, test = "F")
```

## Additional resources

- [ggplot2 book](https://ggplot2-book.org/)
- [ggplot2 gallery](https://r-graph-gallery.com/)