---
layout: page
title: STAT 486/586
tagline: "Logical Operations"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](09-logical.R)



```{r}
library("tidyverse")
```

# TRUE/FALSE

**Always use TRUE/FALSE and not T/F!**

```{r}
T
F
TRUE
FALSE
```

T and TRUE appear the same. 

```{r}
isTRUE(T)
isTRUE(TRUE)
```

But T/F can be reassigned. 

```{r}
T <- FALSE

isTRUE(T)
```

TRUE/FALSE cannot be reassigned.

```{r, error=TRUE}
TRUE <- FALSE
```

The reason is because R has a short list of reserved words

```{r}
?Reserved
```



# NOT (!)

```{r NOT}
!TRUE
!FALSE
```

# OR (|)

```{r OR}
TRUE | TRUE
TRUE | FALSE
FALSE | TRUE
FALSE | FALSE
```

# AND (&)

```{r AND}
TRUE & TRUE
TRUE & FALSE
FALSE & TRUE
FALSE & FALSE
```

# XOR (xor)

```{r XOR}
xor(TRUE, TRUE)
xor(TRUE, FALSE)
xor(FALSE, TRUE)
xor(FALSE, FALSE)
```

# Multiple comparisons

## Elementwise comparisons

The operations above perofrm elementwise comparisons.

```{r}
b <- c(TRUE,FALSE)
x <- rep(b, times = 2)
y <- rep(b, each = 2)
x
y

x | y
x & y
xor(x,y)
!x
```

## Left-to-right comparison

If you want left-to-right comparisons, 
you can use `&&` and `||`.

```{r}
TRUE && TRUE && TRUE
TRUE && TRUE && FALSE
TRUE && FALSE && TRUE
FALSE && FALSE && FALSE

TRUE || TRUE || TRUE
FALSE || FALSE || TRUE
FALSE || FALSE || FALSE
```

You will get a warning if you try to pass a vector to these 
left-to-right comparisons. 

```{r, error=TRUE}
c(TRUE,FALSE) && c(TRUE,TRUE)
```

## any()

```{r}
any(c(TRUE,   TRUE,  TRUE,  TRUE))
any(c(TRUE,  FALSE, FALSE, FALSE))
any(c(FALSE, FALSE, FALSE, FALSE))
```

## all()

```{r}
all(c(TRUE,   TRUE,  TRUE,  TRUE))
all(c(TRUE,  FALSE, FALSE, FALSE))
all(c(FALSE, FALSE, FALSE, FALSE))
```


# Comparisons

## Character

```{r}
"a" == "a"
"a" != "a"
"a" == ""
"a" != "b"
```

### Partial matching

```{r}
grepl("a", "abcd")
grepl("a", letters)
grepl("a", LETTERS, ignore.case = TRUE)
```




## Numeric

```{r}
a <- 1
b <- 2

a < b
a > b
a <= b
a >= b
```

### Equality

```{r}
a == 1
a == 2
a != 1
a != 2

a == b
a != b
```

Be careful with double precision 

```{r}
sin(0)    == 0
sin(2*pi) == 0
```

```{r}
abs(sin(2*pi) - 0) < .Machine$double.eps^0.5
```

```{r}
near(sin(2*pi), 0)
```




# Force TRUE/FALSE

Not all seemingly logical functions return TRUE/FALSE.

```{r}
all.equal(1,1)
all.equal(1,2)

all(c(TRUE, FALSE))
all(c(TRUE, FALSE, NA))
all(c(TRUE, TRUE,  NA))
```

This can cause issues in flow control

```{r, error=TRUE}
if (all(c(TRUE, TRUE,  NA))) {
  # Do something
}
```

## isTRUE()

The `isTRUE()` and `isFALSE()` functions ALWAYS resolve to TRUE/FALSE. 

```{r}
isTRUE(TRUE)
isTRUE(NA)
isTRUE(1)
isTRUE(as.logical(NA))
isTRUE(as.logical(1))  # all non-zero numeric values are coerced to TRUE
isTRUE(as.logical(2))  
isTRUE(as.logical(1e-16))  

isFALSE(FALSE)
isFALSE(NA)
isFALSE(1)
isFALSE(as.logical(NA))
isFALSE(as.logical(0))  # only 0 is coerced to FALSE
isFALSE(as.logical(2)) 
```

```{r}
isTRUE(all.equal(1,2))
isFALSE(all.equal(1,2))

isFALSE(!all.equal(1,1))

isTRUE(NA)
isFALSE(NA)
```
