---
layout: page
title: "R/RStudio Installation"
output: 
  html_document:
    toc: TRUE
---


Although R and RStudio will be available in the lab, you will likely want to 
install R and RStudio on your own personal laptop, desktop, or both. 
Below are installation instructions for R and RStudio. 

## Install R (may require administrator privileges)

Please go to <http://www.r-project.org/> and click on 
[download R](http://cran.r-project.org/mirrors.html). You will be asked to 
select a [CRAN mirror](http://cran.r-project.org/mirrors.html). I select 
<https://mirror.las.iastate.edu/CRAN/> for obvious reasons. 
In the download and install R section, click on 
the correct link depending on your operating system: Linux, (Mac) OS X, or 
Windows. 

### Windows

Click on **base** and then the link to Download R X.X.X for Windows (where the 
X will be numbers representing the current version of R). Install this program 
like any other program on Windows using all the defaults.

### (Mac) OS X

Click on the appropriate **.pkg** file depending on which version of Mac OS X 
you are using. 

Install this program like any other program on OS X using all the 
defaults. 

### Linux

If you are using Linux, then I trust you know what you are doing. You can 
install R from source or you can use a package manager. 


## Install RStudio (may require administrator privileges)

The installation of RStudio is optional, but highly recommended. 
It provides an improved interface to R and has 
[projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects) 
which help to quickly switch between (oddly enough) projects. 
Go to <http://www.rstudio.com/products/rstudio/download/> and choose the correct 
platform under **Installers for ALL Platforms**. 
Install like any other program. 

Now, whenever you want to run R, just run RStudio. 


## Install packages

### Install all necessary packages

The last thing you need to do is to install all the packages that are necessary
for this course. 
To do so, run the following command in the Console in RStudio. 

```{r, message=FALSE, eval=FALSE}
install.packages(c("tidyverse","gridExtra")) # you will need to uncomment this line in the script
```

This process will take about a minute (depending on your internet connection). 

### Background

The base installation of R provides access to many graphical and statistical
methods. 
But a great many more are available as R packages. 
The main repository for R packages is the 
[Comprehensive R Archive Network (CRAN)](https://cran.r-project.org/web/packages/available_packages_by_name.html). 

To install a package from CRAN, you can use the function `install.packages()`. 
For example, to install the `tidyverse` package (which you definitely want to do) 
use 

```{r, eval=FALSE, purl=FALSE}
install.packages("tidyverse")
```

Packages only need to be installed once (until you update R) so you will only 
have to run `install.packages()` once for each package you need. 

By convention, any packages that are needed in a script are included at the 
top of the script using the `library()` function. 
For example, 

```{r, eval=FALSE, purl=FALSE}
library("gridExtra")
```

If you try to run one of the commands above when you don't have the package
installed, you will see an error, e.g. 

```{r, error = TRUE, purl=FALSE}
library("notarealpackage")
```

If you get this error, you should run `install.packages()` using the
package name as the argument to install the package. 

