---
title: "Data Visualization I (ggplot2)"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Data Visualization I (ggplot2)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Before we get back into graphics, it is important to understand some of the 
fundamentals behind what R is doing. 

Please open the `02_graphics.R` script in your working directory. 
If you cannot find this file, you may need to do the following:

1. [change your working directory](02_intro.html#working_directory)
1. write the workshop data and scripts

```{r, eval=FALSE}
MWBDSSworkshop::workshop(write_data = TRUE, write_scripts = TRUE) 
```

## <a name="types"></a> Data types

Objects in R can be broadly classified according to their dimensions:

- scalar
- vector
- matrix
- array (higher dimensional matrix)

and according to the type of variable they contain:

- integer
- numeric
- character (string)
- logical
- factor


### Scalars

Scalars have a single value assigned to the object in R. 

```{r}
a = 3.14159265 
b = "ISDS Workshop" 
c = TRUE
```

Print the objects

```{r}
a
b
c
```


### Vectors

The `c()` function creates a vector in R

```{r}
a = c(1,2,-5,3.6)
b = c("ISDS","Workshop")
c = c(TRUE, FALSE, TRUE)
```

To determine the length of a vector in R use `length()`

```{r}
length(a)
length(b)
length(c)
```

To determine the type of a vector in R use `class()`

```{r}
class(a)
class(b)
class(c)
```


#### Vector construction

Create a numeric vector that is a sequence using : or `seq()`.

```{r}
1:10
5:-2
seq(from = 2, to = 5, by = .05)
```

Another useful function to create vectors is `rep()`

```{r}
rep(1:4, times = 2)
rep(1:4, each  = 2)
rep(1:4, each  = 2, times = 2)
```

Arguments to functions in R can be referenced either by position or by name or
both. 
The safest and easiest to read approach is to name all your arguments.
I will often name all but the first argument. 

#### Accessing vector elements 

Elements of a vector can be accessed using brackets, e.g. [index].

```{r}
a = c("one","two","three","four","five")
a[1]
a[2:4]
a[c(3,5)]
a[rep(3,4)]
```

Alternatively we can access elements using a logical vector where only TRUE 
elements are accessed.

```{r}
a[c(TRUE, TRUE, FALSE, FALSE, FALSE)]
```

You can also remove elements using a negative sign `-`.

```{r}
a[-1]
a[-(2:3)]
```

#### Modifying elements of a vector

You can assign new values to elements in a vector using = or <-.

```{r}
a[2] = "twenty-two"
a
a[3:4] = "three-four" # assigns "three-four" to both the 3rd and 4th elements
a
a[c(3,5)] = c("thirty-three","fifty-five")
a
```



### Matrices

Matrices can be constructed using `cbind()`, `rbind()`, and `matrix()`:

```{r}
m1 = cbind(c(1,2), c(3,4))       # Column bind
m2 = rbind(c(1,3), c(2,4))       # Row bind

m1
all.equal(m1, m2)

m3 = matrix(1:4, nrow = 2, ncol = 2)
all.equal(m1, m3)

m4 = matrix(1:4, nrow = 2, ncol = 2, byrow = TRUE)
all.equal(m3, m4)

m3
m4
```


#### Accessing matrix elements

Elements of a matrix can be accessed using brackets separated by a comma, 
e.g. [row index, column index].

```{r}
m = matrix(1:12, nrow=3, ncol=4)
m
m[2,3]
```

Multiple elements can be accessed at once

```{r}
m[1:2,3:4]
```

If no row (column) index is provided, then the whole row (column) is accessed.

```{r}
m[1:2,]
```

Like vectors, you can eliminate rows (or columns)

```{r}
m[-c(3,4),]
```

Be careful not to forget the comma, e.g. 

```{r}
m[1:4]
```

You can also construct an object with more than 2 dimensions using the 
`array()` function.

### Cannot mix types 

You cannot mix types within a vector, matrix, or array

```{r}
c(1,"a")
```

The number 1 is in quotes indicating that R is treating it as a character rather 
than a numeric.

```{r}
c(TRUE, 1, FALSE)
```

The logicals are converted to numeric (0 for FALSE and 1 for TRUE).

```{r}
c(TRUE, 1, "a")
```

Everything is converted to a character. 

### Activity

Reconstruct the following matrix using the `matrix()` function, then 

1. Print the element in the 3rd-row and 4th column
1. Print the 2nd column
1. Print all but the 3rd row

```{r}
m = rbind(c(1, 12, 8, 6),
          c(4, 10, 2, 9),
          c(11, 3, 5, 7))
m
```

```{r, echo=FALSE}
# Reconstruct the matrix 

# Print the element in the 3rd-row and 4th column

# Print the 2nd column

# Print all but the 3rd row
```


When you have completed the activity, compare your results to the 
[solutions](02_graphics-solution.html#matrix).




## <a name="frames"></a> Data frames

A `data.frame` is a special type of matrix that allows different data types in 
different columns.

We have already seen a `data.frame` with our GI data set. 
Let's read this data in again and take a look.


```{r}
GI = read.csv("GI.csv")
dim(GI)
```


### Access `data.frame` elements

A `data.frame` can be accessed just like a matrix, 
e.g. [row index, column index].

```{r}
GI[1:2, 3:4]
```

`data.frame`s can also be accessed by column names

```{r}
GI[1:2, c("facility","icd9","gender")]
```

or 

```{r, message=FALSE}
library('dplyr') 
GI %>% 
  select(facility, icd9, gender) %>%
  head(n = 2)
```

The `%>%` (pipe) operator allows chaining of commands by passing the result of
the previous command as the first argument of the next command. 
This makes code much easier to read. 
Two equivalent approaches that are harder to read are 

```{r, eval=FALSE}
# Approach 1
head(select(GI, facility, icd9, gender), n = 2)

# Approach 2
GI_select <- select(GI, facility, icd9, gender)
head(GI_select, n = 2)
```

When there are long strings of commands, using the `%>%` (pipe) operator makes
code much easier to read. 
See [here](https://www.r-statistics.com/2014/08/simpler-r-coding-with-pipes-the-present-and-future-of-the-magrittr-package/) 
for more background and information. 



### Different data types in different columns

The function `str()` allows you to see the structure of any object in R. 
Using `str()` on a `data.frame` object tells you 

1. that the object is a `data.frame`,
1. the number of rows and columns,
1. the names of each column,
1. each column's data type, and
1. the first few elements in that column.

```{r}
str(GI)
```



### <a name="factor"></a> Factor

A factor is a data type that represents a categorical variable. 

The default is for any character vector to be converted to a factor when read 
using `read.csv()` or `read.table()`. 
You can change this behavior by setting `stringsAsFactors = FALSE`, 
which is the default in `readr::read_csv()`.

Internally, R codes a factor as an integer and then keeps a table that contains 
the conversion from that integer into the actual value of the factor.

```{r}
nlevels(GI$gender)
levels(GI$gender)          # internal table
GI$gender[1:3]
as.numeric(GI$gender[1:3]) # internal coding
```



#### Converting a numeric variable into a factor

When a categorical variable is encoded as a numeric variable in the original 
data set, R reads them in as numeric. 
To convert them to a factor use `as.factor()` or `factor()`.

```{r}
GI$facility = as.factor(GI$facility)
summary(GI$facility)
```


#### Converting back to the original numeric variable

To obtain the original numeric variable use `as.character()` and `as.numeric()`

```{r}
head(as.character(GI$facility))             # This returns the levels as a character vector
head(as.numeric(as.character(GI$facility))) # This returns the original numeric factor levels
```


#### Creating your own factor

Use the `cut()` function to create a factor from a continuous variable.

```{r}
GI$ageC = cut(GI$age, c(-Inf, 5, 18, 45 ,60, Inf)) # Inf is infinity
table(GI$ageC)
```

This created a new variable in the GI `data.frame` called *ageC*. 






### <a name="dates"></a> Dates

In order to use dates properly, they need to be converted into type `Date`. 

```{r}
GI$date = as.Date(GI$date)
str(GI$date)
```

`as.Date()` will attempt to read dates as "%Y-%m-%d" then "%Y/%m/%d". 
If neither works, it will give an error. 

```{r, eval=FALSE}
?as.Date
```

You can specify other date patterns, e.g. 

```{r, eval=FALSE}
as.Date("12/09/14", format="%m/%d/%y")
```

For those who work with dates often, check out the 
[lubridate](https://cran.r-project.org/web/packages/lubridate/index.html)
package. 


### Activity 

Create a new variable in the GI data set called `icd9code` that cuts icd9 at 
0, 140, 240, 280, 290, 320, 360, 390, 460, 520, 580, 630, 680, 710, 740, 760, 
780, 800, 1000, and Inf. 
Find the `icd9code` that is the most numerous in the GI data set.

```{r, eval=FALSE}
# Create icd9code

# Find the icd9code that is most numerous
```

When you have completed the activity, compare your results to the 
[solutions](02_graphics-solution.html#factor).




## <a name="reshaping"></a> Reshaping data frames in R

There are two general representations of tabular data.

Wide:
```{r, echo=FALSE}
d = data.frame(week = 1:3, 
               GI = c(246,195,212), 
               ILI = c(948, 1020, 1024))
d
```

which is a succinct representation of the data

Long:
```{r, echo=FALSE}
d %>% 
  tidyr::gather(key = syndrome, value = count, -week)
```

which is the form most statistical software wants, 
i.e. there is only one column for the response (count).


### Wide to long

The `tidyr` package provides functions to convert between the two 
representations. First, we need to load the package

```{r}
library('tidyr')
```

Create the wide `data.frame`:
```{r}
d = data.frame(week = 1:3, 
               GI   = c(246,195,212), 
               ILI  = c(948, 1020, 1024))
d
```

To turn the `data.frame` into long format, use `gather()`.

```{r}
m <- d %>%
  gather(key = syndrome, # Creates a column called syndrome
         value = count,  # Creates a column called count
         -week)          # Keeps the column `week` as a column
                         # All other columns (GI and ILI) are gathered
```

This approach is useful if there are a lot of columns that need to be gathered
but only a few that need to remain as columns.
If the opposite is true, i.e. there are only a few columns to be gathered and 
a lot that need to remain as columns, use

```{r}
m2 <- d %>%
  gather(key = syndrome, # Creates a column called syndrome
         value = count,  # Creates a column called count
         GI, ILI)        # Gathers these columns

all.equal(m,m2)
```


### Long to wide

If we want to convert back, use `spread()`

```{r}
m %>%
  spread(key = syndrome, value = count)
```

I find that I use the `gather` function much more often than I use the `spread`
function because data are usually stored in a succinct format but then I need
the data in long format for summaries or figures or statistical analyses.


## <a name="aggregating"></a> Aggregating data frames in R

The GI data set that we have is already in long format and each row is an 
individual. 
We may want to aggregate this information. 
To do so, we will use the `group_by()` and `summarize()` functions in the
`dplyr` package.

```{r}
library('dplyr')
```

For example, perhaps we wanted to know the total number of GI or ILI cases 
across the 3 weeks:

```{r}
m %>%                           # We need to use the melted (long) version of the data set
  group_by(syndrome) %>%        # Do the following for each syndrome
  summarize(total = sum(count)) # Calculate `total` which is the sum of count for each syndrome
```


### Aggregating the GI data set

Let's aggregate the GI data set by week, gender, and age category.

First, we need to create weeks

```{r}
GI$date = as.Date(GI$date) # Make sure the dates are actually dates
GI$week = cut(GI$date, 
              breaks = "weeks", 
              start.on.monday = TRUE) 
```

Now we can summarize

```{r}
GI_count <- GI %>%                 # each row is a single observation
  group_by(week, gender, ageC) %>% # split the data by these variables
  summarize(total = n())           # this counts the number of rows, see ?n

nrow(GI_count)
head(GI_count, 20)
```


### Activity

Aggregate the GI data set by gender, ageC, and icd9code 
(the ones created in the last activity).


```{r, echo=FALSE}
# Aggregate our GI data set by gender, ageC, and icd9code (the ones created in the last activity).
```


When you have completed the activity, compare your results to the 
[solutions](02_graphics-solution.html#aggregate).





## <a name="ggplot2"></a> Basics of `ggplot2`

Previously we produced graphics using the base `graphics` system. 
Although I still use this for producing quick plots, 
I invariably end up using the `ggplot2` package.
This package requires a `data.frame` in long format. 


Load the `ggplot2` package
```{r}
library('ggplot2')
```



### Histogram

A basic histogram in ggplot

```{r}
ggplot(data = GI, aes(x = age)) + geom_histogram(binwidth = 1)
```

For code that looks more similar to the histogram code we saw before, 
you can use 

```{r, eval=FALSE}
qplot(data = GI, x = age, geom = "histogram", binwidth = 1)
```

Many websites and even the `ggplot2` manual have examples using `qplot`. 
I believe this is mainly to ease the transition for individuals who are familiar 
with base `graphics`.
If you are just starting out with R, 
I recommend using the `ggplot` function in `ggplot2` from the beginning.

### Boxplots

A basic boxplot

```{r}
ggplot(data = GI, aes(x = 1, y = age)) + geom_boxplot()
```


### Multiple boxplots

```{r}
ggplot(GI, aes(x = facility, y = age)) + geom_boxplot()
```



### Scatterplots

```{r}
ggplot(GI, aes(x=date, y=age)) + geom_point()
```

### Bar charts

With ggplot, there is no need to count first.

```{r}
ggplot(GI, aes(x=facility)) + geom_bar()
```

An appealing aspect of ``ggplot` is that once the data is in the correct format
it is easy to construct lots of different plots. 


### Activity


Construct a histogram and boxplot for age at facility 37 using ggplot2.

```{r, eval=FALSE}
# Construct a histogram for age at facility 37.

# Construct a boxplot for age at facility 37. 
```

Construct a bar chart for the 3-digit zipcode at facility 37 using ggplot2

```{r, eval=FALSE}
# Construct a bar chart for the 3-digit zipcode at facility 37 
```


When you have completed the activity, compare your results to the 
[solutions](02_graphics-solution.html#ggplot2).




## <a name="customizing"></a> Customizing ggplot2 plots

There are many ways to customize the appearance of ggplot2 plots:

- Colors
- Labels
- Titles
- Characters
- Line types
- Themes



### Colors

```{r}
ggplot(GI, aes(x = age)) + 
  geom_histogram(binwidth = 1, color = 'blue',   fill = 'yellow')
```

```{r}
ggplot(GI, aes(x=date, y=age)) + 
  geom_point(color = 'purple')
```

To find all the colors that R knows, use 
```{r, eval=FALSE}
colors()
```

### Labels

```{r}
ggplot(GI, aes(x = facility, y = age)) + 
  geom_boxplot() + 
  labs(x     = 'Facility ID', 
       y     = 'Age (in years)', 
       title = 'Age by Facility ID')
```


### Characters

```{r}
ggplot(GI, aes(x=date, y=age)) + geom_point(shape = 2, color = 'red')
```

ggplot2 uses the same shape codes as base graphics, see

```{r, eval=FALSE}
?points
```


### Line types

Here I am also using a trick of setting up part of the plot and assigning it to
the object `g`. 
Then you can add elements to the plot and if you don't assign it, 
the plot will be shown. 

```{r}
g = ggplot(GI %>% 
             group_by(week) %>%
             summarize(count = n()), 
           aes(x = as.numeric(week), 
               y = count)) +
  labs(x = 'Week #', 
       y = 'Weekly count')

g + geom_line()
g + geom_line(size=2, color='firebrick', linetype=2)
```

Linetype options can be found in `?par` under `lty`. 
But it is probably more informative to just google it, 
e.g. <http://www.cookbook-r.com/Graphs/Shapes_and_line_types/>.

### Themes

```{r}
g = g + geom_line(size = 1, color = 'firebrick')
g + theme_bw()
```

For other themes, see

```{r, eval=FALSE}
?theme
?theme_bw
```




## <a name="help"></a> Getting help on ggplot2

Although the general R help can still be used, e.g. 

```{r, eval=FALSE}
?ggplot
?geom_point
```

It is much more helpful to google for an answer

    geom_point 
    ggplot2 line colors

The top hits will all have the code along with what the code produces.

### Helpful sites

These sites all provide code. The first two also provide the plots that are produced.

- <http://docs.ggplot2.org/current/>
- <http://www.cookbook-r.com/Graphs/>
- <http://stackoverflow.com/questions/tagged/ggplot2>



### Activity

Play around with ggplot2 to see what kind of plots you can make.

