---
title: "Data Visualization II (ggplot2)"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Data Visualization II (ggplot2)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Close down R, say 'No' to saving the workspace.
In fact, I recommend you change your settings so that 

- the workspace is *never* stored and
- .RData is *not* restored upon startup.

## <a name="workflow"></a> Workflow

The typical workflow in R is 

1. Start R
1. Choose your working directory (choose project in RStudio)
1. Open a script
1. Read in data
1. Check data (interactively)
1. Do analysis

If you are using RStudio, I highly recommend you look into setting up [projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects). 


### Choose your working directory 

Start R and then [choose your working directory](01_intro.html#working_directory).


### Start the workshop

Now start the workshop to get your data files and scripts set up. 
Typically you would choose your working directory to be where you actually have
your data and scripts.

```{r, purl=FALSE, message=FALSE, purl=FALSE}
library('MWBDSSworkshop')
workshop(write_data = TRUE, write_scripts = TRUE, launch_index  = FALSE) 
```


### Open the script

Open the `03_advanced_graphics.R` script.

A convention is to put `library()` calls for all packages that you will use in
the script at the top of the script.

```{r, message=FALSE}
library('tidyverse') # loads dplyr/ggplot2/readr
```


### Read in the data

Typically, you will have a script that reads in the data, converts variables, 
and then performs statistical analyses.

This time we will use the `mutate()` function from the 
[dplyr](http://cran.r-project.org/web/packages/dplyr/index.html) package. 
This function allows you to create new columns in a `data.frame`.

```{r}
# Read in csv files
GI     = read.csv("GI.csv")
icd9df = read.csv("icd9.csv")

# Add columns to the data.frame
GI <- GI %>%
  mutate(
    date      = as.Date(date),
    weekC     = cut(date, breaks="weeks"), 
    week      = as.numeric(weekC),
    facility  = as.factor(facility),
    icd9class = factor(cut(icd9, 
                           breaks = icd9df$code_cutpoint, 
                           labels = icd9df$classification[-nrow(icd9df)], 
                           right  = TRUE)),
    ageC      = cut(age, 
                    breaks = c(-Inf, 5, 18, 45 ,60, Inf)),
    zip3      = trunc(zipcode/100))
```




### Reading other scripts

If the script to read in the data is extensive, it may be better to separate
your scripts into two different files: one for reading in the data and
converting variables and another (or more) to perform statistical analyses.
Then, in the second file, i.e. the one that does an analysis, you can `source()`
the one that just reads in the data, e.g.

```{r, eval=FALSE, purl=FALSE}
source("read_data.R") 
```


### Check the data

At this point, I typically check the data to make sure the `data.frame` is what
I think it is. 
I usually do this in interactive mode, i.e. not in a script.

```{r}
dim(GI)
str(GI)
summary(GI)
```


### Activity

Create a new variable in the GI data set called `weekD` that is a `Date` object
with each observation having the Monday of the week for that observation. 
Check that it is actually a date using the `str()` function.

```{r, eval=FALSE}
# Create weekD variable in GI data set
```

When you have completed the activity, compare your results to the 
[solutions](03_advanced_graphics-solution.html#Date).



## <a name="graph_workflow"></a> Graph workflow

Now to construct a graph, the workflow is

1. construct an appropriate data set and
1. construct the graph.


### Construct the data set

Suppose we want to plot the number of weekly GI cases by facility. 
We need to summarize the data by week and facility. 

```{r}
GI_wf <- GI %>%
  group_by(week, facility) %>%
  summarize(count = n())
```

In interactive mode, you should verify that this data set is correct, e.g. 

```{r, eval=FALSE}
nrow(GI_wf) # Should have number of weeks times number of facilities rows
ncol(GI_wf) # Should have 3 columns: week, facility, count
dim(GI_wf)
head(GI_wf)
tail(GI_wf)
summary(GI_wf)
summary(GI_wf$facility)
```

### Construct the graph

Now, we would like week on the x-axis and count on the y-axis. 

```{r}
ggplot(GI_wf, aes(x = week, y = count)) + 
  geom_point()
```

But, clearly we need to distinguish the facilities. 


### Try colors and shapes to distinguish facilities

Colors:

```{r}
ggplot(GI_wf, aes(x = week, y = count, color = facility)) + 
  geom_point()
```

Shapes:

```{r}
ggplot(GI_wf, aes(x = week, y = count, shape = facility)) + 
  geom_point()
```

`ggplot2` only plots 6 different shapes and provides a warning.


### Activity - weekly GI counts by age category

Construct a data set and then a plot to look at weekly GI cases by age category.

```{r, eval=FALSE}
# Construct a data set aggregated by week and age category

# Construct a plot to look at weekly GI cases by age category.
```

When you have completed the activity, compare your results to the 
[solutions](03_advanced_graphics-solution.html#graph).



## <a name="faceting"></a> Faceting

Faceting is a technique to view many small graphs on a single page.

In ggplot, there are two different ways to construct facets: 
`facet_wrap()` and `facet_grid()`. 
The former performs the faceting for a single variable and constructs the 
matrix of plots automatically. 
The latter performs faceting for two variables putting one vertically and the 
other horizontally.


### Faceting on a single variable

When there are many levels of a single variable, 
`facet_wrap()` can often allow comparisons across that variable.

```{r}
ggplot(GI_wf, aes(x = week, y = count)) + 
  geom_point() + 
  facet_wrap(~ facility)
```

By default, faceting forces all axes to be exactly the same. 
If you want the axis to automatically scale for each facet use `scales="free"`.

```{r}
ggplot(GI_wf, aes(x = week, y = count)) + 
  geom_point() + 
  facet_wrap(~ facility, scales = "free")
```

### Faceting on two variables

When there are two variables (with few levels in each), 
`facet_grid()` can allow relevant comparisons.

Suppose we are interested in weekly GI counts by gender and age category. 
First, construct the data set

```{r}
GI_sa <- GI %>%
  group_by(week, gender, ageC) %>%
  summarize(count = n())
```

Next, construct the plot using `facet_grid(row ~ column)` syntax.

```{r}
ggplot(GI_sa, aes(x = week, y = count)) + 
  geom_point() + 
  facet_grid(gender ~ ageC)
```

### Using faceting and colors/shapes

In addition, we could use faceting together with colors/shapes

```{r}
ggplot(GI_sa, aes(x = week, y = count, shape = gender, color = gender)) + 
  geom_point() + 
  facet_wrap(~ ageC)
```


### Activity - weekly GI counts by zip3 and ageC

Construct a plot of weekly GI counts by zip3 and ageC. 

```{r, eval=FALSE}
# Construct a plot of weekly GI counts by zip3 and ageC. 
```

When you have completed the activity, compare your results to the 
[solutions](03_advanced_graphics-solution.html#facet).




## <a name="subsetting"></a> Filtering

Often our data set is large and we would like to 
[filter the data](intro.html#subsetting) to focus on a particular group or 
time frame. 
You may want to filter by a

- categorical, 
- continuous (numeric), or 
- `Date` variable.

### Filtering by a categorical variable

For a categorical variable, you want to filter by 

- a single category
- a set of categories
- not in a single category
- not in a set of categories

#### Filtering by a single category

Suppose we are only interested in counts for infectious and 
parasitic diseases (IPD).

```{r}
IPD_w <- GI %>%
  filter(icd9class == "infectious and parasitic disease") %>%
  group_by(week) %>%
  summarize(count = n())

ggplot(IPD_w, aes(x = week, y = count)) + 
  geom_point()
```


#### Filtering by a set of categories

Suppose we are interested in looking at both IPD and ill-defined conditions

```{r}
conditions = levels(GI$icd9class)
conditions
```

So we want levels 1 and 3. 

```{r}
IPDp <- GI %>%
  filter(icd9class %in% levels(icd9class)[c(1,3)])

summary(IPDp$icd9class)
```

Now construct the graph

```{r}
IPDp_w <- IPDp %>%
  group_by(week, icd9class) %>%
  summarize(count = n())

ggplot(IPDp_w, aes(x = week, y = count)) + 
  geom_point() + 
  facet_wrap(~ icd9class, scales = "free")
```

#### Filtering by not in a single category

Suppose we are interested in all icd9 codes except IPD. 

```{r}
# Combine filtering and summarizing
nIPD_w <- GI %>%
  filter(icd9class != "infectious and parasitic disease") %>%
  group_by(week, icd9class) %>%
  summarize(count = n())

ggplot(nIPD_w, aes(x = week, y = count)) + 
  geom_point() + 
  facet_wrap(~ icd9class, scales = "free")
```

#### Filtering by not in a set of categories

Suppose we are interested in all icd9 codes except IPD and ill-defined 
conditions.

```{r}
nIPDp_w <- GI %>%
  filter(!(icd9class %in% levels(icd9class)[c(1,3)])) %>%
  group_by(week, icd9class) %>%
  summarize(count = n())

ggplot(nIPDp_w, aes(x = week, y = count)) + 
  geom_point() + 
  facet_wrap(~ icd9class, scales = "free")
```


### Filtering by a continuous (numeric) variable

We may want to subset a continuous (numeric) variable by

- Larger than (or equal to) a number
- Smaller than (or equal to) a number
- In a range of numbers (with our without endpoints)

#### Filtering larger than a number

```{r}
Age60p_w <- GI %>%
  filter(age > 60) %>%
  group_by(week) %>%
  summarize(count = n())

ggplot(Age60p_w, aes(x = week, y = count)) + 
  geom_point()
```

#### Filtering less than or equal to a number 

```{r}
Age_lte30_w <- GI %>%
  filter(age <= 30) %>%
  group_by(week) %>%
  summarize(count = n())

ggplot(Age_lte30_w, aes(x = week, y = count)) + 
  geom_point()
```

#### Filtering within a range

```{r}
Age30to60_w <- GI %>%
  filter(age > 30, age <= 60) %>% # This includes those who are 60, but not those who are 30
  group_by(week) %>%
  summarize(count = n())

ggplot(Age30to60_w, aes(x = week, y = count)) + 
  geom_point()
```

### Filtering a date

Filtering a date works similar to filtering a continuous (numeric) variable 
once you convert a comparison value to a `Date` using the `as.Date()` function.

Let's create a `Date` vector containing the two values we will use for 
filtering.

```{r}
two_dates = as.Date(c("2007-01-01", "2007-12-31"))
```

#### Filtering dates

```{r}
early <- GI %>% filter(date <  two_dates[1])
late  <- GI %>% filter(date >= two_dates[2])
mid   <- GI %>% filter(date >= two_dates[1], date < two_dates[2])
```

Now make plots

```{r}
early_w <- early %>% group_by(week) %>% summarize(count = n())
late_w  <- late  %>% group_by(week) %>% summarize(count = n())
mid_w   <- mid   %>% group_by(week) %>% summarize(count = n())

g1 = ggplot(early_w, aes(x = week, y = count)) + geom_point()
g2 = ggplot(late_w,  aes(x = week, y = count)) + geom_point()
g3 = ggplot(mid_w,   aes(x = week, y = count)) + geom_point()
```

To arrange multiple plots, use the `grid.arrange` function in the `gridExtra` package.

```{r}
# you may need to run install.packages("gridExtra")
if (require('gridExtra'))
  gridExtra::grid.arrange(g1,g3,g2)
```

This set of figures would have been a lot easier to construct if we had simply
created a new variable that had values early, mid, and late and then used 
faceting to create the plot, e.g.

```{r}
cut_dates <- c(as.Date("1900-01-01"),
               two_dates,
               Sys.Date())

GI_time <- GI %>% 
  mutate(time = cut(date, 
                    breaks = cut_dates, 
                    labels = c("early","mid","late"))) %>%
  group_by(week, time) %>%
  summarize(count = n())

ggplot(GI_time,
       aes(x = week, y = count)) +
  geom_point() +
  facet_wrap(~ time, ncol = 1, scales = "free")
```

### Activity - subsetted graphs

Construct a plot for those in zipcode 206xx between Jan 1, 2008 and Dec 31, 2008. 

```{r, eval=FALSE}
# Filter the data to zipcode 206xx between Jan 1, 2008 and Dec 31, 2008

# Aggregate the date for each week in this time frame

# Construct the plot of weekly GI counts in zipcode 206xx.
```

When you have completed the activity, compare your results to the 
[solutions](03_advanced_graphics-solution.html#subsetting).



## <a name="professional"></a> Making professional looking graphics

To make the graphics we are producing look professional, we will need to 
[customize](graphics.html#customizing) the graph according to the final 
production platform, e.g. website, word document, pdf document, etc. 
This will be highly specific to your use. 
Nonetheless, likely you will use the following `ggplot2` functions:

- `scale_color_manual()`
- `labs()`
- `theme()`

But there are many others. 
One thing you will notice is that we can keep updating the graph by updating 
the R object that holds the graph. 

### Base graphic 

Let's start with the following graph

```{r}
GI$weekD = as.Date(GI$weekC) 

GI_sum <- GI %>%
  group_by(weekD, gender, ageC) %>%
  summarize(count = n())

ggplot(GI_sum, aes(x = weekD, y = count, color = gender)) + 
  geom_point(size = 3) + 
  facet_grid(ageC ~ ., scales = "free_y")
```

### Change age category labels

The easiest way to change some of the labels on the graph is to change the 
underlying data. Here we change the levels associated with the `ageC` variable. 

```{r}
levels(GI_sum$ageC)
levels(GI_sum$ageC) = paste("Age:", c("<5","5-18","18-45","45-60",">60"))
table(GI_sum$ageC)
```

```{r}
g = ggplot(GI_sum, aes(x = weekD, y = count, color = gender)) + 
  geom_point(size = 3) + 
  facet_grid(ageC ~ ., scales = "free_y")
g
```

`g` is the R object that contains the graph


### Change colors 

Intuitive colors can make a graph easier to read. 


```{r}
g = g + scale_color_manual(values=c("hotpink","blue"), name='Gender')
g
```


### Change title/axis labels and put legend on the bottom

Use proper names and capitalization in title and axis labels.

```{r}
g = g + 
  labs(title = "Weekly GI cases", x = "Year", y = "Weekly count") + 
  theme(legend.position = "bottom")
g
```

### Try alternate themes

Try alternate themes:

```{r}
g = g + theme_bw()
g
```

To see a list of possibilities:

```{r, eval=FALSE}
?theme_bw
?theme
```


### Adjust font sizes

Depending on the final platform, font sizes may need to be adjusted.

```{r}
g = g + theme(title = element_text(size=rel(2)),
              text = element_text(size=16),
              legend.background = element_rect(fill  = "white", 
                                               size  = .5, 
                                               color = "black"))
g
```


### <a name="exporting"></a> Exporting graphs

To export the graphs, use the `ggsave()` function which saves the last plot that 
you displayed. 
As a default, it uses the size of the current graphics device, 
but you will likely want to modify this. 

```{r}
g
ggsave("plot.pdf")
```

If you look in your current working directory, you will see the plot.pdf file. 

```{r}
ggsave("plot.pdf", width=14, height=8)
```
