---
title: "Lab04 - Graphical Statistics"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

To follow along, use the [lab04 code](lab04.R) also run the following

```{r, eval=FALSE}
install.packages("tidyverse")
```

If the install fails, then run

```{r, eval=FALSE}
install.packages(c("dplyr","ggplot2"))
```

The installation only needs to be done once. 
But we will need to load the packages in every R session where we want to use 
them.
To load the packages, use 

```{r, eval=FALSE}
library("dplyr")
library("ggplot2")
```

alternatively, you can load the entire (not very big) `tidyverse`.

```{r}
library("tidyverse")
```


The main purpose of the lab today is to construct plots using the `ggplot2`
R package.
In order to construct these plots, 
we need to construct an appropriate `data.frame` and we will use 
`dplyr` to help us construct that `data.frame`. 


## Data wrangling

Let's use the built-in R data set `airquality`.
First, let's look at the top of the file using the `head()` function.

```{r}
head(airquality)
```

We can get more information, by going to the help file

```{r, eval=FALSE}
?airquality
```

One issue with this dataset is that the Month/Day columns don't really provide 
us with a `Date`.
Let's create a new column that creates a real `Date`.

```{r}
airquality <- airquality %>%
  mutate(Date = as.Date(paste("1973",Month,Day,sep="/"))) 
```

Now we can get rid of the old Month and Day columns

```{r}
airquality <- airquality %>%
  select(-Month, -Day)
```

# Scatterplot

At this point we can construct individual graphs for our 4 different response
variables: Ozone, Solar.R, Wind, and Temp.

```{r}
ggplot(airquality, aes(x = Date, y = Ozone)) +
  geom_point()
```

or if we wanted a line plot

```{r}
ggplot(airquality, aes(x = Date, y = Ozone)) +
  geom_line()
```

#### Scatterplot activity

Create a scatterplot of solar radiation vs date.

<div id="scatterplot_activity_solution" style="display:none"> 
```{r, purl=FALSE}
ggplot(airquality, aes(x = Date, y = Solar.R)) +
  geom_point()
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('scatterplot_activity_solution') .style.display=='none') {document.getElementById('scatterplot_activity_solution') .style.display=''}else{document.getElementById('scatterplot_activity_solution') .style.display='none'}">Show/Hide Solution</button>


### Faceted scatterplots

What if we want to visualize all 4 response variables vs Date simultaneously.
One way to do this is with facets, but in order to use facets, we will need
to organize our `data.frame` into long format. 

```{r}
d <- airquality %>%
  tidyr::gather(response, value, -Date)
```

Take a look at the resulting `data.frame`.

```{r}
head(d)
summary(d)
table(d$response)
```




```{r, echo=FALSE, purl=FALSE, results="hide", message=FALSE}
knitr::knit("lab04.Rmd", tangle = TRUE)
```
