---
title: "Lab10 - Contrasts"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


## Preparation

To follow along, use the [lab10 code](lab10.R) and make sure the following 
packag is installed:

- lsmeans

You can use the following code to perform the installation:

```{r install_packages, eval=FALSE}
install.packages("lsmeans")
```

Now load the packages

```{r load_packages}
library("dplyr")
library("ggplot2")
# library("Sleuth3")
library("lsmeans")
```

## One explanatory variables

Consider the `fiber` data set in the lsmeans package where
the stregnth of fibers produced by 3 different machines is tested. 

```{r}
ggplot(lsmeans::fiber, aes(machine, strength)) + 
  geom_point() + 
  theme_bw()
```

### Means

We might be interested in the mean strength of fibers produced by each machine.
One approach would be to fit a regression model and predict for a new 
observation for each machine type.

```{r}
m <- lm(strength ~ machine, data = lsmeans::fiber)
nd <- data.frame(machine = c("A","B","C"))
p <- predict(m, 
             newdata = nd, 
             interval = "confidence")
bind_cols(nd, p %>% as.data.frame)
```
Alternatively, use the `lsmeans` function in the 
[lsmeans](https://cran.r-project.org/web/packages/lsmeans/index.html) 
package.

```{r}
lsmeans(m, ~machine)
```

#### Means activity

Consider the `ex0518` data set in the Sleuth3 package.

```{r}
ex0518 <- Sleuth3::ex0518 %>%
  mutate(Treatment = relevel(Treatment, ref="Control"))

ggplot(ex0518, aes(Treatment, Protein)) + 
  geom_point() + 
  theme_bw()
```

Fit a regession model of protein on treatment and 
compute a point estimate as well as a 95% CI for the mean protein level in 
each of the treatment (diet) levels.

<div id="means_activity_solution" style="display:none"> 
```{r, purl=FALSE}
m <- lm(Protein ~ Treatment, data = ex0518)
lsmeans(m, ~Treatment)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('means_activity_solution') .style.display=='none') {document.getElementById('means_activity_solution') .style.display=''}else{document.getElementById('means_activity_solution') .style.display='none'}">Show/Hide Solution</button>

### Comparison of means

Typically we are more interested in saying something about differences in means.
We can either try to specify the contrasts of interest, or we can use 
pre-packaged analyses to extract those contrasts we are interested in.
Suppose that machine C was really the control and we are interested primarily
in comparing 

- machine A to machine C and
- machine A to machine B.

We can use the `contrast`  function in the `lsmeans` package to perform the
comparison.

```{r}
m <- lm(strength ~ machine, data = lsmeans::fiber)
ls <- lsmeans(m, ~ machine)
(co <- contrast(ls, method = "pairwise"))
```
To get the CIs, use `confint`:

```{r}
confint(co)
```

By default, these functions a Tukey multiple comparison adjustment which is the 
most appropriate adjustment for performing all pairwise comparisons. 

You can turn off the adjustment by setting the `adjust` argument to "none", e.g.

```{r}
(co <- contrast(ls, method = "pairwise", adjust="none"))
confint(co)
```

Also, we weren't really interested in looking at all pairwise comparisons.
Instead, we were really mainly interested in looking at machine A/B versus 
machine C (the control).

We can specify just these comparisons by changing the `method` argument to 
"trt.vs.ctrlk" (which stands for treatment vs control).

```{r}
(co <- contrast(ls, method = "trt.vs.ctrlk"))
confint(co)
```
Notice that the adjustment method used here is the `Dunnett` method which is
the appropriate multiple comparison adjustment method for comparing all 
treatment levels to a control. 

#### Contrast activity

Reconsider the `ex0518` data set and provide an estimate and 95% CI for the 
difference in mean protein level for all treatments compared to control.

<div id="contrasts_activity_solution" style="display:none"> 
```{r, purl=FALSE}
m <- lm(Protein ~ Treatment, data = ex0518)
ls <- lsmeans(m, ~ Treatment)
(co <- contrast(ls, method = "trt.vs.ctrlk"))
confint(co)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('contrasts_activity_solution') .style.display=='none') {document.getElementById('contrasts_activity_solution') .style.display=''}else{document.getElementById('contrasts_activity_solution') .style.display='none'}">Show/Hide Solution</button>




## Two categorical explanatory variables

These data are taken from 
[a wood glue experiment](https://woodgears.ca/joint_strength/glue.html).
For simplicity, the data here are taken to be balanced.

```{r}
wood_glue <- data.frame(force = c(185,170,210,240,245,190,210,250,
                                  290,280,260,270,200,280,350,350),
                        wood = rep(c("spruce","maple"),each = 8),
                        glue = rep(c("carpenter's", "weldbond","gorilla","titebond"), each=2, times=2))
```

```{r}
ggplot(wood_glue, aes(wood, force, color=glue, shape=glue)) +
  geom_point() +
  theme_bw()
```

We can fit the regression model with wood, glue, and their interaction.

```{r}
m <- lm(force ~ wood*glue, data = wood_glue)
anova(m)
```

### Compare mean force averaged across wood type

```{r}
(ls <- lsmeans(m, ~ glue))
```

The note here indicates that there may be issues with comparing glue's when
averaged over wood since, by construction of the model, the glue's effect on 
the force may be different depending on the wood type.

If we want to compare the glue's, we can use `contrast` as before.

```{r}
(co <- contrast(ls, "pairwise"))
confint(co)
```

### Compare mean force within each wood type

Since the glue's effect may depend on wood type, 
we may be interested in comparing the glues within each wood type.
To estimate the mean force required to break the joint for every wood-glue 
combination use the `lsmeans` function with `A|B` denoting we want to calculate
the means for A by B. 

```{r}
(ls <- lsmeans(m, ~ glue | wood))
```

We would have gotten the exact same result if we switched `glue` and `wood`, 
but this provides a table in the more directly comparable order.

If we want to compare glue's within the wood type, we again use `contrast`. 

```{r}
(co <- contrast(ls, "trt.vs.ctrlk"))
confint(co)
```

#### Two categorical explanatory variables activity

Consider the `ex1321` data set in the Sleuth3 package. 
Estimate the average treatment effect on the intelligence test across all the 
classes.

<div id="twoC_activity_solution" style="display:none"> 

We should always look at the data first.

```{r, purl=FALSE}
ggplot(Sleuth3::ex1321, aes(Class, Gain, color=Treatment, shape = Treatment)) +
  geom_jitter() + 
  theme_bw()
```

Then we can estimate the average treatment effect across the classes.

```{r, purl=FALSE}
m <- lm(Gain ~ Treatment*Class, data = Sleuth3::ex1321)
(ls <- lsmeans(m, ~ Treatment))
(co <- contrast(ls, "pairwise"))
confint(co)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('twoC_activity_solution') .style.display=='none') {document.getElementById('twoC_activity_solution') .style.display=''}else{document.getElementById('twoC_activity_solution') .style.display='none'}">Show/Hide Solution</button>





## ANCOVA - Categorical explanatory variable with continuous explanatory variable

Reconsider the `fiber` data set in the lsmeans package where
the strength of fibers produced by 3 different machines is tested. 
The machines do not actually produce a uniform size fiber.
So in addition to recording the strength, the diameter of the fiber is also 
recorded.
Since we expect there to be differences in strength depending on the diameter,
we should compare machine after adjusting for diameter. 

```{r}
ggplot(lsmeans::fiber, aes(diameter, strength, color=machine, shape=machine)) + 
  geom_point() + 
  theme_bw()
```

### Comparing treatments at a particular level of the continuous explanatory variable

We will fit the model with an interaction between machine and diameter to allow
the possibility that the effect of diameter depends on the machine.

```{r}
m <- lm(strength ~ diameter*machine, data=fiber)
ls <- lsmeans(m, ~ machine)
co <- contrast(ls, method = "pairwise")
confint(co)
```

Again, 
we are warned that these results may not make sense due to the interaction.
Perhaps we should evaluate these at a particular diameter.

```{r}
ls <- lsmeans(m, ~ machine | diameter)
co <- contrast(ls, method = "pairwise")
confint(co)
```

We can see that, in fact, the analysis provided above was at a diameter of 
24.1333 which happens to be the mean diameter in the data set.

```{r}
mean(fiber$diameter)
```

### Comparing the effect of the continuous variable across machines

We can also look at the effect of the continuous variable across machines and 
see if this is the same, i.e. is the slope the same for all machines.
To do this we use the `lstrends` function.

```{r}
( lst <- lstrends(m, "machine", var = "diameter") )
(co <- contrast(lst, method = "pairwise"))
confint(co)
```



#### ANCOVA activity

Take a look at the bat energy expenditure data set in `case1002` and determine
the relationship between echo-locating ability and energy expenditure after
adjusting for mass. 
In particular, compare the differences in mean energy expenditure between 
echolocating bats and non-echolocating bats as well as echolocating bats and 
birds at the mean mass.
Does comparing at the mean mass make sense?
Also compare the effects of mass on mean energy expenditure between those same 
groups.

<div id="ancova_activity_solution" style="display:none"> 

As always, plot the data. 

```{r, purl=FALSE}
ggplot(Sleuth3::case1002, aes(Mass, Energy, color = Type, shape = Type)) +
  geom_point() + 
  theme_bw()
```

Compare energy expenditure at mean mass.

```{r, purl=FALSE}
m <- lm(Energy ~ Mass*Type, data = Sleuth3::case1002)
ls <- lsmeans(m, ~ Type | Mass)
co <- contrast(ls, "pairwise") # Ideally we would use "trt.vs.ctrlk" here, can anybody figure out how?
confint(co)
```

This doesn't really make sense since there are no echolocating bats whose mass 
is anywhere near this range.

To compare the slopes, use 

```{r, purl=FALSE}
( lst <- lstrends(m, "Type", var = "Mass") )
(co <- contrast(lst, method = "pairwise"))
confint(co)
```

</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('ancova_activity_solution') .style.display=='none') {document.getElementById('ancova_activity_solution') .style.display=''}else{document.getElementById('ancova_activity_solution') .style.display='none'}">Show/Hide Solution</button>




## Custom contrasts

Although pairwise comparisons are common for contrasts, 
we may be interested in more sophisticated contrasts. 
For example, in the mice lifetime diet, we may be interested in comparing 
the standard N/N85 diet to the low calorie diets.
In addition, we may be interested in comparing N/R50 vs R/R50 to look specifically
at pre-weaning calorie restriction.
To perform these analysis, we use custom contrasts (although the second one is 
technically part of the pairwise comarisons).

```{r}
m <- lm(Lifetime ~ Diet, data = Sleuth3::case0501)
ls <- lsmeans(m, "Diet")
#                                           N/N85 N/R40 N/R50  NP R/R50 lopro
co <- contrast(ls, list(`High - Low`    = c(    4,   -1,   -1,  0,   -1,   -1) /4,
                        `Pre-wean: R-N` = c(    0,    0,   -1,  0,    1,    0) ))
confint(co)
```
