---
title: "Lab03 - Probability Distributions in R"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


To follow along, use the [lab03 code](lab03.R).

## Probability distributions in R

Many of the standard probability distributions have functions in R to calculate:

 - probability density/mass function (d)
 - cumulative distribution function (p)
 - quantile function (q)
 - random draws (r)
 
The letter after each indicates the letter the function will start with for 
all distributions.
 
### Uniform distribution
 
The uniform distribution is one of the easiest distributions to understand. 
If Y is a random variable and it is uniformly distributed between a and b, 
then we write $Y \sim Unif(a,b)$. 
 
#### Probability density function
 
The probability density function for a uniform random variable is zero outside
of a and b (indicating that the random variable cannot take values below a
or above b) and is constant at a value of 1/(b-a) from a to b. 
 
```{r}
a <- 0
b <- 1
curve(dunif(x, min = a, max = b), from = -1, to = 2,
      xlab='y', ylab='f(y)', main='Probability density function for Unif(0,1)')
```
 
#### Cumulative distribution function
 
The cumulative distribution function is the integral from negative infinite 
up to y of the probability density function. 
Thus it indicates the probability the random variables will take on values less
than y, i.e. $P(Y<y)$. 
Since the probability density function for the uniform is constant, 
the integral is simply zero from negative infinite up to a, 
then a straight line from (a,0) up to (b,1) and then constant after that.

```{r}
curve(punif(x, min = a, max = b), from = -1, to = 2,
      xlab='y', ylab='f(y)', main='Cumulative distribution function for Unif(0,1)')
```
 
#### Quantile function

The quantile function is the inverse of the cumulative distribution function.
For a given probability p, it finds the value such that the 
probability the random variable is below that value is p.

```{r}
curve(qunif(x, min = a, max = b), from = 0, to = 1,
      xlab='y', ylab='f(y)', main='Quantile function for Unif(0,1)')
```

#### Random draw function

To draw random values from the distribution, use the `r` version of the 
function.
For instance, here is 100 random Unif(a,b) draws represented as a histogram.

```{r}
random_uniforms <- runif(100, min = a, max = b)
hist(random_uniforms, probability = TRUE)
```
