---
layout: post
title: "DLM MCMC"
description: ""
category: [Teaching]
tags: [STAT 615,Monte Carlo,MCMC,DLM,R]
---
{% include JB/setup %}

This post looks at the autocorrelation in a simple DLM when using JAGS, which samples each state individually, and FFBS, which samples the states jointly. 
This is put within the context of a local level model with unknown observation and evolution variance. 


Simulate the data

```{r}
require(methods, quietly=TRUE)
require(dlm, quietly=TRUE)
require(rjags, quietly=TRUE)

# Simulate data
ll = dlmModPoly(1, dV=1, dW=.01, m0=0, C0=1)
o = dlmForecast(ll, 100, sampleNew=1)
```

Fit the model using [JAGS](http://mcmc-jags.sourceforge.net/) (the model looks terrible, but it should work).

```{r} 
jags_ll = "
model {
  for (i in 2:n) {
    y[i-1] ~ dnorm(theta[i], tauV)
    theta[i] ~ dnorm(theta[i-1], tauW)
  }

  theta[1] ~ dnorm(0,1)
  tauV ~ dgamma(1,1)
  tauW ~ dgamma(1,1)
}  
"

dat = list(y=as.numeric(o$newObs[[1]]))
dat$n = length(dat$y)+1
m = jags.model(textConnection(jags_ll), dat, 
               n.adapt=2000)
res = coda.samples(m, c("tauV", "tauW","theta"), n.iter=2000)
```

Fit the model using the [dlm](http://cran.r-project.org/web/packages/dlm/index.html) package in R. 

```{r}
joint = dlmGibbsDIG(o$newObs[[1]], ll, 
                    a.y=1, b.y=1, 
                    a.theta=1, b.theta=1,
                    n.sample=1000, thin=1)
```

Plot traceplots and autocorrelation for the middle state. 

```{r}
par(mfrow=c(2,2))
plot(as.numeric(res[[1]][1001:2000,52]), type="l")
acf(res[[1]][,52])

plot(joint$theta[52,1,],type="l")
acf(joint$theta[52,1,])
```

The elephant in the room here is in the implementation since the JAGS code runs quite a bit faster than the dlm code. 
Of course, the dlm code is written to handle much more complicated DLMs. 
