---
layout: post
title: "ggplot2 ISU palette"
description: ""
category: 
tags: [ggplot2]
---
{% include JB/setup %}

To make figures in R using ggplot2 that use Iowa State University colors, 
you can 
[follow these instructios](https://stackoverflow.com/questions/17180115/manually-setting-group-colors-for-ggplot2)
and construct the palette using 
[ISU colors](http://www.brandmarketing.iastate.edu/brand-elements/color-palette/).

First we need to construct the palette

```{r}
library(ggplot2)

ISU_primary_palette   <- c("#C8102E", "#F1BE48", "#524727", 
                           "#9B945F", "#CAC7A7")

ISU_secondary_palette <- c("#3E4827", "#76881D", "#A2A569",
                           "#003D4C", "#006BA6", "#7A99AC",
                           "#7C2529", "#9A3324", "#BE531C",
                           "#8B5B29", "#B9975B", "#EED484",
                           "#6E6259", "#707372", "#ACA39A")
```

Then, we need to construct the plot.
For demonstration purposes, I use a bar chart, 
but generally I would not be using a bar chart.
Here is a bar chart using the primary palette. 

```{r primary_palette}
n_primary <- length(ISU_primary_palette)

# Name the palette
groups <- paste("group", 1:n_primary, sep="")
names(ISU_primary_palette) <- groups

d <- data.frame(group = groups, value = runif(n_primary))

ggplot(d, aes(x = group, y = value, fill = group)) + 
  geom_bar(stat="identity") + 
  scale_fill_manual(values = ISU_primary_palette) + 
  theme_bw()
```


Here is a bar chart using the secondary palette. 

```{r secondary_palette}
n_secondary <- length(ISU_secondary_palette)

# Name the palette
groups <- paste("group", 1:n_secondary, sep="")
names(ISU_secondary_palette) <- groups

d <- data.frame(group = factor(groups, levels = groups), 
                value = runif(n_secondary))

ggplot(d, aes(x = group, y = value, fill = group)) + 
  geom_bar(stat="identity") + 
  scale_fill_manual(values = ISU_secondary_palette) + 
  theme_bw()
```

This post makes me think that I should just create an R package that has the 
color palette.
