---
layout: page
title: STAT 486/586
tagline: "Distributions"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](12-distributions.R)

```{r}
library("tidyverse"); theme_set(theme_bw())
library("scales") # for breaks_pretty()
```

# R conventions

## Distribution naming

The distributions built into base R can be found using 

```{r, eval=FALSE}
?Distributions
```

Most distributions have 4 functions:

- d<name>: probability mass/density function
- p<name>: cumulative distribution function
- q<name>: quantile function (inverse of the cdf)
- r<name>: random draws



## Reproducible random draws

Simulations will necessarily be random and thus, every time you obtain a random
draw, it will be different. 

For example,
```{r}
runif(5)
runif(5)
```

Many times we will want to create a reproducible set of simulations. 
To do this, you can use the `set.seed()` function. 

For example,
```{r}
set.seed(1)
runif(5)
set.seed(1)
runif(5)
```




# Discrete distributions

Let's start with discrete distributions

## Binomial

### Probability mass function

Recall the pmf for a binomial is 
\[ 
P(Y=y) = {n\choose y} p^y (1-p)^{n-y}
\]
for integer $n>0$, $0<p<1$, and $y=0,1,\ldots,n$.

We can plot this pmf using `dbinom`.

```{r}
n <- 10
p <- 0.3
d <- data.frame(
  y = 0:n
) %>%
  mutate(pmf = dbinom(y, size = n, prob = p))

ggplot(d, aes(x = y, y = pmf, yend = 0, xend = y)) +
  geom_point() + 
  geom_segment()
```

### Cumulative distribution function

We can plot this cdf using `pbinom`.

```{r}
d <- d %>%
  mutate(cdf = pbinom(y, size = n, prob = p))

ggplot(d, aes(x = y, y = cdf, yend = 0, xend = y)) +
  geom_step()
```

More often, we use the cdf to calculate probabilities. 
For example, suppose $Y \sim Bin(30, 0.4)$ and we want to calculate 
$P(10<Y<14)$.
Recall that 
\[
P(10<Y<14) = P(Y\le 13) - P(Y \le 10)
\]
we can do this using

```{r}
n <- 30
p <- 0.4
pbinom(13, size = n, prob = p) - 
  pbinom(1, size = n, prob = p)
```



### Random draws

To obtain random draws from a binomial use the `rbinom` function.

```{r}
draws <- rbinom(1000, size = n, prob = p)

ggplot(mapping = aes(x = draws)) +
  geom_histogram(binwidth = 1)
```

Let's compare our empirical probabilities to the true probabilities. 

```{r}
mean(10 < draws & draws < 14) # vs 
pbinom(13, size = n, prob = p) - 
  pbinom(10, size = n, prob = p)
```


## Poisson

### Probability mass function

A random variable has a Poisson distribution it its probability mass 
function is 
\[
P(Y=y) = \frac{e^{-r}r^y}{y!}
\]
for some rate $r>0$ and integer $y\ge 0$. 

We use `dpois` for the Poisson pmf. 

```{r}
r <- 3
d <- data.frame(y = 0:(3*r)) %>%
  mutate(pmf = dpois(y, lambda = r),
         cdf = ppois(y, lambda = r))

ggplot(d, aes(x = y, y = pmf, xend = y, yend = 0)) + 
  geom_point() + 
  geom_segment() +
  scale_x_continuous(breaks = breaks_pretty())
```

### Cumulative distribution function

We use `ppois` for the Poisson pmf. 

```{r}
ggplot(d, aes(x = y, y = cdf)) + 
  geom_step() +
  scale_x_continuous(breaks = breaks_pretty())
```

### Random draws

To obtain random draws from a Poisson distribution use `rpois`.

```{r}
draws <- rpois(1000, lambda = r)

ggplot(mapping = aes(x = draws)) + 
  geom_histogram(binwidth = 1, color = "white") + 
  scale_x_continuous(breaks = breaks_pretty())
```




## Multinomial distribution

Often times we have a distribution over a categorical variable and each category
has a certain probability. 
This is called a multinomial distribution and is a generalization of the 
binomial distribution to more than two categories. 

### Probability mass function

Let $X_i$ for $i = 1,\ldots, k$ be the number of observations in category $i$.
Then a random vector $(X_1,\ldots,X_k)$ has a multinomial distribution if the 
probability mass function is 
\[ 
f(x_1,\ldots,x_k) = 
\frac{n!}{x_1!\ldots x_k!}p_1^{x_1}\times \cdots \times p_k^{x_k}
\]
with $\sum_{i=1}^k x_i = n$ and probabilities $p_k>0$ and $\sum_{i=1}^k p_k = 1$. 
### Random draws

Unlike all the other distributions with their consistent naming convention,
we will use the `sample()` function to draw from a multinomial distribution. 
This will result in telling us the categories rather than the counts,
but we can always compute the counts later. 

```{r}
sample(c("Cheerios","Trix","Cocoa puffs"), 
       size = 10, replace = TRUE, prob = c(.4,.5,.1))
```


# Continuous distributions

## Uniform

The uniform distribution, particularly on (0,1), is the result of many 
pseudorandom number generators and is also the basis of many algorithms to
obtain other random variables. 

### 


## Normal (Gaussian)

### Probability density function

```{r}
m <- 0
s <- 1 # standard deviation

d <- data.frame(x = seq(m - 3*s, m + 3*s, length = 1001)) %>%
  mutate(pdf = dnorm(x, m, s),
         cdf = pnorm(x, m, s))

ggplot(d, aes(x = x, y = pdf)) +
  geom_line()
```


### Cumulative distribution function

```{r}
ggplot(d, aes(x = x, y = cdf)) +
  geom_line()
```

### Critical values

The `qnorm` function provides the inverse of the normal CDF. 
We can plot this function

```{r}
ggplot(data.frame(p = 0:1)) +
  stat_function(fun = qnorm, n = 1001, args = list(mean = m, sd = s))
```

The most common use of this function is to obtain z critical values for us 
in constructing confidence intervals. 

```{r}
qnorm(.975)
```

### Random draws

We can use the `rnorm` function to sample random draws from the normal 
distribution. 

```{r}
draws <- rnorm(20)

ggplot() +
  geom_histogram(mapping = aes(x = draws), bins = 10)
```

```{r}
ggplot() +
  geom_histogram(
    mapping = aes(
      x = draws, 
      y = after_stat(density)), 
    bins = 10) +
  stat_function(fun = dnorm, 
                n = 1001, 
                args = list(
                  mean = mean(draws), 
                  sd = sd(draws)),
                color = 'red') +
  xlim(min(draws)-1, max(draws) + 1)
```

If we have a much larger sample size, 
the histogram will look approximately normal.

```{r}
draws <- rnorm(1e4)

ggplot() +
  geom_histogram(
    mapping = aes(
      x = draws, 
      y = after_stat(density)), 
    bins = 30) +
  stat_function(fun = dnorm, 
                n = 1001, 
                args = list(
                  mean = mean(draws), 
                  sd = sd(draws)),
                color = 'red') +
  xlim(min(draws)-1, max(draws) + 1)
```


## Student's t

### Probability density function

```{r}
df <- 10

d <- data.frame(x = seq(-4, 4, length = 1001)) %>%
  mutate(pdf = dt(x, df = df),
         cdf = pt(x, df = df))

ggplot(d, aes(x = x, y = pdf)) + 
  geom_line() +
  stat_function(fun = dnorm, color = 'red')
```

### Calculative p-values

Suppose you have the following data

```{r}
y <- round(rnorm(10), 2)
y
```

Which has the following summary statistics

```{r}
n <- length(y); n 
m <- mean(y); m
s <- sd(y); s
```

If you are interested in the following hypothesis test:
\[ 
H_0: \mu=1 \quad \mbox{vs} \quad H_A: \mu \ne 1
\]
then the test statistic is 

```{r}
t <- (m-1)/(s/sqrt(n)); t
```

Under the null hypothesis, this statistic has a $t$ distribution with $n-1$ 
degrees of freedom. 
We can then compute the p-value using 

```{r}
pvalue <- 2*pt(-abs(t), df = n-1); pvalue
```

Of course, we could use the built-in functions in R

```{r}
t.test(y, mu = 1)$p.value
```

### Critical values

The main use of the quantile function (or inverse CDF) is to compute 
t critical values. 
For example, 
```{r}
tcrit <- qt(.95, df = n-1); tcrit
```

So a 90\% CI for the mean is 

```{r}
m + c(-1,1)*tcrit*s/sqrt(n)
```

Of course, we could just use built-in functions

```{r}
t.test(y, conf.level = 0.9)$conf.int
```



