---
layout: page
title: "Normal analysis"
output: 
  html_document:
    toc: true
---

This page provides a summary of statistical methods in R used to analyze 
normal data. 
It begins by considering a single group and then expands to multiple groups. 
For completing course activities, you should copy relevant code from here and
modifying it as necessary. 

In the code below, we will make use of the `tidyverse` package, so
make sure to load it at the beginning of your R session.

```{r}
library("tidyverse")
theme_set(theme_bw()) # set a figure theme
```

## Single group

Normal data, written $Y_i \stackrel{ind}{\sim} N(\mu,\sigma^2)$ for 
individuals $i=1,\ldots,n$ is characterized by a group of continuous observations
with 

- common mean $\mu$,
- common standard deviation $\sigma$,
- normal distribution (bell-shaped histogram), and
- each observation is independent. 

If our sample is based on a random sample from our population, 
then we can make inferences about that entire population. 

Suppose, we obtain 33 observations with a sample mean of 13 and a sample
standard deviation of 2. 
In R, we can write 

```{r}
n  <- 33
mn <- 13
s  <- 2
se <- s/sqrt(n) # standard error (used often below)
```

Our posterior belief about the unknown population mean $\mu$ 
is represented by the following curve

```{r}
# Probability density function for a location-scale t-distribution
dlst <- function(x, df, location, scale) {
  dt((x-location)/scale, df = df)/scale
}

d <- data.frame(mu = seq(from = mn - 4*se, to = mn + 4*se, length = 1001)) %>%
  mutate(density = dlst(mu, df = n-1, location = mn, scale = se))

ggplot(d, aes(x = mu, y = density)) + 
  geom_line() + 
  labs(x = expression(mu),
       y = "Posterior belief",
       title = "Posterior belief about population mean")
```

We can calculate probabilities based on this curve by finding the area under
the curve. 
For example, to find the probability that $\mu$ is between 
12 and 13.5 we use 

```{r}
pt( (13.5-mn)/se, df = n-1) - pt( (12-mn)/se, df = n-1)
```

which corresponds to the area

```{r}
ggplot(d, aes(x = mu, y = density)) + 
  stat_function(fun = dlst, xlim = c(12,13.5), geom="area", fill = "red",
                args = list(df = n-1, location = mn, scale = se)) + 
  geom_line() + 
  labs(x = expression(mu),
       y = "Posterior belief",
       title = "Belief that the mean is between 12 and 13.5")
```

A 95% credible interval is calculated using

```{r}
a <- 1-0.95
qt(c(a/2, 1-a/2), df = n-1) * se + mn
```




## Multiple groups

Normal data for multiple groups is written 
$Y_{i,g} \stackrel{ind}{\sim} N(\mu_g, \sigma_g^2)$ 
for
$g=1,\ldots,G$ where $g$ is the number of groups and 
$i = 1,\ldots,n_g$ where $n_g$ is the number of individuals in group $g$. 
For group $g$, we have 

- common mean $\mu_g$,
- common standard deviation $\sigma_g$,
- normal distribution (bell-shaped histogram), and
- each observation is independent. 

And, we have independence of observations across the two groups. 

If our sample is based on a random sample from our population, 
then we can make inferences about that entire population. 

```{r, echo=FALSE,purl=FALSE}
set.seed(20220217)

d <- data.frame(group = rep(c("group1","group2"), times = c(15,20))) %>%
  mutate(y = rnorm(n()) * 2^(group == "group2") + 
           66 + 1*(group == "group1"),
         y = round(y))

write_csv(d, file = "normal-data.csv")
```

Rather than having summary data, 
we will instead record our data in a file and read it in to R.

```{r}
d <- read_csv("normal-data.csv")

head(d)
```

We can calculate summary statistics from these data

```{r}
sm <- d %>% 
  group_by(group) %>%
  summarize(n = n(),
            mn = mean(y),
            s = sd(y)) %>%
  mutate(se = s/sqrt(n))

sm
```

We can plot our posterior beliefs using the following code

```{r}
# Function to create the posterior
create_posterior <- function(sm) {
  
  data.frame(mu = seq(sm$mn-4*sm$se, sm$mn+4*sm$se, length=1001)) %>%
    mutate(posterior = dlst(mu, df = sm$n-1, location = sm$mn, scale = sm$se))
}

# Construct the curves
curves <- sm %>%
  group_by(group) %>%
  do(create_posterior(.)) %>%
  mutate(group = factor(group)) # so that we can use it as a linetype

# Plot curves
ggplot(curves, aes(x = mu, y = posterior, color = group, linetype = group)) +
  geom_line() + 
  labs(x = expression(mu),
       y = "Posterior belief",
       title = "Posterior beliefs for the mean in two groups.") 
```

If we want to calculate beliefs that one groups mean is 
larger than the other, we use a Monte Carlo approach.
This approach first simulates values for $\mu_1$ and $\mu_2$. 

```{r}
n_reps <- 100000
mu1 <- rt(n_reps, df = sm$n[1]-1) * sm$se[1] + sm$mn[1]
mu2 <- rt(n_reps, df = sm$n[2]-1) * sm$se[2] + sm$mn[2]
```

Then we calculate the proportion of times that $\mu_1$ is larger than $\mu_2$
which is calculated in R using the `mean` function.

```{r}
mean(mu1 > mu2)
```

We can also use these samples to obtain a 95% credible interval for the 
difference between the two probabilities of success. 

```{r}
a <- 1-0.95
quantile(mu1 - mu2, probs = c(a/2, 1-a/2))
```


For additional groups a similar process can be followed. 
