---
title: "Data Management in R"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


## Getting started

Install the `dplyr` package using 

```{r, eval=FALSE}
install.packages("dplyr")
```

Change your working directory (see below). 

Save these to your working directory:

 - [lab02.R](lab02.R)
 - [experiments.csv](experiments.csv)
 - [treatment_codes.csv](treatment_codes.csv)


## <a name="reading">Reading Data into R</a>

In this section, we will learn how to read in csv or Excel files into R. 
We focus on csv files because they are simplest to import, 
they can be easily exported from Excel (or other software), 
and they are portable, 
i.e. they can be used in other software.

### <a name="working_directory">Changing your working directory</a>

One of the first tasks after starting R is to change the working directory. 
To set,

- in RStudio: Session > Set Working Directory > Choose Directory... (Ctrl + Shift + H)
- in R GUI (Windows): File > Change Dir... 
- in R GUI (Mac): Misc > Change Working Directory...

Or, you can just run the following command
```{r, eval=FALSE}
setwd(choose.dir(getwd()))
```

Make sure you have write access to this directory. 



### Reading data into R

Data are stored in many different formats. 
I will focus on data stored in a csv file, 
but mention approaches to reading in
data stored in Excel, SAS, Stata, SPSS, and database formats.

#### Reading a csv file into R

The most common way I read data into R is through a csv file. 
csv stands for comma-separated value file and is a standard file format for
data. 
The utils package (which is installed and loaded with base R) has a 
function called `read.csv` for reading csv files into R. 
For example, 

```{r}
experiments = read.csv("experiments.csv")
```

In order for this code to execute correctly, 
you need to have the [experiments.csv](experiments.csv) in your working 
directory.

This created a `data.frame` object in R called `experiments`. 

The utils package has the `read.table()` function which is a more general 
function for reading data into R and it has many options. 
We could have gotten the same results if we had used the following code:

```{r, tidy=FALSE}
experiments2 = read.table("experiments.csv", 
                 header=TRUE, # There is a header.
                 sep=",")     # The column delimiter is a comma.
```

To check if the two data sets are equal, use the following

```{r}
all.equal(experiments, experiments2)
```


The `read.csv` function is available in base R, but these days I will often
use the `read_csv` function in the 
[readr](https://cran.r-project.org/web/packages/readr/index.html).
See below for how to install packages using the `install.packages` function.

```{r, eval=FALSE}
install.packages("readr") # run this command if the readr package is not installed
library('readr')
experiments <- read_csv("experiments.csv")
```

The two main differences are 

- this function does NOT make R friendly column names (and therefore the column names are "pretty")
- this function does NOT turn strings into factors


#### Activity

Read the [treatment_codes.csv](treatment_codes.csv) into R 
and save it into an R object named `treatment_codes`.
If you have extra time, try to figure out how to use the `read_csv` function
from the `readr` package to read the file in.

<details><summary>Click for solution</summary>  
```{r, purl=FALSE}
treatment_codes <- read.csv("treatment_codes.csv")
```

```{r, eval=FALSE, purl=FALSE}
install.packages("readr") # only need to do this once
library("readr")          # need to do this every R session
treatment_codes <- read_csv("treatment_codes.csv")
```
</details>



### <a name="excel">Reading Excel files</a>

My main suggestion for reading Excel files into R is to 

1. Save the Excel file as a csv
1. Read the csv file into R using `read.csv`

This approach will work regardless of any changes Excel makes in its document 
structure.

Reading an Excel file into R is done using the `read_excel` function from 
the [readxl](https://cran.r-project.org/web/packages/readxl/index.html) R package. 
Unfortunately many scenarios can cause this process to not work. 
When it works, it looks like this

```{r, eval=FALSE}
install.packages('readxl')
library('xl')
d = read_excel("filename.xlsx", sheet = 1) # or
d = read_excel("filename.xlsx", sheet ="sheetName")
```

Again, if these approaches don't work, you can `Save as...` a csv file in Excel.


### Read SAS, Stata, or SPSS data files

The `haven` package provides functionality to read in SAS, Stata, and SPSS 
files. 
An example of reading a SAS file is

```{r, eval=FALSE}
install.packages('haven')
library('haven')
d = read_sas('filename.sas7bdat')
```

#### Read a database file

There are many different types of databases, so the code you will need will be
specific to the type of database you are trying to access. 

The [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html) package,  
which we will discussing today, has a [number of functions to read from some 
databases](https://cran.r-project.org/web/packages/dplyr/vignettes/databases.html).
The code will look something like
```{r, eval=FALSE}
library('dplyr')
my_db <- src_sqlite("my_db.sqlite3", create = T)
```

The [RODBC](https://cran.r-project.org/web/packages/RODBC/index.html) package 
has a 
[number of functions to read from some databases](http://www.statmethods.net/input/dbinterface.html).
The code might look something like

```{r, eval=FALSE}
install.packages("RODBC")
library('RODBC')

# RODBC Example
# import 2 tables (Crime and Punishment) from a DBMS
# into R data frames (and call them crimedat and pundat)

library(RODBC)
myconn <-odbcConnect("mydsn", uid="Rob", pwd="aardvark")
crimedat <- sqlFetch(myconn, "Crime")
pundat <- sqlQuery(myconn, "select * from Punishment")
close(myconn)
```



## Manipulating `data.frame`s in R

Some basic `data.frame` information is available using base R functions. 
But manipulating the `data.frame` becomes a lot easier using the 
[dplyr](https://cran.r-project.org/web/packages/dplyr/index.html).

You can check to see if a package is installed, by running

```{r}
library("dplyr", quietly = TRUE) # if you don't use 
```

If you get an error, you will likely need to install the package and
then re-run the previous command.

```{r, eval=FALSE}
install.packages("dplyr")
```


### Basic `data.frame` information

The dimensions of the `data.frame` can be found using `dim`, `nrow`, and `ncol`.

```{r}
dim(experiments)
nrow(experiments)
ncol(experiments)
```

The contents of the `data.frame` can be assessed using the following functions:

```{r}
names(experiments)
head(experiments)
tail(experiments)
str(experiments)
```

To start running some statistics on the `data.frame` a quick function is the 
`summary` function:

```{r}
summary(experiments)
```

Notice (under `win` column) that R will tell you if any observations are 
`Not Available` using `NA`.


### Subsetting a `data.frame`

First, let's make sure the `dplyr` package is loaded 
and `experiments` is loaded.

```{r}
library("dplyr")
experiments     <- read.csv("experiments.csv")
```

There are numerous ways in `dplyr` to subset observations:

- `filter()`
- `distinct()`
- `sample_frac()`
- `sample_n()`
- `slice()`
- `top_n()`

By far the most common one for me is to `filter()`.
This function will return all rows that match a set of logical criteria. 
For example, perhaps we want to extract all observations with treatment code
`AI`.

```{r}
experiments %>%
  filter(code == "AI") # notice the double equal
```

You can also filter by multiple criteria. 

```{r}
experiments %>%
  filter(win, n_cards_played > 30) %>% # win is already logical
  summary
```



#### Activity

Use the `treatment_codes` `data.frame` to keep only observations that used the
`random` strategy and 6 owls. 
If you get done with this quickly, keep only observations that used 2 or more
owls but less than 6. 

<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
# Read in data if you haven't done it already
treatment_codes <- read.csv("treatment_codes.csv")

treatment_codes %>%
  filter(strategy == "random", n_owls == 6) %>%
  summary

treatment_codes %>%
  filter(n_owls >= 2, n_owls < 6) %>%
  summary
```
</details>

### Subset the columns

If we want to subset the columns, we use the `select()` function. 
If used by itself, then it will keep any named columns and remove any names 
pre-fixed with a "-". 

```{r}
experiments %>% 
  select(n_cards_played, n_suns_played, n_owls_left) %>%
  names

experiments %>%
  select(-code, -rep, -win, -owl_score) %>%
  names
```

Since there are still many rows here, I only print out the column names.

Sometimes, it would be more efficient to perform some *fuzzy* matching using 
one of these functions:

- `contains()`
- `ends_with()`
- `everything()`
- `matches()`
- `num_range()`
- `one_of()`
- `starts_with()`

For example, keep only columns that start with "n_"

```{r}
experiments %>% 
  select(starts_with("n_")) %>%
  names
```



#### Activity

Use the `treatment_codes` `data.frame` to extract all the codes (and only codes)
for treatments that had two players and three owls. 

<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
# Read in the data if you haven't already
treatment_codes <- read.csv("treatment_codes.csv")

treatment_codes %>%
  filter(n_players == 2, n_owls == 3) %>%
  select(code)
```
</details>


### Making new variables

Proper data management requires that raw data remain raw and that scripts are
written to create analyzable data sets. 
These scripts will often need to construct new variables for future analyses.
The `dplyr` mutate function can be used to add new columns to the `data.frame`.

```{r}
experiments %>% 
  mutate(suns_spots_remaining   = 13 - n_suns_played,             # there are 13 sun spots
         average_owl_position   = owl_score / n_owls_left) %>% 
  head
```

Notice the `NaN` when we divided by zero. 
If there are no owls left, then it makes more sense that the average owl 
position is zero (which indicates the nest). 

```{r}
experiments %>% 
  mutate(suns_spots_remaining   = 13 - n_suns_played,
         average_owl_position   = owl_score / n_owls_left,
         average_owl_position   = ifelse(is.nan(average_owl_position), 
                                         0,                            # return 0 if NaN
                                         average_owl_position)) %>%    # otherwise leave alone
  head
```


#### Activity

For each observation, calculate the number of non sun cards played.

<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
experiments %>% 
  mutate(n_non_sun_cards_played = n_cards_played - n_suns_played) %>%    
  head
```
</details>



### Summarizing data

We will often want to summarize large `data.frame`s. 
The `dplyr` `summarize` function will allow us to do this summarization. 

```{r}
experiments %>%
  summarize(proportion_wins     = mean(win),
            mean_cards_played   = mean(  n_cards_played),
            median_cards_played = median(n_cards_played),
            min_cards_played    = min(   n_cards_played),
            max_cards_played    = max(   n_cards_played),
            sd_cards_played     = sd(    n_cards_played),
            var_cards_played    = var(   n_cards_played))
```

Often, we will want to perform these summarizations on groups. 
In order to do this we will first

1. Group the `data.frame` and then
1. Perform the summarization within each group

```{r}
experiments %>%
  group_by(win) %>%
  summarize(mean_cards_played   = mean(  n_cards_played),
            median_cards_played = median(n_cards_played)) %>%
  ungroup()  # this isn't necessary here, but there may be times when you want to ungroup
```

#### Activity

Calculate the average number of cards played after grouping by the 
number of owls left.

<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
experiments %>%
  group_by(n_owls_left) %>%
  summarize(mean_cards_played   = mean(  n_cards_played),
            median_cards_played = median(n_cards_played))
```
</details>



### Combining `data.frame`s

The `experiment` `data.frame` is set up with *blinding*. 
In order to *unblind* these data, we will combine it with the `treatment_codes`
`data.frame`. 

These two `data.frame`s have the `code` column in common:

```{r}
# Read in the data if you haven't already
treatment_codes <- read.csv("treatment_codes.csv")

names(experiments)
names(treatment_codes)
```

So, now let's create a new `data.frame` that has combines the information in 
both of the other `data.frame`s. 

```{r}
results <- experiments %>%
  left_join(treatment_codes) # Keeps all rows in experiments and only the relevant rows in treatment_codes
```

You were told that it was combining by code (which is correct), but you might
have wanted to enforce that yourself.

```{r}
results2 <- experiments %>%
  left_join(treatment_codes, by="code")
all.equal(results, results2)
```


The warning here is not important, although it is annoying. 
To avoid these warnings, I often use 

```{r}
treatment_codes <- read.csv("treatment_codes.csv", stringsAsFactors = FALSE)
experiments     <- read.csv("experiments.csv",     stringsAsFactors = FALSE)
```

or, using the 
[readr](https://cran.r-project.org/web/packages/readr/index.html) package. 

```{r, eval=FALSE}
library(readr)
treatment_codes <- read_csv("treatment_codes.csv")
experiments     <- read_csv("experiments.csv")
```



There are other ways of joining `data.frame`s in R, but `left_join` is by far
the one that I use the most often. 



#### Activity - Putting it all together

Perform the following:

1. Combine the `experiment` and `treatment` `data.frame`s by column `code`.
1. Remove the `code` column (as it is no longer needed).
1. Subset to include only observations that

  - use the "last owl farthest" strategy,
  - have four or more owls, and
  - three players.
  
1. Calculate the number of games and win proportion after grouping by 

  - number of owls
  - number of players
  - number of cards per player

<details><summary>Click for solution</summary>  
```{r, purl=FALSE}
experiments %>%
  left_join(treatment_codes, by="code") %>%
  select(-code) %>%
  filter(strategy == "last_owl_farthest", 
         n_owls >= 4,
         n_players == 3) %>%
  group_by(n_players, n_owls, n_cards_per_player) %>%
  summarize(
    n_games = n(),
    win_proportion = mean(win))
```
</details>



### `dplyr` cheat sheet

[Here](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) 
is a two-page summary (cheat sheet) of many `dplyr` commands.
