---
layout: page
title: STAT 486/586
tagline: "Basic R - Data Wrangling"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](07-wrangling.R)

We will start with loading up the necessary packages using the `library()` 
command.

```{r}
library("tidyverse")
```

# Read data 

Before, we can start reading data in from our file system, 
we need to 1) understand the concept of a working directory and 
2) have a file that we can read. 

## Working directory



```{r}
getwd()
```

### Set working directory

You can set the working directory in RStudio by going to 

    Session > Set Working Directory > Choose Working Directory

Alternatively, we can use the `setwd()` function. 

```{r, eval=FALSE}
?setwd
```

Use RStudio projects!! This will set your working directory to the same 
folder every time you open that project. 


## Write csv

In order to read a file, we need an appropriate file. 
Comma-separated value (csv) files are a de-facto standard for transporting 
data. 
In this file type, each column is separated by a comma and each row is 
separated by a newline. 

```{r}
readr::write_csv(ToothGrowth, file = "ToothGrowth.csv")
```



## Read csv

```{r}
d <- readr::read_csv("ToothGrowth.csv")
```

```{r}
all.equal(ToothGrowth, d)
```

```{r}
str(ToothGrowth)
str(d)
```

### Options

```{r}
d <- readr::read_csv("ToothGrowth.csv",
                     col_types = cols(
                       len = col_double(),
                       supp = col_factor(),
                       dose = col_double()
                     ))
```

## Read Excel

Check out 

```{r, eval=FALSE}
install.packages("readxl")
?readxl::read_excel
```


## Databases

To read other types of files, including databases, 
start with [these suggestions in R4DS](https://r4ds.had.co.nz/data-import.html#other-types-of-data). 


# CSV file format

There is no official standard for CSV file formats, 
but there are attempts to define a standard or at least outline what is standard.

[Library of Congress](https://www.loc.gov/preservation/digital/formats/fdd/fdd000323.shtml)

[The Internet Society](https://www.rfc-editor.org/rfc/rfc4180)

## Formatting

- commas (or semi-colons)
- line return (CR or LF, mabye CRLF)
- header (optional)
- quotes (optional)


## Bad formatting

### Additional rows

The file format only suggests an (optional) header row. 
Otherwise all rows should strictly be data. 

Some (Canvas....ahem!) add additional row(s) between the header and the data. 
This should not be done as it makes it difficult for standard software
including R to read it since data is assumed to immediately come after the
header.

If you want to include additional rows, you can include rows above the header
row. 
See the `skip` argument of `readr::read_csv()`.

#### Subtotals

A common use of spreadsheets is to calculate subtotals either by adding
additional rows or columns.
Although these can be removed, it is annoying and therefore should be 
avoided. 
Simply calculate subtotals in a different Excel sheet or copy the data over
to a new Excel sheet. 




# Data wrangling

Data often come in multiple files and need to be *wrangled* to get it into the 
format necessary for analysis. 



## Bind

A common issue is that data are created on a periodic basis and each period
has its own file. 
In these circumstances, you need to *bind* the data together. 

### Bind rows

```{r}
dA <- ToothGrowth %>% filter(supp == "VC")
dB <- ToothGrowth %>% filter(supp == "OJ")
bind_rows(dA, dB)
```

```{r}
d <- list(
  A = ToothGrowth %>% filter(supp == "VC", dose == 0.5),
  B = ToothGrowth %>% filter(supp == "VC", dose == 1),
  C = ToothGrowth %>% filter(supp == "VC", dose == 2),
  D = ToothGrowth %>% filter(supp == "OJ", dose == 0.5),
  E = ToothGrowth %>% filter(supp == "OJ", dose == 1),
  F = ToothGrowth %>% filter(supp == "OJ", dose == 2)
)
```

```{r}
class(d)
d[[1]]
d$B
```

```{r}
bind_rows(d)
```

### Bind columns

Hmm...maybe don't bother??

## Join

Data sets are often in multiple files with each file contain some of the
information. 

### Left_join

The most common join operation I use is `left_join()`. 

Rather than refactoring the `supp` column in `ToothGrowth`, 
I would probably just create a new tibble that contains the necessary 
information. 

```{r}
suppTranslation <- tibble(
  supp = c("VC","OJ"),
  `Delivery Method` = c("Ascorbic Acid","Orange Juice")
)
```

Now, I can simply `left_join()` these two data sets:

```{r}
left_join(ToothGrowth, suppTranslation, by = "supp")
```

### Right_join

Of course, if the tibbles were in a different order 
(perhaps due to a data pipeline), I could use `right_join()`. 

```{r}
right_join(suppTranslation, ToothGrowth, by = "supp")
```


### Anti-join

One way to filter observations is to use the `dplyr::filter()` function.
If the observations you want to remove are already part of a data.frame,
then you can use the `anti_join()` function. 

```{r}
len10 <- filter(ToothGrowth, len < 10)

anti_join(ToothGrowth, len10, by = c("len", "supp", "dose"))
```
### Other joins

There are also `inner_join()`, `full_join()`, `semi_join()`, and 
`nest_join()`. I don't use these, but probably I should.


## Transform

Another very common need is to convert data from a relatively *wide* format
to a *long* format for the purpose of visualization or modeling. 
Wide is generally a more efficient storage structure while most statistical
software expects a long format. 

For example, the `AirPassengers` data.frame contains monthly air passengers
across years. 

```{r}
AirPassengers
```

Formally, `AirPassengers` is a time series object and thus R treats it a bit
differently. 
Let's convert it to a data.frame. 
Can you understand the piped code below?

```{r}
months <- c("Jan","Feb","Mar","Apr","May","Jun",
             "July","Aug","Sep","Oct","Nov","Dec")

d <- AirPassengers %>% 
  matrix(ncol=12) %>%
  as.data.frame() %>%
  setNames(months) %>%
  mutate(year = 1949:1960) %>%
  relocate(year)
```

Now we have a wide data frame, but we will typically want a long data.frame
for visualization or analysis. 

### Longer

To convert to a longer data.frame where each row is a single observation, 
we can use `pivot_longer()`. 

```{r}
dl <- d %>%
  pivot_longer(Jan:Dec, names_to = "month", values_to = "passengers") %>%
  mutate(month = factor(month, levels = months)) # to order the months
```

```{r}
table(dl$month)
```

#### Create monthly totals

```{r}
dl %>%
  group_by(month) %>%
  summarize(
    n = n(),
    mean = mean(passengers),
    sd = sd(passengers)
  ) 
```

Plot yearly comparisons

```{r}
ggplot(dl, aes(x = month, y = passengers, 
               group = year, color = year)) +
  geom_line() + 
  scale_y_log10() +
  theme_bw()
```

Fit a regression model

```{r}
m <- lm(log(passengers) ~ year + month, data = dl)
summary(m)
```

### Wider

Less often, we want to convert back to a wide format.

```{r}
dl %>% pivot_wider(id_cols = year, 
                   names_from = "month", 
                   values_from = "passengers")
```


# Binary R file formats

Generally, I don't suggest story data in binary formats,
but these formats can be useful to store intermediate data.
For example, if there is some important results from a statistical 
analysis that takes a long time to perform 
(I'm looking at you Bayesians)
you might want to store the results in a binary format. 

## RData

There are two functions that will save `RData` files:
`save()` and `save.image()`. 
The latter will save everything in the environment while the former will 
only save what it is specifically told to save. 
Both of these are read using the `load()` function. 

```{r}
a <- 1
save(a, file = "a.RData")
```

## RDS

An RDS file contains a single R object. 
These files can be written using `saveRDS()` and read using `readRDS()`.

```{r}
saveRDS(a, file = "a.RDS")
rm(a)
```

When you read this file, you need to save it into an R object. 

```{r}
b <- readRDS("a.RDS")
```

## Cleanup 

```{r}
unlink("ToothGrowth.csv")
unlink("a.RData")
unlink("a.RDS")
```



# Other resources

[Data transformation cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-transformation.pdf)
