---
title: "Lab08 - Categorical explanatory variables"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


To follow along, use the [lab08 code](lab08.R) and make sure the following 
package is installed:

- Sleuth3

You can use the following code to perform the installation:

```{r install_packages, eval=FALSE}
install.packages("Sleuth3")
```

Now load the packages

```{r load_packages}
library("tidyverse")
library("Sleuth3")
```

## Overview

Categorical variables in R are represented either by a 

- character
- factor

A factor is a special character variable that 

- has an underlying numerical representation,
- a lookup table to find the values, and
- an inherent ordering.




## Characters

Let's create a test data set.

```{r}
n <- 12
test <- data.frame(y = rnorm(n))
test$char <- rep(c("a","b","A","B"), each=3)
```

If you look at the type of `char`, it will be a character (vector). 

```{r}
str(test$char)
```

If you try to run a `summary()`, 
you won't get much.

```{r}
summary(test)
summary(test$char)
```

Instead, you want to use the `table()` function.

```{r}
table(test$char)
```

When you use the `readr::read_csv()` function to read in data,
character columns are left as characters.
First we need to write the csv file.

```{r}
write.csv(test, file = "test.csv", row.names = FALSE)
```

Now let's read it in. 

```{r}
test2 <- readr::read_csv("test.csv")
```
```{r}
summary(test2)
```

#### `readr::read_csv()` Activity

Use the `case0501` data set from the `Sleuth3` package to practice 
writing to a csv file and reading from the csv file.
Verify that Diet is read in as a character vector. 

<div id="readcsv_activity_solution" style="display:none"> 
```{r, purl=FALSE}
write.csv(Sleuth3::case0501, file = "case0501.csv", row.names = FALSE)
tmp <- readr::read_csv("case0501.csv")
summary(tmp)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('readcsv_activity_solution') .style.display=='none') {document.getElementById('readcsv_activity_solution') .style.display=''}else{document.getElementById('readcsv_activity_solution') .style.display='none'}">Show/Hide Solution</button>




## Factors

When you read in data using the `read.csv()` function, 
all character vectors will automatically be converted to factor vectors. 

```{r}
case0501 <- read.csv(file = "case0501.csv") # We are overwriting the Sleuth3::case0501 data.frame
```

Note that `Diet` is a factor. 

```{r}
str(case0501$Diet)
```

Although printing out this variable makes it look like a character:

```{r}
case0501$Diet
```

There is actually an internal representation via integers:

```{r}
as.numeric(case0501$Diet)
```

This representation can be accessed using `levels()`:

```{r}
nlevels(case0501$Diet)
levels(case0501$Diet)
```

The ordering here is alphabetical.

When you run `summary()` or `table()`, 
R uses this ordering for displaying results:

```{r}
summary(case0501$Diet)
table(  case0501$Diet)
```

You can convert factors into characters

```{r}
case0501$Diet_char <- as.character(case0501$Diet)
str(case0501)
```

and characters back to factors

```{r}
case0501$Diet_fact <- as.factor(case0501$Diet_char)
str(case0501)
all.equal(case0501$Diet, case0501$Diet_fact)
```


#### Activity

Practice reading and writing files using `read.csv()` and `readr::read_csv()`.
Specifically, do the following

1. Write `case0501` to a csv using `write.csv()`. Do not write the rownames.
1. Read in the created csv file using both `read.csv()` and `readr::read_csv()`. What differences are there?
1. Use the help file for `read.csv()` to read in the data without converting Diet 
to a factor.
1. Convert Diet in the `readr::read_csv()` function to a factor. 
1. Write `case0501` to a csv using `readr::write_csv()` and see what the differences are
from the first step.

<div id="readcsv2_activity_solution" style="display:none"> 
Here is what I would do
```{r, purl=FALSE, eval=FALSE}
# Write csv file
write.csv(case0501, file = "tmp.csv", row.names = FALSE)

# Read csv file
d  <- read.csv("tmp.csv") 
d2 <- readr::read_csv("tmp.csv") 

# Don't convert to factor when reading in
d  <- read.csv("tmp.csv", stringsAsFactors = FALSE)

# Convert Diet into a factor
d2 <- d2 %>%
  mutate(Diet = factor(Diet))

readr::write_csv(case0501, path = "tmp2.csv") # Automatically doesn't include rownames
```
Depending on your application, you may decide that `read.csv()` is what you 
typically want to use or you may decide the `readr::read_csv()` is what you 
typically want to use.
</div>
<button title="Show a solution" type="button" onclick="if(document.getElementById('readcsv2_activity_solution') .style.display=='none') {document.getElementById('readcsv2_activity_solution') .style.display=''}else{document.getElementById('readcsv2_activity_solution') .style.display='none'}">Show/Hide Solution</button>




### Creating factors

We have already seen that you can create factors using the `as.factor()` 
function. 
But sometimes we want a bit more control. 
In particular, we may be interested in affecting the ordering of the 
levels of the factor. 

To adjust the ordering of the levels of the factor, 
use the `levels` argument of the `factor()` function.

```{r}
newDiet = factor(case0501$Diet, 
                 levels = c("NP","N/N85","N/R40","N/R50","lopro","R/R50"))
table(newDiet) # compare to table(case0501$Diet)
```



## Using categorical variables in a regression

Let's make sure we have the correct `data.frame` for the following. 

```{r}
case0501 <- Sleuth3::case0501 %>%
  mutate(Diet_char = as.character(Diet),
         Diet_fact = as.factor(Diet_char))
```

### Using a character vector as the explanatory variable

Let's perform the Lifetime-Diet ``regression'' using the character vector for
Diet. 

```{r}
m <- lm(Lifetime ~ Diet_char, data = case0501)
summary(m)
```

Notice that it automatically constructed all our indicator variables. 
It automatically chose the lopro diet as the reference level (because it came
first alphabetically). 

If you leave Diet as a character vector then you cannot control the reference 
level. 

### Using a factor vector as the explanatory variable

Let's perform the Lifetime-Diet ``regression'' using the factor vector for
Diet. 

```{r}
m <- lm(Lifetime ~ Diet_fact, data = case0501)
summary(m)
```

These results are identical to the results using the character vector. 

### Adjusting the reference level

Now suppose you want to run the regression using N/N85 (normal diet) as the 
reference level. 
Do this using the `relevel()` function and its `ref` argument.

```{r}
case0501$Diet_fact <- relevel(case0501$Diet_fact, ref = "N/N85")
table(case0501$Diet_fact)
```

Notice how N/N85 is now the first level. 

Now  when you run the regression the reference level will be N/N85. 

```{r}
m <- lm(Lifetime ~ Diet_fact, data = case0501)
summary(m)
```

The authors of the `Sleuth3` package had already made N/N85 the reference level
so if you run the regression with the original data, N/N85 will be the 
reference level.

```{r}
m <- lm(Lifetime ~ Diet, data = Sleuth3::case0501)
summary(m)
```


#### Check the regression activity

Verify that the regression is correct by constructing indicator variables for 
the non-reference level explanatory variable levels and running the regression
to reproduce the results above.

<div id="factor_activity_solution" style="display:none"> 
```{r, purl=FALSE}
d <- case0501 %>% 
  mutate(X1 = Diet == "N/R40",
         X2 = Diet == "N/R50",
         X3 = Diet == "NP",
         X4 = Diet == "R/R50",
         X5 = Diet == "lopro")

m2 <- lm(Lifetime ~ X1+X2+X3+X4+X5, data = d)
summary(m2)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('factor_activity_solution') .style.display=='none') {document.getElementById('factor_activity_solution') .style.display=''}else{document.getElementById('factor_activity_solution') .style.display='none'}">Show/Hide Solution</button>



#### Data analysis activity 

Use the `ex0518` `data.frame` from the `Sleuth3` package.

Adjust the order of the levels of treatment to have increasing CPFA and make
"Control" the reference level. 
Create a plot of protein versus treatment. 
Run a regression of protein on treatment to determine which CPFA levels have 
a mean protein level that is significantly different from the control. 
Look at the diagnostic plots to determine if there are any clear violations of
model assumptions.


<div id="data_analysis_activity_solution" style="display:none"> 

Get the data and adjust the levels of treatment. 

```{r, purl=FALSE}
d <- Sleuth3::ex0518 %>%
  mutate(Treatment = factor(Treatment, 
                            levels = c("Control","CPFA50","CPFA150",
                                       "CPFA300","CPFA450","CPFA600")))
```

Now create the plot.

```{r, purl=FALSE}
library(ggplot2)
ggplot(d, aes(Treatment, Protein)) +
  geom_jitter(height=0, width=.25) +
  theme_bw()
```

Notice that the ordering on the x-axis is according to the ordering of the 
levels of treatment. 

Now run the regression. 

```{r, purl=FALSE}
m <- lm(Protein ~ Treatment, data = d)
summary(m)
```

Based on this regression analysis, CPFA levels at 300 or above have 
significantly different mean protein levels. 

Finally look at the diagnostic plots.

```{r, purl=FALSE}
opar = par(mfrow=c(2,3))
plot(m, 1:6, ask=FALSE)
par(opar)
```

There is no clear violation here. 

</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('data_analysis_activity_solution') .style.display=='none') {document.getElementById('data_analysis_activity_solution') .style.display=''}else{document.getElementById('data_analysis_activity_solution') .style.display='none'}">Show/Hide Solution</button>


#### Regression using continuous explanatory variable activity

Use the previous data to construct a continuous explanatory variable that has 
the CPFA level. 
Then run a regression using this explanatory variable and assess model 
assumptions using a diagnostic plot. 

<div id="data_analysis2_activity_solution" style="display:none"> 

I'm not necessarily expecting you to come up with this approach that uses 
`gsub()` and `as.numeric()`. 

```{r, purl=FALSE}
d <- d %>%
  mutate(cpfa = as.numeric(gsub("CPFA","", Treatment)),
         cpfa = ifelse(Treatment == "Control", 0, cpfa))
summary(d)
```

Run the regression 

```{r, purl=FALSE}
m <- lm(Protein ~ cpfa, data = d)

opar = par(mfrow=c(2,3))
plot(m, 1:6, ask = FALSE)
par(opar)

plot(d$cpfa, m$residuals)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('data_analysis2_activity_solution') .style.display=='none') {document.getElementById('data_analysis2_activity_solution') .style.display=''}else{document.getElementById('data_analysis2_activity_solution') .style.display='none'}">Show/Hide Solution</button>


