---
title: "Lab03 - Probability Distributions in R"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


To follow along, use the [lab03 code](lab03.R).

# Probability distributions in R

Many of the standard probability distributions have functions in R to calculate:

 - probability density/mass function (d)
 - cumulative distribution function (p)
 - quantile function (q)
 - random draws (r)
 
The letter after each indicates the letter the function will start with for 
all distributions.
 

## Discrete distributions

If the support of the random variable is a finite or countably infinite number
of values, then the random variable is discrete.
Discrete random variables have a 
[probability mass function (pmf)](https://en.wikipedia.org/wiki/Probability_mass_function).
This pmf gives the probability that a random variable will take on each value 
in its support. 
The [cumulative distribution function (cdf)](https://en.wikipedia.org/wiki/Cumulative_distribution_function)
provides the probability the random variable is less than or equal to a particular
value.
The quantile function is the inverse of the cumulative distribution function, 
i.e. you provide a probability and the quantile function returns the value of the
random variable such the cdf will return that probability.

### Binomial distribution

Whenever you repeat an experiment and you assume 

1. the probability of success is the same each time and 
1. each trial of the experiment is independent of the rest (as long as you know the probability of success).

Then, if you record the number of success out of the total, you have a 
[binomial distribution](https://en.wikipedia.org/wiki/Binomial_distribution).

For example, consider an experiment with probability of success of 0.7 and 13
trials, i.e. $X\sim Bin(13,0.7)$. 

We can use the pmf to calculate the probability of a particular outcome
of the experiment.
For example, what is the probability of seeing 6 successes?
We can use the `dbinom` function. 

```{r}
n <- 13
p <- 0.7
dbinom(6, size = n, prob = p)
```

The entire pmf is 

```{r}
x <- 0:n
plot(x, dbinom(x, size = n, prob = p), main = "Probability mass function for Bin(13,0.7)")
```
 
If we want to calculate the probability of observing an outcome less than 
or equal to a particular value, we can use the cumulative distribution function.
For example, what is the probability of observing 9 or fewer successes?

```{r}
pbinom(9, size = n, prob = p)
```
 
Here is the entire cdf. 

```{r}
plot(x, pbinom(x, size = n, prob = p), type="s", main = "Cumulative distribution function for Bin(13,0.7)")
```

For a discrete random variable, the cdf is a step function since the function 
*jumps* whenever it comes across a value in the support for the random variable.

Here is the quantile function using the `qbinom` function:

```{r}
p_seq <- seq(from = 0, to = 1, length = 101)
plot(p_seq, qbinom(p_seq, size = n, prob = p), type="s", main = "Quantile function for Bin(13,0.7)")
```
 
Finally, we can draw random values from this binomial distribution using the 
`rbinom` function. 

```{r}
draws <- rbinom(100, size = n, prob = p) # draw 100
brks  <- (0:(n+1)) - 0.5
hist(draws, breaks = brks, main = "Random draws from Bin(13,0.7)") 
```
 
Rather than having the number of draws, we often want the percentage of draws. 
For reference, we can add the true probabilities using the pmf 
(shown in red).

```{r}
hist(draws, breaks = brks, probability = TRUE)
points(x, dbinom(x, size = n, prob = p), col="red")
```


#### Activity

Plot the pmf, cdf, and quantile function for the binomial distribution with 
probability of success 0.25 and 39 trials, i.e. $X\sim Bin(39,0.25)$. 
Then sample 999 random binomials with 39 trials and probability of success 0.25
and plot them on a histogram with the true probability mass function.

<div id="binomial_activity_solution" style="display:none"> 
```{r, purl=FALSE}
n <- 39
p <- 0.25
x <- 0:n

plot(x, dbinom(x, size = n, prob = p),                     main = 'Probability mass function for Bin(39,0.25)')
plot(x, pbinom(x, size = n, prob = p),         type = 's', main = 'Cumulative distribution function for Bin(39, 0.25)')
plot(p_seq, qbinom(p_seq, size = n, prob = p), type = 's', main = 'Quantile function for Bin(39, 0.25)')

draws <- rbinom(999, size = n, prob = p)
brks  <- (0:(n+1)) - 0.5
hist(draws, breaks = brks, probability = TRUE, main = "Random draws from Bin(39,0.25)")
points(x, dbinom(x, size = n, prob = p), col="red")
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('binomial_activity_solution') .style.display=='none') {document.getElementById('binomial_activity_solution') .style.display=''}else{document.getElementById('binomial_activity_solution') .style.display='none'}">Show/Hide Solution</button>



 
 
### Poisson distribution

While the binomial distribution has an upper limit ($n$), 
we sometimes run an experiment and are counting *successes* without any 
**technical** upper limit. 
These experiments are usually run for some amount of time or over some amount
of space.
For example, 

- the number of photos observed by a detector in a minute
- the number of times an a/c unit comes on in an hour
- the number of buildings in a square mile
- the number of transitors on a circuit board

When there is no technical upper limit (even if the probability of more is 
extremely small), then a [Poisson distribution](https://en.wikipedia.org/wiki/Poisson_distribution)
can be used. 
The Poisson distribution has a single parameter, the rate that describes, on
average, how many of the *things* are expected to be observed.
We write $X\sim Po(\lambda)$ where $\lambda$ is the rate parameter.

Suppose we record the number of network failures in a day and on average we 
see 2 failures per day.
The number of network failures in a day has no upper limit, so we'll use the 
Poisson distribution. 

Here is the pmf:

```{r}
rate <- 2
x <- 0:10 # with no upper limit we need to decide on an upper limit

plot(x, dpois(x, lambda = rate), main = "Probability mass function for Po(2)") 
```
 
Here is the cdf:

```{r}
plot(x, ppois(x, lambda = rate), type="s", main = "Cumulative distribution function for Po(2)")
```
 
Here is the quantile function:

```{r}
plot(p_seq, qpois(p_seq, lambda = rate), type="s", ylim=c(0,10), main = "Quantile function for Po(2)") # Change the y limits for comparison purposes
```

And random draws

```{r}
draws <- rpois(100, lambda = rate)

hist(draws, breaks = (0:(max(draws)+1)) - 0.5, probability = TRUE, main = "Random draws from Po(2)")
points(x, dpois(x, lambda = rate), col="red")
```


#### Activity

Plot the pmf, cdf, and quantile function for a Poisson distribution with rate
23, i.e. $X\sim Po(23)$.
Then sample 999 Poisson random variables with rate 23
and plot them on a histogram with the true probability mass function.

<div id="poisson_activity_solution" style="display:none"> 
```{r, purl=FALSE}
rate <- 23
x <- 5:40 # with no upper limit we need to decide on an upper limit

plot(x, dpois(x, lambda = rate),                   main = 'Probability mass function for Po(23)')
plot(x, ppois(x, lambda = rate),         type="s", main = 'Cumulative distribution function for Po(23)')
plot(p_seq, qpois(p_seq, lambda = rate), type="s", main = 'Quantile function for Po(23)')

draws <- rpois(999, lambda = rate) 
hist(draws, breaks = (min(draws):(max(draws)+1)) - 0.5, probability = TRUE, main = "Random draws from Po(23)")
points(x, dpois(x, lambda = rate), col="red")
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('poisson_activity_solution') .style.display=='none') {document.getElementById('poisson_activity_solution') .style.display=''}else{document.getElementById('poisson_activity_solution') .style.display='none'}">Show/Hide Solution</button>




## Continuous distributions

In constrast to discrete random variables, 
[continuous random variables](http://stattrek.com/probability-distributions/discrete-continuous.aspx?Tutorial=Stat)
can take on an uncountably infinite number of values. 
The easiest way for this to happen is that the random variable can take on any
value between two specified values (and infinity counts). 

Continuous random variables have a 
[probability density function (pdf)](https://en.wikipedia.org/wiki/Probability_density_function)
instead of a pmf. 
When integrated from a to b, this pdf gives the probability the random variable
will take on a value between a and b.
Continuous random variables still have a cdf, quantile function, and random
generator that all still have the same interpretation. 
 
### Uniform distribution
 
The simplest continuous random variable is the uniform random variable. 
If $Y$ is a random variable and it is uniformly distributed between a and b, 
then we write $Y\sim Unif(a,b)$ and this means that $Y$ can take on any value 
between a and b with equal probability. 
 
The probability density function for a uniform random variable is zero outside
of a and b (indicating that the random variable cannot take values below a
or above b) and is constant at a value of 1/(b-a) from a to b. 
 
```{r}
a <- 0
b <- 1

# The curve function expects you to give a function of `x` and then it 
# (internally) creates a sequence of values from `from` and to `to` and creates
# plots similar to what we had before, but using a line rather than points.
curve(dunif(x, min = a, max = b), from = -1, to = 2,
      xlab='y', ylab='f(y)', main='Probability density function for Unif(0,1)')
```
 
The cumulative distribution function is the integral from negative infinite 
up to y of the probability density function. 
Thus it indicates the probability the random variables will take on values less
than y, i.e. $P(Y<y)$. 
Since the probability density function for the uniform is constant, 
the integral is simply zero from negative infinite up to a, 
then a straight line from (a,0) up to (b,1) and then constant after that.

```{r}
curve(punif(x, min = a, max = b), from = -1, to = 2,
      xlab='y', ylab='F(y)', main='Cumulative distribution function for Unif(0,1)')
```

The quantile function is the inverse of the cumulative distribution function.
For a given probability p, it finds the value such that the 
probability the random variable is below that value is p.

```{r}
curve(qunif(x, min = a, max = b), from = 0, to = 1,
      xlab='p', ylab='F^{-1}(p)', main='Quantile function for Unif(0,1)')
```

To draw random values from the distribution, use the `r` version of the 
function.
For instance, here is 100 random Unif(a,b) draws represented as a histogram.

```{r}
random_uniforms <- runif(100, min = a, max = b)
hist(random_uniforms, probability = TRUE, main = "Random draws from Unif(0,1)")
curve(dunif(x, min = a, max = b), add = TRUE, col="red")
```


#### Activity

Plot the pdf, cdf, and quantile function for a uniform distribution on the 
interval (13, 65), i.e. $X\sim Unif(13,65)$. 
Then sample 999 random uniforms on the interval (13,65) and plot a histogram
of these draws with the probability density function.


<div id="uniform_activity_solution" style="display:none"> 
```{r, purl=FALSE}
a <- 13
b <- 65

opar = par(mfrow=c(2,2)) # Create a 2x2 grid of figures
curve(dunif(x, min = a, max = b), from = a-1, to = b+1,
      xlab='y', ylab='f(y)', main='Probability density function for Unif(13,65)')
curve(punif(x, min = a, max = b), from = a-1, to = b+1,
      xlab='y', ylab='f(y)', main='Cumulative distribution function for Unif(13,65)')
curve(qunif(x, min = a, max = b), from = 0, to = 1,
      xlab='y', ylab='f(y)', main='Quantile function for Unif(13,65)')
random_uniforms <- runif(999, min = a, max = b)
hist(random_uniforms, probability = TRUE, main = "Random draws from Unif(13,65)")
curve(dunif(x, min = a, max = b), add=TRUE, col="red")
par(opar)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('uniform_activity_solution') .style.display=='none') {document.getElementById('uniform_activity_solution') .style.display=''}else{document.getElementById('uniform_activity_solution') .style.display='none'}">Show/Hide Solution</button>



### Normal distribution

The most important distribution for this course is the 
[normal (Gaussian) distribution](https://en.wikipedia.org/wiki/Normal_distribution).
The normal distribution has two parameters: the mean ($\mu$) and the variance 
($\sigma^2$) and we write $X\sim N(\mu,\sigma^2)$.
If $\mu=0$ and $\sigma^2=1$, we call the associated random variable a **standard
normal**.
The probability density function for the normal distribution is the well-known
bell-shaped curve.

```{r}
mu    <- 0
sigma <- 1 # standard deviation

curve(dnorm(x, mean = mu, sd = sigma), # notice the 3rd argument is the sd
      from = -4, to = 4,
      main = "PDF for a standard normal")
```

The quantile function has a sigmoid shape.

```{r}
curve(pnorm(x, mean = mu, sd = sigma), 
      from = -4, to = 4,
      main = "CDF for a standard normal",
      ylab = "F(x)")
```

The quantile function will be used later in the semester to help construct
confidence/credible intervals.

```{r}
curve(qnorm(x, mean = mu, sd = sigma),
      from = 0, to = 1, 
      main = "Quantile function for a standard normal")
```

Then we can take random draws

```{r}
draws <- rnorm(100, mean = mu, sd = sigma)
hist(draws, probability = TRUE)
curve(dnorm(x, mean = mu, sd = sigma), add = TRUE, col = "red")
```

#### Activity

Plot the pdf, cdf, and quantile function for a normal distribution with mean 
-4 and variance 3, i.e. $X\sim N(-4,3)$. 
Then sample 999 random N(-4,3) and plot a histogram
of these draws with the probability density function.


<div id="normal_activity_solution" style="display:none"> 
```{r, purl=FALSE}
mu    <- -4
sigma <- sqrt(3) # standard deviation!!!

opar = par(mfrow=c(2,2))
curve(dnorm(x, mean = mu, sd = sigma), 
      from = mu-4*sigma, to = mu+4*sigma,
      main = "PDF for a normal")

curve(pnorm(x, mean = mu, sd = sigma), 
      from = mu-4*sigma, to = mu+4*sigma,
      main = "CDF for a normal")

curve(qnorm(x, mean = mu, sd = sigma),
      from = 0, to = 1, 
      main = "Quantile function for a normal")

draws <- rnorm(999, mean = mu, sd = sigma)
hist(draws, probability = TRUE)
curve(dnorm(x, mean = mu, sd = sigma), add = TRUE, col = "red")
par(opar)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('normal_activity_solution') .style.display=='none') {document.getElementById('normal_activity_solution') .style.display=''}else{document.getElementById('normal_activity_solution') .style.display='none'}">Show/Hide Solution</button>


#### Quantiles of a normal distribution

For any normal random variable, determine the probability it is 

- within 1 standard deviation of its mean
- within 2 standard deviation of its mean
- within 3 standard deviation of its mean

<div id="normal_quantiles_activity_solution" style="display:none"> 
Since this is apparently for any normal, we will just use the standard normal. 
So if $X\sim N(\mu, \sigma^2)$, we must determine 
$P(|X|<c) = P(-c < X < c) = P(X<c) - P(X< -c)$ for $c$ in 1,2,3.
```{r, purl=FALSE}
# Try any values for mu and sigma
mu    <- 0
sigma <- 1

pnorm(mu+1*sigma, mean = mu, sd = sigma) - pnorm(mu-1*sigma, mean = mu, sd = sigma)
pnorm(mu+2*sigma, mean = mu, sd = sigma) - pnorm(mu-2*sigma, mean = mu, sd = sigma)
pnorm(mu+3*sigma, mean = mu, sd = sigma) - pnorm(mu-3*sigma, mean = mu, sd = sigma)

```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('normal_quantiles_activity_solution') .style.display=='none') {document.getElementById('normal_quantiles_activity_solution') .style.display=''}else{document.getElementById('normal_quantiles_activity_solution') .style.display='none'}">Show/Hide Solution</button>



## Binomial questions of interest

Suppose a manufacturing line produces temperature sensors and has a sensor
failure proportion equal to 1.5%. 
In a given day, the plant tests 70 sensors. 
If we assume the sensors are independent (given the failure proportion), then 
$X\sim Bin(70,0.15)$. 

Answer the following questions:

- What is the probability no sensors fail in one day?
- What is the probability 3 or more sensors fail in one day?

<div id="binomial_questions_activity_solution" style="display:none">

The second question is tricky, because we need $P(X\ge 3) = 1-P(X<3) = 1-P(X\le 2)$. 

```{r, purl=FALSE}
n <- 70
p <- 0.015

dbinom(0, size = n, prob = p)
1-pbinom(2, size = n, prob = p)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('binomial_questions_activity_solution') .style.display=='none') {document.getElementById('binomial_questions_activity_solution') .style.display=''}else{document.getElementById('binomial_questions_activity_solution') .style.display='none'}">Show/Hide Solution</button>

```{r, echo=FALSE, purl=FALSE, results="hide", message=FALSE}
knitr::knit("lab03.Rmd", tangle = TRUE)
```


