---
title: "Lab06 - Built-in analyses in R"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---


To follow along, use the [lab06 code](lab06.R).
You can also save [creativity.csv](creativity.csv) into your working directory
if you want.

## Binomial data

### Bayesian analysis

Let $Y\sim Bin(n,\theta)$.
Recall that our default prior for $\theta$ is $\theta \sim Be(1,1)$, which is
equivalent to a $Unif(0,1)$ prior.
The posterior under this prior is $\theta|y \sim Be(1+y,1+n-y)$. 
In order to perform this analysis, 
we simply use the beta distribution in R. 

Suppose you observe 9 successes out of 13 attempts. 

```{r}
a <- b <- 1
n <- 13
y <- 9
```

The posterior is 

```{r}
curve(dbeta(x, a+y, b+n-y))
```

The posterior expectation is 

```{r}
(a+y)/(b+n-y)
```

A 95\% equal-tail credible interval is 

```{r}
qbeta(c(.025,.975), a+y, b+n-y)
```

The probability that $\theta$ is greater than 0.5, i.e. $P(\theta>0.5|y)$ is 

```{r}
1-pbeta(0.5, a+y, b+n-y)
```

#### Jeffreys prior activity

An alternative prior is called Jeffreys prior and it corresponds to a 
Be(0.5,0.5) prior. 
Suppose you observed 17 successes out of 20 attempts and you are willing to assume 
independence and a common probability of success.
Use Jeffreys prior on this probability of success to do the following 

- Plot the posterior
- Calculate the posterior median
- Calculate a one-sided upper 95% credible interval
- Calculate the probability that the probability of success is greater than 0.9.


<div id="beta_jeffreys_activity_solution" style="display:none"> 
```{r, purl=FALSE}
a <- b <- 0.5
n <- 20
y <- 17
curve(dbeta(x, a+y, b+n-y))
qbeta(0.5, a+y, b+n-y)
c(qbeta(0.05, a+y, b+n-y), 1)
1-pbeta(0.9, a+y, b+n-y)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('beta_jeffreys_activity_solution') .style.display=='none') {document.getElementById('beta_jeffreys_activity_solution') .style.display=''}else{document.getElementById('beta_jeffreys_activity_solution') .style.display='none'}">Show/Hide Solution</button>


### Non-Bayesian analysis

Fortunately many analysis have built in functions in R. 
If you observe 78 successes out of 100 attempts, 
then you can use the `prop.test()` function to generate a number of statistics
automatically.

```{r}
n <- 100
y <- 78
prop.test(y,n)
```

Sometimes you want all of this output, 
but other times you want to extract exactly what you want.
To do this we will save the object created by `prop.test()` 
and use the `names()` or `str()` function to determine what is in this object.

```{r}
pt <- prop.test(y,n)
names(pt)
str(pt)
```
Then we can use the `$` to extract what we want.
The estimate is 

```{r}
pt$estimate
```

An approximate 95\% confidence interval is 

```{r}
pt$conf.int
```

If we wanted an aproximate 90\% one-sided lower confidence interval, 
we can obtain it using 

```{r}
prop.test(y, n, alternative = "less", conf.level = 0.9)$conf.int
```

#### Activity

Suppose you observed 694 success out of 934 attempts. 
Compute a 95\% equal-tail confidence interval using `prop.test` 
and compare this to a 95\% equal-tail confidence interval you construct using
the Central Limit Theorem.

<div id="binomial_clt_activity_solution" style="display:none"> 
```{r, purl=FALSE}
y <- 694
n <- 934
prop.test(y, n)$conf.int

p <- y/n
p +c(-1,1)*qnorm(.975)*sqrt(p*(1-p)/n)

# If you turn off the continuity correction, you will get closer
prop.test(y, n, correct = FALSE)$conf.int
```
If you really want to figure out what the function is doing, 
you can look at the function by just typing `prop.test` and hitting enter.
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('binomial_clt_activity_solution') .style.display=='none') {document.getElementById('binomial_clt_activity_solution') .style.display=''}else{document.getElementById('binomial_clt_activity_solution') .style.display='none'}">Show/Hide Solution</button>

### Small sample sizes

The `prop.test` function uses the Central Limit Theorem to perform its analysis.
When n is small, you should instead use the `binom.test` function.

Suppose you observe 9 successes out of 13 attempts. 

```{r}
n <- 13
y <- 9
binom.test(y, n)
```

#### Activity

Suppose you observe 0 success out of 77 attempts. 
Compare 95% confidence intervals given by `prop.test` and `binom.test` to 
an interval you construct based on the CLT and to 95% credible intervals.

<div id="binom_ci_comparison_activity_solution" style="display:none"> 

So this is a bit of a trick question since there was 0 successes. 
When you run `prop.test` and `binom.test` you are given one-sided confidence 
intervals.
The CLT interval doesn't exist since the standard error is zero.
The appropriate credible interval to use is a one-sided interval.

```{r, purl=FALSE}
y <- 0
n <- 77
prop.test(y, n)$conf.int
binom.test(y, n)$conf.int
qbeta(c(0,.95), 1+y, 1+n-y)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('binom_ci_comparison_activity_solution') .style.display=='none') {document.getElementById('binom_ci_comparison_activity_solution') .style.display=''}else{document.getElementById('binom_ci_comparison_activity_solution') .style.display='none'}">Show/Hide Solution</button>





## Normal data

Suppose $Y_i\stackrel{ind}{\sim} N(\mu,\sigma^2)$. 
Recall that a Bayesian analysis with the default prior 
$p(\mu,\sigma^2) \propto 1/\sigma^2$ provides the same analysis for $\mu$ as a
non-Bayesian analysis. 
That is 

- MLE and Bayes estimator for $\mu$ are the same
- confidence and credible intervals for $\mu$ are exactly the same

This is because the posterior for $\mu$ is 
$$
\mu|y \sim t_{n-1}(\overline{y}, S^2/n)
$$
which means that 
$$
\left.\frac{\mu-\overline{y}}{s/\sqrt{n}} \right|y \sim t_{n-1}(0,1).
$$
while the sampling distribution for $\overline{y}$ is such that 
$$
T=\left.\frac{\overline{Y}-\mu}{S/\sqrt{n}} \right|y \sim t_{n-1}(0,1).
$$
Suppose you observe the following data
```{r}
(y <- rnorm(10, mean = 3.2, sd = 1.1))
```
Then you can manually construct an MLE and Bayesian posterior using `mean()`.

```{r}
(ybar <- mean(y))
```

and a 95% credible/confidence interval using

```{r}
n <- length(y)
s <- sd(y)

ybar + c(-1,1)*qt(.975, df = n-1)*s/sqrt(n)
```

### Built-in analysis

You can use the `t.test()` function to perform this for you.

```{r}
t.test(y)
```

and extract quantities using

```{r}
tt <- t.test(y)
tt$estimate
tt$conf.int
```

### Using the `lm()` function

An alternative to `t.test()` that will be extremely useful throughout the 
remainder of the semester is the `lm()` function.
The `lm` stands for linear model and regression is a special case.

```{r}
lm(y~1) # The 1 stands for a single group, i.e. just an intercept
```

This function is sophisticated enough that very little is reported when you 
just run the function.
Instead, most of the information is in the `summary()` function.

```{r}
m <- lm(y~1)
summary(m)
```

At this point, some of this should start to become familiar. 
The Estimate is the sample mean, the Std. Error is the standard error
of this mean, and the t value is the ratio of these two. 
The residual standard error is the MLE for $\sigma$ and the degrees of freedom
is the number of observations minus the number of parameters for the mean (n-1 
in this case).

The `confint()` function will produce the confidence/credible intervals

```{r}
confint(m)
```

#### Activity

Using the following data, compare the point estimate and confidence/credible
intervals obtained using the `t.test()` and `lm()` functions to estimates and 
intervals you create yourself

```{r}
set.seed(1)
y <- rnorm(1001, mean = 256, sd = 34.6)
```

<div id="normal_activity_solution" style="display:none"> 
```{r, purl=FALSE}
t.test(y)
(m <- lm(y~1))
confint(m)
mean(y) + c(-1,1)*qt(.975, df = length(y)-1)*sd(y)/sqrt(length(y))
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('normal_activity_solution') .style.display=='none') {document.getElementById('normal_activity_solution') .style.display=''}else{document.getElementById('normal_activity_solution') .style.display='none'}">Show/Hide Solution</button>




## Reading data from files

First, let's write some data to files. 

```{r}
# Generate some simulated data
n <- 100
d <- data.frame(rep = 1:n,
                response = sample(c("Yes","No"), n, replace=TRUE, prob = c(.2,.8)),
                measurement = rnorm(n, mean = 55, sd = 12))

# Write it to a file
# make sure you have set your working directory to someplace where you want this
# file to be written
write.csv(d, 
          file = "data.csv",
          row.names = FALSE)
```

Alternatively, you could use the `write_csv()` function in the 
[readr](https://cran.r-project.org/web/packages/readr/index.html) package. 

```{r, eval=FALSE}
install.packages("readr")
library("readr") 
write_csv(d, path = "data.csv")
```

Now let's read this data back in. 

```{r}
my_data <- read.csv("data.csv")
```

Take a look at the data to make sure it looks correct:

```{r}
head(my_data)
str(my_data)
```

### Binomial data

To use `prop.test()` and `binom.test()`, 
you need to calculate the number of successes and the number of attempts.

```{r}
y <- sum(my_data$response == "Yes")
n <- length(my_data$response)
prop.test(y, n)
binom.test(y, n)
```

### Normal data

To analyze the normal data, you can just use `t.test()` directly.

```{r}
t.test(my_data$measurement)
```

The data are already located in a `data.frame`. 
You can use this in the `lm()` function. 

```{r}
m <- lm(measurement~1, data = my_data)
summary(m)
confint(m)
```


#### Online activity

Read in the data at [creativity.csv](creativity.csv) 
and then construct confidence/credible intervals for mean creativity score
for both the Intrinsic and Extrinsic groups. 

<div id="online_activity_solution" style="display:none"> 

There are a variety of ways to do this. 
I will construct two new data frames to contain the Intrinsic and Extrinsic 
data and then construct the intervals.

```{r, purl=FALSE}
creativity <- read.csv("http://www.jarad.me/courses/stat401Eng/labs/lab06/creativity.csv")

library(dplyr)
intrinsic <- creativity %>% filter(Treatment == "Intrinsic")
extrinsic <- creativity %>% filter(Treatment == "Extrinsic")

confint(lm(Score~1, data = intrinsic))
confint(lm(Score~1, data = extrinsic))
```

An alternative approach is to use the `subset` argument

```{r, purl=FALSE}
confint(lm(Score~1, data = creativity, subset = Treatment == "Intrinsic"))
confint(lm(Score~1, data = creativity, subset = Treatment == "Extrinsic"))
```

If you want to find out more about these data, 
take a look at the help file for `case0101` in the 
[Sleuth3](https://cran.r-project.org/web/packages/Sleuth3/index.html) package.

```{r, eval=FALSE}
install.packages("Sleuth3")
library("Sleuth3")
?case0101
```

</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('online_activity_solution') .style.display=='none') {document.getElementById('online_activity_solution') .style.display=''}else{document.getElementById('online_activity_solution') .style.display='none'}">Show/Hide Solution</button>

