---
layout: page
title: STAT 486/586
tagline: "R Objects"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](08-objects.R)



```{r}
library("tidyverse")
```

# Vectors

```{r}
nums <- c(2.5, -3.2, 9, pi, 2, 0.2)
ints <- c(1L, 5L, 99L, 103L)
chrs <- c("my character vector")
lgls <- c(TRUE, FALSE, FALSE, TRUE)
```

## is.vector()

```{r}
is.vector(nums)
is.vector(ints)
is.vector(chrs)
is.vector(lgls)
```

## length()

```{r}
length(nums)
length(ints)
length(chrs)
length(lgls)
```

## typeof()

```{r}
typeof(nums)
typeof(ints)
typeof(chrs)
typeof(lgls)
```

```{r}
ints
typeof(c(1,5,99,103))
```

## mode()

```{r}
mode(nums)
mode(ints)
mode(chrs)
mode(lgls)
```

## storage.mode()

```{r}
storage.mode(nums)
storage.mode(ints)
storage.mode(chrs)
storage.mode(lgls)
```



## class()

```{r}
class(nums)
class(ints)
class(chrs)
class(lgls)
```

[This section in the R manual attempts to disambiguate these different functions](https://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html#Objects).

# Attributes

```{r}
attributes(nums)
```

## names()

```{r}
names(nums) <- LETTERS[1:length(nums)]
nums
names(nums)
```

```{r}
attributes(nums)
```

## dim()

```{r}
dim(nums)
```

```{r}
dim(nums) <- c(2,3)
nums
names(nums)
dim(nums)
attributes(nums)
```

```{r}
is.vector(nums)
typeof(nums)
```

# Matrices

```{r}
is.matrix(nums)
colnames(nums) <- LETTERS[1:ncol(nums)]
rownames(nums) <- letters[1:nrow(nums)]
nums
attributes(nums)
typeof(nums)
mode(nums)
storage.mode(nums)
class(nums)
```

# Data.frame

```{r}
nums <- as.data.frame(nums)
is.matrix(nums)
is.data.frame(nums)
nums
attributes(nums)
typeof(nums)
mode(nums)
storage.mode(nums)
class(nums)
```

# List

```{r}
nums <- as.list(nums)
is.data.frame(nums)
is.list(nums)
length(nums)

nums
nums[[1]]
nums$B
attributes(nums)
typeof(nums)
mode(nums)
storage.mode(nums)
class(nums)
```

```{r}
l <- list(
  x = 1:10,
  y = rnorm(10)
)
l$model <- lm(l$y ~ l$x)

l
l$model
attributes(nums)
typeof(nums)
mode(nums)
storage.mode(nums)
class(nums)

attributes(l$model)
typeof(l$model)
mode(l$model)
storage.mode(l$model)
class(l$model)
```

# Data frames

```{r}
attributes(ToothGrowth)
typeof(ToothGrowth)
mode(ToothGrowth)
storage.mode(ToothGrowth)
class(ToothGrowth)
```

Columns in a data.frame my have their own attributes

```{r}
attributes(ToothGrowth$supp)
```

# Complex

```{r}
comps <- c(1i, 2+2i, 3+4i)
attributes(comps)
typeof(comps)
mode(comps)
storage.mode(comps)
class(comps)
```

# Raw

```{r}
raws <- raw(3)
attributes(raws)
typeof(raws)
mode(raws)
storage.mode(raws)
class(raws)
```


# Factor

Factors are a special type of character object that has an internal integer
representation and a lookup table.

```{r}
is.factor(ToothGrowth$supp)

attributes(ToothGrowth$supp)
typeof(ToothGrowth$supp)
mode(ToothGrowth$supp)
storage.mode(ToothGrowth$supp)
class(ToothGrowth$supp)
```

```{r}
ToothGrowth$supp
summary(ToothGrowth$supp)
```

## Lookup table

```{r}
as.numeric(ToothGrowth$supp)    # integer representation
as.character(ToothGrowth$supp)

nlevels(ToothGrowth$supp)
levels(ToothGrowth$supp)   # LOOKUP table
```

## Reorder levels

By default, the order of levels is alphabetical. 

```{r}
my_char <- c(letters[1:3], LETTERS[1:3])
my_char

my_fact <- as.factor(my_char)
my_fact
levels(my_fact)
```

To rearrange, the levels use `factor()` with the `levels` argument

```{r}
my_fact2 <- factor(my_fact, levels = c(letters[1:3], LETTERS[1:3]))
my_fact2
levels(my_fact2)
```

The ordering of levels in a `ggplot` are based on the ordering in the factor.

```{r}
ggplot(ToothGrowth, aes(x = supp, y = len)) + 
  geom_boxplot()
```

If you use a character vector, the ordering will be alphabetical.


This is particularly important when using numbers as alphabetical ordering
is generally not what you want. 

```{r}
my_fact <- factor(c("a1", "a2", "a10"))
levels(my_fact)
```

## Reference level

By default, R will use the first level as the reference level in a regression
model. 

```{r}
m <- lm(len ~ supp, data = ToothGrowth)
coef(m)
```

If we want a different level to be first, we can just move one level to the 
beginning rather than setting all levels.

```{r}
d <- ToothGrowth %>%
  mutate(supp = relevel(supp, ref = "VC"))

m <- lm(len ~ supp, data = d)
coef(m)
```






# Resources

- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)
- [Hands on Programming with R](https://rstudio-education.github.io/hopr/)
