---
layout: page
title: STAT 486/586
tagline: "Control Flow"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](10-control.R)



```{r}
library("tidyverse")
```


Control structures in R provide conditional flow as well as looping. 
An R `expression` is evaluated within the loop or as the result of a
conditional statement. 

The following are example R expressions

```{r}
# First expression
1+2

# Second expression
a <- 1; b <- 2; a+b

# Third expression
{
  a <- 1
  b <- 2
  a + b
}
```

See 

```{r}
?expression
```

# Conditionals

Conditionals include if-else type statements

## if

```{r}
if (TRUE) {
  print("This was true!")
}
```

```{r}
this <- TRUE
if (this) {
  print("`this` was true!")
}
```

```{r}
if (1<2) {
  print("one is less than two!")
}
if (1>2) {
  print("one is greater than two!")
}
```

Using variables

```{r}
a <- 1
b <- 2
if (a < 2) {
  print("`a` is less than 2!")
}
if (a < b) {
  print("`a` is less than `b`!")
}
```

## if-else

```{r}
if (a < b) {
  print("`a` is less than `b`!")
} else {
  print("`b` is not less than `a`!")
}
```


```{r}
if (a > b) {
  print("`a` is greater than `b`!")
} else {
  print("`a` is not greater than `b`!")
}
```

```{r}
if (a > b) {
  print("`a` is greater than `b`!")
} else if (dplyr::near(a,b)) {
  print("`a` is near `b`!")
} else {
  print("`a` must be greater than b")
}
```

You can omit the brackets { } for single line expressions

```{r}
if (a < b)
  print("`a` is less than `b`!")
```

## ifelse

The `ifelse()` function takes a logical vector as a first argument and then
two scalars. 

```{r}
ifelse(c(TRUE, FALSE, TRUE), "this was true", "this was false")
```


## switch

A rarely used function is `switch()` which implements a case-switch comparison.

```{r}
this <- "a"
switch(this,
       a = "`this` is `a`",
       b = "`this` is `b`",
       "`this` is not `a` or `b`")

this <- "b"
switch(this,
       a = "`this` is `a`",
       b = "`this` is `b`",
       "`this` is not `a` or `b`")

this <- "c"
switch(this,
       a = "`this` is `a`",
       b = "`this` is `b`",
       "`this` is not `a` or `b`")
```



# Loops

There are 3 base types of loops: `for`, `while`, and `repeat`. 


## for 

These loops use the reserved words `for` and `in`, 

### Iterating over integers

The most common use of a `for` loop is to loop over integers. 

```{r}
for (i  in 1:10) {
  print(i)
}
```


### Combine with control structures

```{r}
for (i  in 1:10) {
  if (i > 5)
    print(i)
}
```

### Iterate over non-integers

Can also iterate over non-integers

```{r}
for (d in c(2.3, 3.5, 4.6)) {
  print(d)
}
```

```{r}
for (c in c("my","char","vector")) {
  print(c)
}
```

While these are possible, I find I rarely use them. 


### `seq_along()`

Be careful when iterating over objects that a potentially NULL. 

```{r}
this <- NULL
for (i in 1:length(this)) {
  print(i)
}
```

Since `this` had no length, you probably didn't want to enter the `for` loop
at all. 
To be safe, you can use `seq_along()`.

```{r}
for (i in seq_along(this)) {
  print(i)
}
```

```{r}
my_chars <- c("my","char","vector")
for (i in seq_along(my_chars)) {
  print(paste(i, ":", my_chars[i]))
}
```

### `seq_len()`

For `data.frame`s use `seq_len()` with `nrow()`. 

```{r}
for (i in seq_len(nrow(ToothGrowth))) {
  if (ToothGrowth$supp[i] == "OJ" & 
      near(ToothGrowth$dose[i], 2) &
      ToothGrowth$len[i] > 25) {
    print(ToothGrowth[i,])
  }
}
```

Like with `if` and `else` statements, 
`for` loops can omit the brackets { } for single line expressions.

```{r}
for (i in 1:10) 
  print(i)
```

## while

The `while()` loop can be used to construct while loops.

```{r}
a <- TRUE
while (a) {
  print(a)
  a <- FALSE
}
```

```{r}
i <- 0
while (i < 10) {
  print(i)
  i <- i + 1
}
```

```{r}
x <- 2
while (x < 1) { # Evaluated before the loop
  print("We entered the loop.")
}

while (x < 100) { # Evaluated after each loop
  x <- x*x
  print(x)
}
```

Be careful about infinite loops

```{r, eval=FALSE}
while (TRUE) {
  # do something
}
```

You can typically **Escape** out of infinite loops. 
Often, you will want to make sure the infinite loop never occurs. 

```{r}
max_iterations <- 1000
i <- 1
while (TRUE & (i < max_iterations) ) {
  i <- i + 1
  # Do something
}
print(i)
```


## repeat

An alternative to `while()` is `repeat` combined with `break`. 

```{r}
i <- 10
repeat {
  print(i)
  i <- i + 1
  if (i > 13)
    break
}
```

```{r}
i <- 1
repeat {
  print(i)
  i <- i + 1
  if (i %% 2) { # %% is the mod function, 0 is FALSE and 1 is TRUE
    next        # skips to next iteration of repeat
  }
  if (i > 14)
    break
}
```

