---
layout: page
title: STAT 486/586
tagline: "Basic R - dplyr"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_float: true
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[R code](06-dplyr.R)

We will start with loading up the necessary packages using the `library()` 
command.

```{r}
library("tidyverse")
```

# Data transformation

The dplyr package provides a number of convenience functions for dealing with
`data.frame` objects in R. 
As a reminder, a `data.frame` is a special matrix that allows different
columns to have different data types. 

Although, we will focus on the [dplyr package](https://dplyr.tidyverse.org/), 
you may want to look up the [data.table package](https://cran.r-project.org/web/packages/data.table/index.html). 
Generally, dplyr tries to make the code readable while data.table tries to 
make the most efficient code. 

Technically, the output of a dplyr function is a `tibble` which is a 
specialized `data.frame`. 

## tibble

```{r, eval=FALSE}
?tibble
```

```{r}
ToothGrowth
```

vs 

```{r}
as_tibble(ToothGrowth)
```



## Filtering observations

Filter means to remove a set of observations according to some criterion. 

### base R

```{r}
ToothGrowth[ToothGrowth$supp == "VC", ]
subset(ToothGrowth, supp == "VC")
```

or

```{r}
ToothGrowth[ToothGrowth$len < 10, ]
subset(ToothGrowth, len < 10)
```

```{r}
subset(ToothGrowth, len < 10 & supp == "VC") # using the AND operator &
```

### dplyr::filter

Note that `dplyr::filter` masks `stats::filter()`. 
We will be explicit here and call the `filter` function from the `dplyr` package
using the double colon operator.

```{r, eval=FALSE}
?`::`
```

```{r}
dplyr::filter(ToothGrowth, supp == "VC")
```

```{r}
dplyr::filter(ToothGrowth, len < 10)
```

```{r}
dplyr::filter(ToothGrowth, len < 10, supp == "VC")
```

## arrange

Arrange arranges the `data.frame` according to a variable. 

### base R

```{r}
ToothGrowth[order(ToothGrowth$len), ]
```

What happens if we want to arrange by supp and then by dose?
I don't know. 

### dplyr::arrange

```{r}
arrange(ToothGrowth, len)
```

```{r}
arrange(ToothGrowth, supp, dose)
```

## Selecting variables

Here we will select which columns to keep. 

### base::subset

```{r}
ToothGrowth[, c("len","dose")]
```

```{r}
ToothGrowth[, c(1,3)]
```


```{r}
ToothGrowth[, -2]
```

```{r}
subset(ToothGrowth, select = c(len, dose))
```


### dplyr::select

```{r}
dplyr::select(ToothGrowth, len, dose)
```

```{r}
dplyr::select(ToothGrowth, c(1,3))
```

```{r}
dplyr::select(ToothGrowth, -2)
```

```{r}
dplyr::select(ToothGrowth, -supp)
```

#### tidyselect::starts_with()

When data.frames get large, it is annoying to type all the variables you 
want to keep. 
Often times we want to keep a collection of variables that have certain 
properties in their name. 

```{r}
select(ToothGrowth, starts_with("len"))
```

There are a variety of these helper functions, see

```{r, eval=FALSE}
?starts_with
```

## Renaming variables

Many times variable names needed to be modified. 

### base::names()

```{r}
d <- ToothGrowth
names(d)
names(d) <- c("Length","Delivery","Dose")
head(d)
```

Valid variable names, see

```{r, eval=FALSE}
?make.names
```

```{r}
names(d)[2] <- "Delivery Method"
names(d)
```

This can get you into trouble

```{r, error=TRUE}
d$Delivery Method
```

but you can use "bad" variable names 

```{r}
d$`Delivery Method`
```

### dplyr::rename()

```{r}
d <- ToothGrowth
names(d)

d <- rename(d,
  Length = len,
  `Delivery Method` = supp,
  Dose = dose
)
names(d)
```



## Adding and modifying variables

We will often want to create new variables or modify existing variables. 

### base::`$`

We have already seen the extract operator to extract a column out of a 
data.frame.
Here we will use it to create a new variable. 


```{r, eval=FALSE}
d <- ToothGrowth
d$`Length (mm)` = d$len / 1000 # original length was in microns
head(d)
```

### dplyr::mutate()

```{r}
mutate(ToothGrowth, `Length (mm)` = len)
```


## Data summaries

Often we will want to create summaries of our data.

### base::mean

```{r}
mean(ToothGrowth$len)
```

```{r}
table(ToothGrowth$supp)
```



### dplyr::summarize()

```{r}
dplyr::summarize(ToothGrowth, mean_len = mean(len))
```

```{r}
dplyr::summarize(
  ToothGrowth,
  n_VC = sum(supp == "VC"),
  n_OJ = sum(supp == "OJ"),               
)
```



# Data pipelines

According to [IBM](https://www.ibm.com/topics/data-pipeline), 

> A data pipeline is a method in which raw data is ingested from various data sources and then ported to data store, like a data lake or data warehouse, for analysis. Before data flows into a data repository, it usually undergoes some data processing. This is inclusive of data transformations, such as filtering, masking, and aggregations, which ensure appropriate data integration and standardization.

## %>%

The pipe operator is a simple function that takes everything on the left-hand
side (lhs) of the pipe and "pipes" it as the first argument to the function
on the right-hand side. 

```{r}
16 %>% sqrt()
```

This becomes especially useful when used in conjuction with additional pipe
operators. 

```{r}
256 %>% sqrt() %>% sqrt()
```

The original pipe operator has moved around a bit, but (I believe) it is now
defined in the [magrittr](https://cran.r-project.org/web/packages/magrittr/index.html).
Now, base R has its own pipe operator as of version 4.1.0.

```{r}
16 |> sqrt()
```

Here is a [brief history of the pipe operator in R](http://adolfoalvarez.cl/blog/2021-09-16-plumbers-chains-and-famous-painters-the-history-of-the-pipe-operator-in-r/). 


## Combining operations

```{r}
ToothGrowth %>%
  filter(dose == 0.5, supp == "VC") %>%
  summarize(
    n = n(), 
    mean = mean(len),
    sd = sd(len))
```


## split-apply-combine

Often, we want to split our data up, apply some operations to it, 
and then combine the splits back together. 
In the `ToothGrowth` data.frame, 
we may want to calculate some summary statistics for both delivery methods. 

### dplyr::group_by()

```{r}
ToothGrowth %>%
  group_by(supp, dose) %>%
  summarize(
    n = n(),
    mean = mean(len),
    sd = sd(len),
    .groups = "drop"
  ) 
```

## Advanced plotting


```{r}
s <- ToothGrowth %>%
  group_by(supp, dose) %>%
  summarize(
    n = n(),
    mean = mean(len),
    sd = sd(len),
    .groups = "drop"
  ) %>%
  mutate(
    ucl = mean + qt(.975, n-1)*sd/sqrt(n),
    lcl = mean - qt(.975, n-1)*sd/sqrt(n)
  )

dw <- 0.1
ggplot(s, 
       aes(x = dose, color = supp, shape = supp)) +
  
  geom_point(
    data = ToothGrowth,
    aes(y = len),
    position = position_jitterdodge(dodge.width = dw, jitter.width = 0.05)) +
  
  geom_pointrange(
    data = s, 
    aes(y = mean, ymin = lcl, ymax = ucl),
    position = position_dodge(width = dw), 
    shape = 0,
    show.legend = FALSE
    ) +
  
  geom_line(
    aes(y = mean, group = supp),
    position = position_dodge(width = dw)) +
  
  labs(
    x = "Dose (mg/day)", 
    y = "Length (\u00b5m)", # unicode \u00b5 is the Greek letter mu
    title = "Odontoblast length vs Vitamin C in Guinea Pigs",
    color = "Delivery Method",
    shape = "Delivery Method") +
  
  theme_bw() +
  theme(legend.position = c(0.8, 0.2),
        legend.background = element_rect(fill = "white",
                                        color = "black"))
  
```

# 
