---
title: "Lab01 - Introduction to R"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---


To follow along, use the [lab01 code](lab01.R).


## Detailed introduction

For an extremely detailed introduction, please see

```{r, eval=FALSE, purl=FALSE}
help.start()
```

In this documentation, the above command will be executed at the command prompt, see below. 

## Brief introduction to R

From `help.start()`:

> R is an integrated suite of software facilities for data manipulation, calculation and graphical display.

and from <https://www.rstudio.com/products/RStudio/>:

> RStudio is an integrated development environment (IDE) for R. 

### R interface

In contrast to many other statistical software packages that use a point-and-click interface, e.g. SPSS, JMP, Stata, etc, R has a command-line interface. The command line has a command prompt, e.g. `>`, see below.

```{r, eval=FALSE, purl=FALSE}
>
```

This means, that you will be entering commands on this command line and hitting enter to execute them, e.g. 

```{r, eval=FALSE, purl=FALSE}
help()
```

Use the **up arrow** to recover past commands.

```{r, eval=FALSE, purl=FALSE}
hepl()
help() # Use up arrow and fix
```


### R GUI (or [RStudio](http://www.rstudio.com/))

Most likely, you are using a graphical user interface (GUI) and therefore, 
in addition, to the command line, you also have a windowed version of R with 
some point-and-click options, e.g. File, Edit, and Help.

In particular, there is an editor to create a new R script. So rather than 
entering commands on the command line, you will write commands in a script and 
then send those commands to the command line using `Ctrl-R` (PC) or 
`Command-Enter` (Mac).

```{r}
a = 1 
b = 2
a+b
```

Multiple lines can be run in sequence by selecting them and then using 
`Ctrl-R` (PC) or `Command-Enter` (Mac).




### Intro Activity

One of the most effective ways to use this documentation is to cut-and-paste 
the commands into a script and then execute them. 

Cut-and-paste the following commands into a **new script** and then run those 
commands directly from the script using `Ctrl-R` (PC) or `Command-Enter` (Mac).

```{r}
x = 1:10
y = rep(c(1,2), each=5)
m = lm(y~x)
s = summary(m)
```

Now, look at the result of each line

```{r, eval=FALSE}
x
y
m
s
s$r.squared
```


<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
x
y
m
s
s$r.squared
```
</details>






## Using R as a calculator

### Basic calculator operations

All basic calculator operations can be performed in R. 

```{r}
1+2
1-2
1/2
1*2
```

For now, you can ignore the [1] at the beginning of the line, 
we'll learn about that when we get to vectors. 


### Advanced calculator operations

Many advanced calculator operations are also available.

```{r}
(1+3)*2 + 100^2  # standard order of operations
sin(2*pi)        # the result is in scientific notation, i.e. -2.449294 x 10^-16 
sqrt(4)
10^2
log(10)          # the default is base e
log(10, base=10)
```



### Using variables

A real advantage to using R rather than a calculator (or calculator app) is the 
ability to store quantities using variables.

```{r}
a = 1
b = 2
a+b
a-b
a/b
a*b
```

### Assignment operators =, <-, and ->

When assigning variables values, you can also use arrows <- and -> and you will 
often see this in code, e.g. 

```{r}
a <- 1
2 -> b
c = 3  # is the same as <-
```

Now print them.

```{r}
a
b
c
```

### Using informative variable names

While using variables alone is useful, it is much more useful to use informative
variables names.

```{r}
# Rectangle
length <- 4
width  <- 3

area <- length * width
area


# Circle
radius <- 2
area   <- pi*radius^2 # this overwrites the previous `area` variable
circumference <- 2*pi*radius
area
circumference


# (Right) Triangle
opposite     <- 1
angleDegrees <- 30
angleRadians <- angleDegrees * pi/180
(adjacent     <- opposite / tan(angleRadians)) # = sqrt(3)
(hypotenuse   <- opposite / sin(angleRadians)) # = 2
```



### Calculator Activity

#### [Bayes' Rule](http://en.wikipedia.org/wiki/Bayes%27_theorem)

Suppose an individual tests positive for a disease, what is the probability the 
individual has the disease? Let 

- $D$ indicates the individual has the disease
- $N$ means the individual does not have the disease
- $+$ indicates a positive test result
- $-$ indicates a negative test 

The above probability can be calculated using [Bayes' Rule](http://en.wikipedia.org/wiki/Bayes%27_rule):

\[ P(D|+) = \frac{P(+|D)P(D)}{P(+|D)P(D)+P(+|N)P(N)} = 
\frac{P(+|D)P(D)}{P(+|D)P(D)+(1-P(-|N))\times(1-P(D))} \]

where

- $P(+|D)$ is the [sensitivity](http://en.wikipedia.org/wiki/Sensitivity_and_specificity) of the test
- $P(-|N)$ is the [specificity](http://en.wikipedia.org/wiki/Sensitivity_and_specificity) of the test
- $P(D)$ is the [prevalence](http://en.wikipedia.org/wiki/Prevalence) of the disease

Calculate the probability the individual has the disease if the test is positive when

- the specificity of the test is 0.95, 
- the sensitivity of the test is 0.99, and
- the prevalence of the disease is 0.001.

```{r, echo=FALSE}
# Find the probability the individual has the disease if 
# specificity is 0.95, sensitivity is 0.99, and prevalence is 0.001
```


<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
specificity = 0.95
sensitivity = 0.99
prevalence = 0.001
probability = (sensitivity*prevalence) / (sensitivity*prevalence + (1-specificity)*(1-prevalence))
probability
```
</details>






## <a name="types">Data types</a>

Objects in R can be broadly classified according to their dimensions:

- scalar
- vector
- matrix
- array (higher dimensional matrix)

and according to the type of variable they contain:

- integer
- numeric
- character (string)
- logical
- factor


### Scalars

Scalars have a single value assigned to the object in R. 

```{r}
a = 3.14159265 
b = "STAT 587 (Eng)" 
c = TRUE
```

Print the objects

```{r}
a
b
c
```


### Vectors

The `c()` function creates a vector in R

```{r}
a = c(1,2,-5,3.6)
b = c("STAT","587", "(Eng)")
c = c(TRUE, FALSE, TRUE, TRUE)
```

To determine the length of a vector in R use `length()`

```{r}
length(a)
length(b)
length(c)
```

To determine the type of a vector in R use `class()`

```{r}
class(a)
class(b)
class(c)
```


#### Vector construction

Create a numeric vector that is a sequence using : or `seq()`.

```{r}
1:10
5:-2
seq(from = 2, to = 5, by = .05)
```

Another useful function to create vectors is `rep()`

```{r}
rep(1:4, times = 2)
rep(1:4, each  = 2)
rep(1:4, each  = 2, times = 2)
```

Arguments to functions in R can be referenced either by position or by name or
both. 
The safest and easiest to read approach is to name all your arguments.
I will often name all but the first argument. 


#### Accessing vector elements 

Elements of a vector can be accessed using brackets, e.g. [index].

```{r}
a = c("one","two","three","four","five")
a[1]
a[2:4]
a[c(3,5)]
a[rep(3,4)]
```

Alternatively we can access elements using a logical vector where only TRUE 
elements are accessed.

```{r}
a[c(TRUE, TRUE, FALSE, FALSE, FALSE)]
```

You can also see all elements except some using a negative sign `-`.

```{r}
a[-1]
a[-(2:3)]
```

#### Modifying elements of a vector

You can assign new values to elements in a vector using = or <-.

```{r}
a[2] = "twenty-two"
a
a[3:4] = "three-four" # assigns "three-four" to both the 3rd and 4th elements
a
a[c(3,5)] = c("thirty-three","fifty-five")
a
```



### Matrices

Matrices can be constructed using `cbind()`, `rbind()`, and `matrix()`:

```{r}
m1 = cbind(c(1,2), c(3,4))       # Column bind
m2 = rbind(c(1,3), c(2,4))       # Row bind

m1
all.equal(m1, m2)

m3 = matrix(1:4, nrow = 2, ncol = 2)
all.equal(m1, m3)

m4 = matrix(1:4, nrow = 2, ncol = 2, byrow = TRUE)
all.equal(m3, m4)

m3
m4
```


#### Accessing matrix elements

Elements of a matrix can be accessed using brackets separated by a comma, 
e.g. [row index, column index].

```{r}
m = matrix(1:12, nrow=3, ncol=4)
m
m[2,3]
```

Multiple elements can be accessed at once

```{r}
m[1:2,3:4]
```

If no row (column) index is provided, then the whole row (column) is accessed.

```{r}
m[1:2,]
```

Like vectors, you can eliminate rows (or columns)

```{r}
m[-c(3,4),]
```

Be careful not to forget the comma, e.g. 

```{r}
m[1:4]
```

You can also construct an object with more than 2 dimensions using the 
`array()` function.


### Cannot mix types 

You cannot mix types within a vector, matrix, or array

```{r}
c(1,"a")
```

The number 1 is in quotes indicating that R is treating it as a character rather 
than a numeric.

```{r}
c(TRUE, 1, FALSE)
```

The logicals are converted to numeric (0 for FALSE and 1 for TRUE).

```{r}
c(TRUE, 1, "a")
```

Everything is converted to a character. 


### Activity

Using the matrix below,  

1. Print the element in the 3rd-row and 4th column
1. Print the 2nd column
1. Print all but the 3rd row

```{r}
m = rbind(c(1, 12, 8, 6),
          c(4, 10, 2, 9),
          c(11, 3, 5, 7))
m
```

If you have extra time, to try create this same matrix using the `matrix()` function.

```{r, echo=FALSE}
# Print the element in the 3rd-row and 4th column

# Print the 2nd column

# Print all but the 3rd row

# Reconstruct the matrix if time allows
```


<details><summary>Click for solution</summary> 
```{r, purl=FALSE}
# Print the element in the 3rd-row and 4th column
m[3,4]

# Print the 2nd column
m[,2]

# Print all but the 3rd row
m[-3,]

# Reconstruct the matrix if time allows
n = matrix(c(1,12,8,6,4,10,2,9,11,3,5,7), nrow=3, ncol=4, byrow=TRUE)
n
all.equal(m,n)
```
</details>


## <a name="frames">Data frames</a>

A `data.frame` is a special type of matrix that allows different data types in 
different columns.


```{r}
class(warpbreaks) # warpbreaks is a built-in data.frame
```


### Access `data.frame` elements

A `data.frame` can be accessed just like a matrix, 
e.g. [row index, column index].

```{r}
warpbreaks[1:3,1:2]
```

`data.frame`s can also be accessed by column names. 
In order to determine the column names use the `names()` function.

```{r}
names(warpbreaks)
warpbreaks[1:3, c("breaks","wool")]
```





### Different data types in different columns

The function `str()` allows you to see the structure of any object in R. 
Using `str()` on a `data.frame` object tells you 

1. that the object is a `data.frame`,
1. the number of rows and columns,
1. the names of each column,
1. each column's data type, and
1. the first few elements in that column.

```{r}
str(warpbreaks)
```




## Packages

Much of the functionality in R is available in R packages. 
Many of these packages are hosted on 
[CRAN](https://cran.r-project.org/) and 
[mirrored](https://cran.r-project.org/mirrors.html) to many locations,
e.g. [ISU CRAN mirror](https://mirror.las.iastate.edu/CRAN/).
[Bioconductor](https://www.bioconductor.org/) also hosts a number of 
packages primarily for bioinformatics uses. 
Packages can also be installed directly from GitHub. 
We will focus on packages available on CRAN. 

### Installing packages from CRAN

One of the first packages we will need in these labs is `ggplot2`.
Since this package is hosted on CRAN, 
we can install it using
```{r, eval=FALSE}
install.packages("ggplot2")
```
You may have to choose a CRAN repository. 
I suggest choosing the repository that is geographically the closest, 
e.g. [ISU CRAN mirror](https://mirror.las.iastate.edu/CRAN/).

When you install a package, the installation process will automatically install
any dependencies for that package. 
For example, `ggplot2` depends on `rlang`, `tibble`, and other packages.

The `ggplot2` package is part of the [tidyverse](https://www.tidyverse.org/)
and we will be using many of the packages in the tidyverse throughout this
course. 
Thus I suggest you go ahead and install the `tidyverse` package which is 
primarily a wrapper to install `ggplot2`, `dplyr`, `readr`, and other packages.

```{r, eval=FALSE}
install.packages("tidyverse")
```

### Installing packages from GitHub

Packages hosted on CRAN undergo substantial testing to ensure the technicalities
of package creation are satisfied. 
This testing does not automatically ensure the package does what it says it 
does. 
Many package authors don't want to go through the hassle of this testing and 
thus host their package on GitHub. 
In order to install a package directly from GitHub, 
you need to first install the `devtools` package and,
if you are on a Windows system, [Rtools](https://cran.r-project.org/bin/windows/Rtools/).

To install the `devtools` package, use 
```{r, eval = FALSE}
install.packages("devtools")
```
and to install Rtools, follow [these instructions](https://cran.r-project.org/bin/windows/Rtools/).

Once devtools has been installed, you can use the `install_github()` function
from the `devtools` package to install a package from CRAN.
To install a package, you need to know the GitHub username and the repository
name. 
For example, to install the `swgoh` package from user `jarad` (that's me),
you can use 
```{r, eval = FALSE}
devtools::install_github("jarad/swgoh")
```

### Loading packages

A package only needs to be installed once on your system (until you upgrade R),
but you will need to either load the package when you want to use it or
explicitly indicate the package name when using functions in the package. 
To load the package in an R session, use
```{r, eval = FALSE}
library("ggplot2")
```
If you want to explicitly call the function, then you use 
<package name>::<function name>. 
For example
```{r, eval = FALSE}
devtools::install_github("jarad/swgoh")
```
uses the `install_github()` function from the `devtools` package.

Personally, I have started to do both in my scripts. 
That is, at the top of the script I include a library call for every package
that is required by the script. 
Then, throughout the script, I explicitly call each function that comes from
a manually loaded package. 


## tidyverse R style guide

Coding in R, like any language, is a process. 
It is helpful if you have a structure to start from. 
The [tidyverse R style guide](https://style.tidyverse.org/) provides 
suggestions for best practices when coding in R.




## Getting help for R

### Learning R

To learn R, you may want to try the swirl package. 
To install, use 

```{r, eval=FALSE}
install.packages("swirl")
```

After installation, use the following to get started

```{r, eval=FALSE}
library("swirl")
swirl()
```

### General help

As you work with R, there will be many times when you need to get help. 

My basic approach is 
 
1. Use the help contained within R.
1. Perform an internet search for an answer.
1. Find somebody else who knows.

In all cases, knowing the R keywords, e.g. a function name, will be extremely 
helpful.


### Help within R I

If you know the function name, then you can use `?<function>`, e.g.

```{r, eval=FALSE}
?mean
```

The structure of help is 

- Description: quick description of what the function does
- Usage: the arguments, their order, and default values (if any)
- Arguments: more thorough description about the arguments
- Value: what the funtion returns
- See Also: similar functions
- Examples: examples of how to use the function


### Help within R II

If you cannot remember the function name, then you can use 
`help.search("<something>")`, e.g.

```{r, eval=FALSE}
help.search("mean")
```

Depending on how many packages you have installed, you will find a lot or a 
little here. 



### Internet search for R help

I google for `<something> R`, e.g. 

    calculate mean R

Some useful sites are 

- <http://www.cookbook-r.com/>
- <http://www.r-tutor.com/r-introduction>
- <http://www.statmethods.net/>
- <http://stackoverflow.com/questions/tagged/r>
- <http://www.ats.ucla.edu/stat/r/>
- <https://sites.google.com/site/rapplicationforbiosurveillance/home/discussions>



### <a name="help">Getting help on ggplot2</a>

Although the general R help can still be used, e.g. 

```{r, eval=FALSE}
?ggplot
?geom_point
```

It is much more helpful to google for an answer

    geom_point 
    ggplot2 line colors

The top hits will all have the code along with what the code produces.

#### Helpful sites

These sites all provide code. The first two also provide the plots that are produced.

- <http://docs.ggplot2.org/current/>
- <http://www.cookbook-r.com/Graphs/>
- <http://stackoverflow.com/questions/tagged/ggplot2>
