---
title: "Lab02 - Graphical Statistics"
author: "Jarad Niemi"
date: "`r Sys.Date()`"
output: html_document
---

### Setup

To follow along, use the [lab02 code](lab02.R) also run the following

```{r, eval=FALSE}
install.packages("tidyverse")
```

If the install fails, then run

```{r, eval=FALSE}
install.packages(c("dplyr","tidyr","ggplot2"))
```

The installation only needs to be done once. 
But we will need to load the packages in every R session where we want to use 
them.
To load the packages, use 

```{r, eval=FALSE}
library("dplyr")
library("tidyr")
library("ggplot2")
```

alternatively, you can load the entire (not very big) `tidyverse`.

```{r}
library("tidyverse")
```



## Constructing plots

The main purpose of the lab today is to construct plots using the `ggplot2`
R package.
In order to construct these plots, 
we need to construct an appropriate `data.frame` and we will use 
`dplyr` to help us construct that `data.frame`. 




Let's use the built-in R data set `airquality`.
Before we start plotting let's take a quick look at the data.

```{r}
dim(airquality)
head(airquality)
tail(airquality)
summary(airquality)
```

For built in datasets, 
we can get more information by going to the help file.

```{r, eval=FALSE}
?airquality
```

One issue with this dataset is that the Month/Day columns don't really provide 
us with a `Date`.
Let's create a new column that creates a real `Date`.

```{r}
airquality <- airquality %>%
  mutate(Date = as.Date(paste("1973",Month,Day,sep="/"))) 
```

If you deal with dates a lot, 
you should check out the [lubridate](https://cran.r-project.org/package=lubridate) 
package.

### Histogram

All `ggplot2` graphics require a `data.frame` containing the data
and this `data.frame` is always the first argument to a `ggplot` 
call.
After this, we specify some aesthetics using the `aes()` function.
Then, we tell `ggplot2` what kind of graphics to construct.

```{r}
ggplot(airquality,     # data.frame containing the data
       aes(x=Ozone)) + # a column name from the data.frame
  geom_histogram()     # create a histogram
```

If you want to avoid the message, 
you can specify the number of bins to use.

```{r}
ggplot(airquality, aes(x=Ozone)) + 
  geom_histogram(bins = 40)
```

If you want plot on the density scale 
(so that you can compare to a pdf),
use the following:


```{r}
ggplot(airquality, aes(x=Ozone)) + 
  geom_histogram(aes(y=..density..), bins = 40)
```


#### Histogram activity

Create a histogram of solar radiation on the density scale with 50 bins.

<div id="histogram_activity_solution" style="display:none"> 
```{r, purl=FALSE}
ggplot(airquality, aes(x=Solar.R)) + 
  geom_histogram(aes(y=..density..), bins = 50)
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('histogram_activity_solution') .style.display=='none') {document.getElementById('histogram_activity_solution') .style.display=''}else{document.getElementById('histogram_activity_solution') .style.display='none'}">Show/Hide Solution</button>




### Boxplots

The syntax for boxplots is similar except that the variable you are
interest in is the y aesthetic.

```{r}
ggplot(airquality,     
       aes(x=1,y=Ozone)) + 
  geom_boxplot()     
```

Comparing boxplots

```{r}
ggplot(airquality,     
       aes(x=Month, y=Ozone, group=Month)) + 
  geom_boxplot()     
```

I dislike boxplots as it hides the data behind a 5 point summary in 
the shape of a box and whiskers. 
I prefer to see the data and the boxplot together.

```{r}
ggplot(airquality,     
       aes(x=Month, y=Ozone, group=Month)) + 
  geom_boxplot(color='grey',                 # make the boxes not so obvious
               outlier.shape = NA) +         # remove outliers, 
  geom_jitter() +                            # because they get plotted here
  theme_bw()                                 # Change the theme to remove gray
```

#### Boxplot activity

Create a scatterplot of wind speed by month and add a boxplot for each month
in the background.
See if you can google to find out how to swap the axes, i.e. have month on the 
y-axis and Wind on the x-axis.

<div id="boxplot_activity_solution" style="display:none"> 
```{r, purl=FALSE}
ggplot(airquality,     
       aes(x=Month, y=Wind, group=Month)) + 
  geom_boxplot(outlier.shape = NA, color='grey') +         
  geom_jitter() +                       
  theme_bw() +
  coord_flip()
```

Flipping the axes makes the comparisons vertical and therefore, I think, easier.
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('boxplot_activity_solution') .style.display=='none') {document.getElementById('boxplot_activity_solution') .style.display=''}else{document.getElementById('boxplot_activity_solution') .style.display='none'}">Show/Hide Solution</button>



### Scatterplot

At this point we can construct individual graphs for our 4 different response
variables: Ozone, Solar.R, Wind, and Temp.
Perhaps we want to understand the temporal variability for Ozone.
We can use a scatterplot of Ozone vs Date.

```{r}
ggplot(airquality, aes(x = Date, y = Ozone)) +
  geom_point()
```

or if we wanted a line plot

```{r}
ggplot(airquality, aes(x = Date, y = Ozone)) +
  geom_line()
```

Notice that the line is disconnected wherever we have missing data.

Perhaps we want to understand the relationship between solar radiation and 
ozone. 

```{r}
ggplot(airquality, aes(x = Solar.R, y = Ozone)) +
  geom_point()
```



#### Scatterplot activity

Create a scatterplot of wind speed versus temperature.


<div id="scatterplot_activity_solution" style="display:none"> 
```{r, purl=FALSE}
ggplot(airquality, aes(x = Temp, y = Wind)) +
  geom_point()
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('scatterplot_activity_solution') .style.display=='none') {document.getElementById('scatterplot_activity_solution') .style.display=''}else{document.getElementById('scatterplot_activity_solution') .style.display='none'}">Show/Hide Solution</button>


If we want to put all the response variables on the same plot, we can color 
them.
In order to do this, we will need to organize our `data.frame` into long format. 

```{r}
airquality_long <- airquality %>%
  select(-Month, -Day) %>%              # Remove these columns
  tidyr::gather(response, value, -Date)
```

Take a look at the resulting `data.frame`.

```{r}
dim(airquality)
dim(airquality_long)

head(airquality_long)
summary(airquality_long)
table(airquality_long$response)
```

```{r}
ggplot(airquality_long, 
       aes(x = Date, y = value, 
           linetype = response,
           color = response, 
           group = response)) +
  geom_line()
```

Notice that the legend is automatically created.
This is not something that is done in base R graphics.

Honestly, this doesn't look very good, 
so it is better to facet the plot.



### Faceted scatterplots

Facets are often a better way of representing multiple variables.

```{r}
ggplot(airquality_long, aes(Date, value)) +
  geom_point() + 
  facet_wrap(~response)
```

Since the axes are quite different for the different responses, 
we can allow them to vary in the different facets.

```{r}
ggplot(airquality_long, aes(Date, value)) +
  geom_line() + 
  facet_wrap(~response,scales="free_y")
```




### Customizing

If we only had the long version of the `data.frame`, we can reconstruct the
wide version by using the following

```{r}
airquality2 <- airquality_long %>%
  tidyr::spread(response, value)
```

Sometimes it is helpful to save the plot as an R object 
so that it can be updated in the future. 
To save the plot, just use the assignment operator, e.g. 

```{r}
g <- ggplot(airquality2,
       aes(x = Temp, y = Wind)) +
  geom_point()

g # Then you can see the plot by just typing the object name
```

We would like this plot to be a bit more informative, 
so we will add some informative labels.

```{r}
g <- g +
  labs(x = "Temperature (F)",
       y = "Wind speed (mph)",
       title = "New York (May-September 1973)")

g
```

As you have seen before, we can also change the theme. 
I prefer the simple "bw" theme.

```{r}
g <- g + theme_bw()
g
```

We can add a regression line.

```{r}
g <- g + geom_smooth(method="lm")
g
```

Alternatively, you can combine all the steps

```{r}
ggplot(airquality2,
       aes(x = Temp, y = Wind)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  labs(x = "Temperature (F)",
       y = "Wind speed (mph)",
       title = "New York (May-September 1973)") + 
  theme_bw()
```



#### Plot creation activity

Use the `cars` dataset to construct and customize a figure displaying the
relationship between the stopping distance and speed of a car.

<div id="plot_creation_activity_solution" style="display:none"> 
```{r, purl=FALSE}
ggplot(cars,
       aes(x=speed, y=dist)) +
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(x = "Speed (mph)",
       y = "Stopping distance (ft)",
       title = "Stopping distance as a function of speed (1920s)") +
  theme_bw()
```
</div> 
<button title="Show a solution" type="button" onclick="if(document.getElementById('plot_creation_activity_solution') .style.display=='none') {document.getElementById('plot_creation_activity_solution') .style.display=''}else{document.getElementById('plot_creation_activity_solution') .style.display='none'}">Show/Hide Solution</button>


If you want to save the plot, use the ggsave function, e.g.

```{r,eval=FALSE}
ggsave(filename = "plot.png", plot = g, width = 5, height = 4)
```



```{r, echo=FALSE, purl=FALSE, results="hide", message=FALSE}
knitr::knit("lab02.Rmd", tangle = TRUE)
```
